---
title: "MG-970: Create proteomics mock data"
---
Related Jira issues: MG-970

This notebook generates mock data to support developing proteomics interfaces:
* Protein - Differential Expression CT
* ***Not yet implemented: Individual Protein Expression (details box plot pages)***

Mock data for the Protein - Differential Expression CT supports the use cases outlined in this document:
https://docs.google.com/document/d/1IWsp41hKJN8xcWsAvtsIBmrTfdwi8Z55V48_j92z-As/edit?tab=t.0#heading=h.lmosv24ok9i6

## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}
if (!require("jsonlite", quietly = TRUE)) { install.packages("jsonlite") }
if (!require("purrr")) { install.packages("purrr") }

library(synapser)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(jsonlite)
```


## Variables and functions
Sources the required variables and functions.
```{r}
# *******************************
# STATIC VARIABLES
# *******************************
# mouse values
MODEL_NAME      <- 'LOAD2'
MODEL_GROUP     <- 'LOAD2'
MATCHED_CONTROL <- 'LOAD1'
MODEL_TYPE      <- 'Late Onset AD'
TISSUE          <- 'Hemibrain'
SEXES           <- c('Females', 'Males')
AGES            <- c(4, 12, 18, 24)


# gene/protein values
ENSEMBL_PREFIX <- 'ENSMUSG'
BIODOMAINS <- c('Apoptosis',
                   'APP Metabolism',
                   'Autophagy',
                   'Cell Cycle',
                   'DNA Repair',
                   'Endolysosome',
                   'Epigenetic',
                   'Immune Response',
                   'Lipid Metabolism',
                   'Metal Binding and Homeostasis',
                   'Mitochondrial Metabolism',
                   'Myelination',
                   'Oxidative Stress',
                   'Proteostasis',
                   'RNA Spliceosome',
                   'Structural Stabilization',
                   'Synapse',
                   'Tau Homeostasis',
                   'Vasculature')


# *********
# FUNCTIONS
# *********

# Builds a Proteomics DE CT object
# ensembl_gene_id, gene_symbol, uniprotid, unique_id, display_symbol, biodomains[], name{}, matched_control, model_group, model_type, tissue, sex, age_cols...
build_row <- function(ensembl_gene_id, gene_symbol, uniprotid, unique_id, display_symbol, biodomains, name, matched_control, model_group, model_type, tissue, sex, age_rows = list()) {


  base_row <- list(
    ensembl_gene_id = ensembl_gene_id,
    gene_symbol = gene_symbol,
    uniprotid = uniprotid,
    unique_id = unique_id,
    display_symbol = display_symbol,
    biodomains = biodomains,
    name = name,
    matched_control = matched_control,
    model_group = model_group,
    model_type = model_type,
    tissue = tissue,
    sex = sex
  )

  row <- c(base_row, age_rows) %>% purrr::discard(~ is.null(.x) || all(is.na(.x)))

  row
}

# Extract the specified number of values from a vector
# if sample_size > length(vec), duplicate values will be included
sample_vector <- function(sample_size, vec) {

  padding_needed <- sample_size - length(vec)

  if (padding_needed <= 0) {
    sampled_indices <- sample(seq_along(vec), size = sample_size, replace = FALSE)
    return(vec[sampled_indices])
  }

  else {
    base_indices <- sample(seq_along(vec), size = length(vec), replace = FALSE)
    base_pool <- vec[base_indices]

    padding_indices <- sample(seq_along(vec), size = padding_needed, replace = TRUE)
    padding_pool <- vec[padding_indices]

    return(c(base_pool, padding_pool))
  }
}

# Generate a heatmap data object
generate_heatmap_object <- function(key_name, min = -2, max = 2) {
  log2_fc <- runif(1, min = min, max = max)
  adj_p_val <- runif(1, min = 0, max = 1)

  value_list <- list(
    log2_fc = round(log2_fc, 5),
    adj_p_val = round(adj_p_val, 5)
  )

  heatmap_obj <- list()
  heatmap_obj[[key_name]] <- value_list

  return(heatmap_obj)
}

generate_link_object <- function(key_name, link_url = NULL, link_text = NULL) {

  value_list <- list(
    link_url = link_url,
    link_text = link_text
  )

  link_obj <- list()
  link_obj[[key_name]] <- value_list

  return(link_obj)
}
```


# *********
# Mock Data
# *********
Requirement summary
* >50 unique genes, 1:1, all match the same typeahead value (can be one letter)
* 1:3+ x2+ genes, must exist in RNA DE
* TBD gene with RNA but no protein result
* TBD (fake?) gene with protein but no RNA
* uniprotid contains a hyphen


# 1. basic_results
* 60 pairs of objects (m/f with matching metadata and unique measurements)
* 1 gene : 1 uniprotid
* All gene_symbol values include "GeneP" for typeahead
* Some objects don't have '18 months' results
```{r}
# Populate gene/protein fields: ensembl_gene_id, gene_symbol, uniprotid, unique_id, display_symbol
ids <- 0:59
pad_number <- "1"
object_dict <- tibble(

  ensembl_gene_id = paste0(ENSEMBL_PREFIX, gsub(" ", pad_number, sprintf("%11d", ids))),
  gene_symbol = paste0("GeneX_", ids),
  uniprotid = paste0(
    "P", gsub(" ", pad_number, sprintf("%5d", ids)),
    sample(c("", "-1", "-2", "-3"), size = length(ids), replace = TRUE, prob = c(0.7, 0.2, 0.07, 0.03))
  ),
  unique_id = paste0(ensembl_gene_id, uniprotid),
  display_symbol = paste0(gene_symbol, " (", uniprotid, ")")
)

# biodomains[], name{}, matched_control, model_group, model_type, tissue, sex, age_cols...

# generate link object for name field
name_link_object <- generate_link_object("name", link_url = paste0("/models/", MODEL_NAME), link_text = MODEL_NAME)[["name"]]

# Populate fields that match across sexes
# biodomains, name, matched_control, model_group, model_type, tissue
object_dict <- object_dict %>%

  # 15% chance this object won't have 24 months data
  mutate(has_24_months = runif(nrow(object_dict)) >= 0.15) %>%

  rowwise() %>%
  mutate(
    biodomains = list(sample_vector(sample(1:10, 1), BIODOMAINS)),
    name            = list(name_link_object),
    matched_control = MATCHED_CONTROL,
    model_group     = MODEL_GROUP,
    model_type      = MODEL_TYPE,
    tissue          = TISSUE,
  ) %>%
  ungroup()

# Expand to include an object per sex for each gene:protein
object_dict_expanded <- object_dict %>%
  crossing(sex = SEXES)

# Generate row-specific heatmap data
basic_results <- object_dict_expanded %>%
  mutate(
    # 1. Map over an existing column to execute exactly once per row
    `4 months`  = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("4 months")[["4 months"]]),
    `12 months` = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("12 months")[["12 months"]]),
    `18 months` = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("18 months")[["18 months"]]),

    # 2. Loop through the logical flags one-by-one so if() only ever sees a single TRUE/FALSE
    `24 months` = purrr::map(has_24_months, ~ if (.x) generate_heatmap_object("24 months")[["24 months"]] else NULL)
  ) %>%
  select(-has_24_months)


basic_results
```


2. rna_exists_multiple_proteins_results
* 15 pairs of objects (m/f with matching metadata and unique measurments)
* 1 gene : 3 uniprotid
* Some objects don't have '18 months' results
```{r}
ids <- 100:114
ensembl_id_in_rna_results <- c("ENSMUSG00000000001", "ENSMUSG00000000001", "ENSMUSG00000000001",
                               "ENSMUSG00000000056", "ENSMUSG00000000056", "ENSMUSG00000000056",
                               "ENSMUSG00000000214", "ENSMUSG00000000214", "ENSMUSG00000000214",
                               "ENSMUSG00000000296", "ENSMUSG00000000296", "ENSMUSG00000000296",
                               "ENSMUSG00000000340", "ENSMUSG00000000340", "ENSMUSG00000000340"
                               )
gene_symbol_in_rna_results <- c("Gnai3", "Gnai3", "Gnai3",
                                "Narf", "Narf", "Narf",
                                "Th", "Th", "Th",
                                "Tpd52l1", "Tpd52l1", "Tpd52l1",
                                "Dbt", "Dbt", "Dbt"
                                )

object_dict <- tibble(

  ensembl_gene_id = ensembl_id_in_rna_results,
  gene_symbol = gene_symbol_in_rna_results,
  uniprotid = paste0(
    sprintf("P%05d", ids),
    sample(c("", "-1", "-2", "-3"), size = length(ids), replace = TRUE, prob = c(0.7, 0.2, 0.07, 0.03))
  ),
  unique_id = paste0(ensembl_gene_id, uniprotid),
  display_symbol = paste0(gene_symbol, " (", uniprotid, ")")
)

# biodomains[], name{}, matched_control, model_group, model_type, tissue, sex, age_cols...

# generate link object for name field
name_link_object <- generate_link_object("name", link_url = paste0("/models/", MODEL_NAME), link_text = MODEL_NAME)[["name"]]

# Populate fields that match across sexes
# biodomains, name, matched_control, model_group, model_type, tissue
object_dict <- object_dict %>%

  # 15% chance this object won't have 18 months data
  mutate(has_24_months = runif(nrow(object_dict)) >= 0.15) %>%

  rowwise() %>%
  mutate(
    biodomains = list(sample_vector(sample(1:10, 1), BIODOMAINS)),
    name            = list(name_link_object),
    matched_control = MATCHED_CONTROL,
    model_group     = MODEL_GROUP,
    model_type      = MODEL_TYPE,
    tissue          = TISSUE,
  ) %>%
  ungroup()

# Expand to include an object per sex for each gene:protein
object_dict_expanded <- object_dict %>%
  crossing(sex = SEXES)

# Generate row-specific heatmap data
rna_exists_multiple_proteins_results <- object_dict_expanded %>%
  mutate(
    # 1. Map over an existing column to execute exactly once per row
    `4 months`  = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("4 months")[["4 months"]]),
    `12 months` = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("12 months")[["12 months"]]),
    `18 months` = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("18 months")[["18 months"]]),

    # 2. Loop through the logical flags one-by-one so if() only ever sees a single TRUE/FALSE
    `24 months` = purrr::map(has_24_months, ~ if (.x) generate_heatmap_object("24 months")[["24 months"]] else NULL)
  ) %>%
  select(-has_24_months)


rna_exists_multiple_proteins_results

```


3. no_rna_protein_exists_results
* 15 pairs of objects (m/f with matching metadata and unique measurements)
* 1 gene : 3 uniprotid
* Some objects don't have '18 months' results
```{r}
# Populate gene/protein fields: ensembl_gene_id, gene_symbol, uniprotid, unique_id, display_symbol
ids <- 0:14
pad_number <- "3"

object_dict <- tibble(

  ensembl_gene_id = paste0(ENSEMBL_PREFIX, gsub(" ", pad_number, sprintf("%11d", ids))),
  gene_symbol = paste0("GeneP_", ids),
  uniprotid = paste0(
    "P", gsub(" ", pad_number, sprintf("%5d", ids)),
    sample(c("", "-1", "-2", "-3"), size = length(ids), replace = TRUE, prob = c(0.7, 0.2, 0.07, 0.03))
  ),
  unique_id = paste0(ensembl_gene_id, uniprotid),
  display_symbol = paste0(gene_symbol, " (", uniprotid, ")")
)

# biodomains[], name{}, matched_control, model_group, model_type, tissue, sex, age_cols...

# generate link object for name field
name_link_object <- generate_link_object("name", link_url = paste0("/models/", MODEL_NAME), link_text = MODEL_NAME)[["name"]]

# Populate fields that match across sexes
# biodomains, name, matched_control, model_group, model_type, tissue
object_dict <- object_dict %>%

  # 15% chance this object won't have 24 months data
  mutate(has_24_months = runif(nrow(object_dict)) >= 0.15) %>%

  rowwise() %>%
  mutate(
    biodomains = list(sample_vector(sample(1:10, 1), BIODOMAINS)),
    name            = list(name_link_object),
    matched_control = MATCHED_CONTROL,
    model_group     = MODEL_GROUP,
    model_type      = MODEL_TYPE,
    tissue          = TISSUE,
  ) %>%
  ungroup()

# Expand to include an object per sex for each gene:protein
object_dict_expanded <- object_dict %>%
  crossing(sex = SEXES)

# Generate row-specific heatmap data
no_rna_protein_exists_results <- object_dict_expanded %>%
  mutate(
    # 1. Map over an existing column to execute exactly once per row
    `4 months`  = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("4 months")[["4 months"]]),
    `12 months` = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("12 months")[["12 months"]]),
    `18 months` = purrr::map(ensembl_gene_id, ~ generate_heatmap_object("18 months")[["18 months"]]),

    # 2. Loop through the logical flags one-by-one so if() only ever sees a single TRUE/FALSE
    `24 months` = purrr::map(has_24_months, ~ if (.x) generate_heatmap_object("24 months")[["24 months"]] else NULL)
  ) %>%
  select(-has_24_months)

no_rna_protein_exists_results
```


# Assemble and upload collection

```{r}
# *********************************************
# Generate ui_config.json and upload to Synapse
# *********************************************

raw_tibble_list <- list(
  basic_results,
  rna_exists_multiple_proteins_results,
  no_rna_protein_exists_results
)

# Since we are using a tibble, we need to do some extra stuff to omit the null 18 months keys entirely
json_list <- purrr::map(raw_tibble_list, function(df) {
  df %>%
    purrr::transpose() %>%
    purrr::map(~ purrr::discard(.x, is.null))
})

json_output <- toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
# cat(json_output)

# write json file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

write(json_output, file.path(output_dir, "proteomics_mocks.json"))
```


# Upload to synapse
*****************************************************
Validate ./output/proteomics_mocks.json before upload
*****************************************************
```{r}
synLogin()
syn_file <- File(
              file.path(output_dir, "ui_config.json"),
              description = "ui_config.json for Model-AD Explorer 2.0.",
              parent = "syn51498054"
            )

res <- synStore(
          syn_file,
          forceVersion = FALSE,
          executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-402_Create ui_config source file.rmd"
        )
# print synapse_id and new version number
print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```


-----
TESTS
-----


## Test build_row
```{r}
object <- build_row(
  ensembl_gene_id = "ENSMUSG00000022565",
  gene_symbol = "PLEC",
  uniprotid = "Q9QXS1",
  unique_id = "ENSMUSG00000022565Q9QXS1",
  display_symbol = "PLEC (Q9QXS1)",
  biodomains = c("Apoptosis","Autophagy","Cell Cycle"),
  name = "LOAD2",
  matched_control = "LOAD1",
  model_group = "LOAD2",
  model_type = "Late Onset AD",
  tissue = "Hemibrain",
  sex = "Female"
)

json_list <- c(
  object
)

json_output <- toJSON(json_list, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)

```


# test sample_data
```{r}
sample_vector(1, sex)
sample_vector(2, sex)
sample_vector(3, sex)
sample_vector(10, sex)

sample_vector(1, biodomains)
sample_vector(2, biodomains)
sample_vector(3, biodomains)
sample_vector(19, biodomains)
```

# test generate heatmap object
```{r}
object <- generate_heatmap_object("12 months")

json_list <- c(
  object
)

json_output <- toJSON(json_list, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)

```


# test generate link object
```{r}
object <- generate_link_object("name", "/models/LOAD2", "LOAD2")

json_list <- c(
  object
)

json_output <- toJSON(json_list, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)
```

sanity check results
```{r}
all_mocks <- rbind(
  basic_results,
  rna_exists_multiple_proteins_results,
  no_rna_protein_exists_results
)

nrow(all_mocks)

```

check for duplicate uniprotids
```{r}

all_mocks <- rbind(
  basic_results,
  rna_exists_multiple_proteins_results,
  no_rna_protein_exists_results
)

all_mocks %>%
  # 1. Select only rows with sex == Female
  filter(sex == "Females") %>%

  # 2. Look for duplicate uniprotid values among those rows
  group_by(uniprotid) %>%
  filter(n() > 1) %>%
  ungroup() %>%

  # Arrange them side-by-side for easy inspection
  arrange(uniprotid)

```