---
title: "MG-970: Create proteomics mock data"
---
Related Jira issues: MG-970

This notebook generates mock data to support developing proteomics interfaces:
* Protein - Differential Expression CT
* ***Not yet implemented: Individual Protein Expression (details box plot pages)***

Mock data for the Protein - Differential Expression CT supports the use cases outlined in this document:
https://docs.google.com/document/d/1IWsp41hKJN8xcWsAvtsIBmrTfdwi8Z55V48_j92z-As/edit?tab=t.0#heading=h.lmosv24ok9i6

## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}
if (!require("jsonlite", quietly = TRUE)) { install.packages("jsonlite") }
if (!require("purrr")) { install.packages("purrr") }

library(synapser)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(jsonlite)
```


## Variables and functions
Sources the required variables and functions.
```{r}
# *******************************
# STATIC VARIABLES
# *******************************
# mouse values
MODEL_NAME      <- 'LOAD2'
MODEL_GROUP     <- 'LOAD2'
MATCHED_CONTROL <- 'LOAD1'
MODEL_TYPE      <- 'Late Onset AD'
TISSUE          <- 'Hemibrain'
SEXES           <- c('Female', 'Male')
AGES            <- c(4, 12, 18, 24)

ensembl_regex = "^ENSMUSG\\d{11}$"

# gene/protein values
ENSEMBL_PREFIX <- 'ENSMUSG'
BIODOMAINS <- c('Apoptosis',
                'APP Metabolism',
                'Autophagy',
                'Cell Cycle',
                'DNA Repair',
                'Endolysosome',
                'Epigenetic',
                'Immune Response',
                'Lipid Metabolism',
                'Metal Binding and Homeostasis',
                'Mitochondrial Metabolism',
                'Myelination',
                'Oxidative Stress',
                'Proteostasis',
                'RNA Spliceosome',
                'Structural Stabilization',
                'Synapse',
                'Tau Homeostasis',
                'Vasculature')


# *********
# FUNCTIONS
# *********

# Generate a heatmap data object with deterministic values based on ensembl_gene_id + sex
generate_heatmap_object <- function(key_name, id = "default", min = -2, max = 2) {
  # Create a deterministic numeric seed from the string combination
  hash_seed <- sum(utf8ToInt(paste0(key_name, id))) * 1000

  # Sine-based pseudo-random noise generator (yields values bound between 0 and 1)
  noise_fc    <- (sin(hash_seed * 12.9898) * 43758.5453) %% 1
  noise_p_val <- (sin(hash_seed * 78.2330) * 43758.5453) %% 1

  # Scale noise to target ranges
  log2_fc   <- min + noise_fc * (max - min)
  adj_p_val <- noise_p_val

  value_list <- list(
    log2_fc   = round(log2_fc, 5),
    adj_p_val = round(adj_p_val, 5)
  )

  heatmap_obj <- list()
  heatmap_obj[[key_name]] <- value_list

  return(heatmap_obj)
}

# Generate a link data object
generate_link_object <- function(key_name, link_url = NULL, link_text = NULL) {

  value_list <- list(
    link_url = link_url,
    link_text = link_text
  )

  link_obj <- list()
  link_obj[[key_name]] <- value_list

  return(link_obj)
}

# Generate a list of biodomains; length and members are deterministic for a given i value
get_deterministic_biodomains <- function(i, domains) {
  n_domains <- length(domains)
  # Cycle domain counts from 1 to 5 across rows
  target_len <- ((i - 1) %% 5) + 1
  # Cycle start index through BIODOMAINS
  start_idx  <- ((i - 1) %% n_domains) + 1

  # Select slice with wraparound
  indices <- ((seq_len(target_len) + start_idx - 2) %% n_domains) + 1
  return(domains[indices])
}
```


# *********
# Mock Data
# *********
Requirement summary
* >50 unique genes, 1:1, all match the same typeahead value (can be one letter)
* 1:3+ x2+ genes, must exist in RNA DE
* TBD gene with RNA but no protein result
* TBD (fake?) gene with protein but no RNA
* uniprotid contains a hyphen


# 1. basic_results
* 60 pairs of objects (m/f with matching metadata and unique measurements)
* 1 gene : 1 uniprotid
* All gene_symbol values include "GeneX" for typeahead
* Some objects don't have '24 months' results
```{r}
# Populate gene/protein fields: ensembl_gene_id, gene_symbol, uniprotid, unique_id, display_symbol

# two-digit base id values + pad length == 11 == valid numeric part of ensembl_gene_id
ids <- 10:69

# pad number that does not allow duplicate id generation (1 vs 11 vs 111...)
pad_number <- "9"

# pattern used to deterministically generate unique_ids, some with isoform suffixes: no suffix (70%), -1 (20%), -2 (6.7%), -3 (3.3%)
isoform_pattern <- c(
  rep("", 7), rep("-1", 2), "-2",
  rep("", 7), rep("-1", 2), "-2",
  rep("", 7), rep("-1", 2), "-3"
)

object_dict <- tibble(

  ensembl_gene_id = paste0(ENSEMBL_PREFIX, gsub(" ", pad_number, sprintf("%11d", ids))),
  gene_symbol = paste0("GeneX_", ids),
  uniprotid = paste0(
    "P", gsub(" ", pad_number, sprintf("%5d", ids)),
    rep(isoform_pattern, length.out = length(ids))
  ),
  unique_id = paste0(ensembl_gene_id, uniprotid),
  display_symbol = paste0(gene_symbol, " (", uniprotid, ")")
)

# Populate fields that match across sexes: biodomains, name, matched_control, model_group, model_type, tissue

# generate link object for name field
name_link_object <- generate_link_object("name", link_url = paste0("/models/", MODEL_NAME), link_text = MODEL_NAME)[["name"]]

object_dict <- object_dict %>%
  mutate(
        biodomains = purrr::map(
      ensembl_gene_id,
      ~ {
        # Extract numeric digits from ensg as double float to prevent integer overflow
        gene_num <- as.numeric(gsub("\\D", "", .x))
        # rows for the same gene always get the same biodomains
        get_deterministic_biodomains(gene_num, BIODOMAINS)
      }
    ),
    name            = list(name_link_object),
    matched_control = MATCHED_CONTROL,
    model_group     = MODEL_GROUP,
    model_type      = MODEL_TYPE,
    tissue          = TISSUE

# Expand to include an object for each sex with identical metadata
  ) %>% crossing(sex = SEXES)


# populate row-specific heatmap data values

# pattern used to deterministically omit 24 month data (15% overall)
pattern_24m <- c(
  FALSE, FALSE,  # Gene Pair 1: Both missing
  FALSE, TRUE,   # Gene Pair 2: Female missing only
  TRUE,  FALSE,  # Gene Pair 3: Male missing only
  FALSE, FALSE,  # Gene Pair 4: Both missing
  rep(TRUE, 32)  # Genes 5-20: Neither missing
)

# Generate row-specific heatmap data
basic_results <- object_dict %>%
  dplyr::mutate(
    has_24_months = rep(pattern_24m, length.out = nrow(object_dict))
  ) %>%
  dplyr::mutate(
    # Populate timepoint data with deterministic values
    `4 months`  = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("4 months", paste0(.x, .y))[["4 months"]]),
    `12 months` = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("12 months", paste0(.x, .y))[["12 months"]]),
    `18 months` = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("18 months", paste0(.x, .y))[["18 months"]]),
    `24 months` = purrr::map2(has_24_months, paste0(ensembl_gene_id, sex), ~ {
      if (.x) generate_heatmap_object("24 months", .y)[["24 months"]] else NULL
    })
  ) %>%
  select(-has_24_months)


basic_results
```

## Validate basic results

### deterministic output
1. Generate the basic_results df, then run this to store it
```{r}
basic_results_old <- basic_results
```

2. Generate the basic_results df again, then run this to compare it with the previous run
```{r}
all.equal(basic_results, basic_results_old)
```

### fields
```{r}

df <- basic_results
row_count <- 120

# 1. Total Rows & Sex Distribution
stopifnot(
  "Row count must be exactly 120" = nrow(df) == row_count,
  "Must contain 60 Female objects" = sum(df$sex %in% c("Female", "f", "female")) == row_count / 2,
  "Must contain 60 Male objects"   = sum(df$sex %in% c("Male", "m", "male")) == row_count / 2
)

# 2. Key Uniqueness & Pair Counts
stopifnot(
  "Must have exactly 60 unique Ensembl IDs" = length(unique(df$ensembl_gene_id)) == row_count / 2,
  "Must have exactly 60 unique UniProt IDs" = length(unique(df$uniprotid)) == row_count / 2,
  "Must have exactly 60 unique Gene Symbols"= length(unique(df$gene_symbol)) == row_count / 2,
  "Gene symbols must all start with GeneX_" = startsWith(df$gene_symbol, "GeneX_")
)

# 3. String Construction Logic
# Compound values
stopifnot(
  "unique_id must equal ensg + uniprot" = df$unique_id == paste0(df$ensembl_gene_id, df$uniprotid),
  "display_symbol must equal gene_symbol + (uniprot)" = df$display_symbol == paste0(df$gene_symbol, " (", df$uniprotid, ")")
)

# Ensembl Gene ID Regex Format Validation
stopifnot(
  "All ensembl_gene_id values must match pattern ^ENSMUSG\\d{11}$" =
    grepl(ensembl_regex, df$ensembl_gene_id)
)

# Expected isoforms are present
stopifnot(
  "Must contain at least one '-1' isoform" = any(endsWith(df$uniprotid, "-1")),
  "Must contain at least one '-2' isoform" = any(endsWith(df$uniprotid, "-2")),
  "Must contain at least one '-3' isoform" = any(endsWith(df$uniprotid, "-3"))
)

# 4. M/F Pair Equality on Shared Metadata
pair_validation <- df %>%
  group_by(ensembl_gene_id) %>%
  dplyr::summarise(
    pair_count      = n(),
    has_both_sexes  = n_distinct(sex) == 2,
    same_uniprot    = n_distinct(uniprotid) == 1,
    same_symbol     = n_distinct(gene_symbol) == 1,
    same_tissue     = n_distinct(tissue) == 1,
    same_matched_ctrl = n_distinct(matched_control) == 1,
    same_biodomains = identical(biodomains[[1]], biodomains[[2]]),
    .groups = "drop"
  )

stopifnot(
  "Every gene must appear as an exact M/F pair" = pair_validation$pair_count == 2,
  "Every gene pair must contain two distinct sexes" = pair_validation$has_both_sexes,
  "UniProt IDs must match across M/F pair"       = pair_validation$same_uniprot,
  "Gene symbols must match across M/F pair"      = pair_validation$same_symbol,
  "Tissue values must match across M/F pair"     = pair_validation$same_tissue,
  "Biodomains lists must match across M/F pair"  = pair_validation$same_biodomains
)

# 5. Month Data Population & Missingness
stopifnot(
  "4 months must never be NULL"  = !purrr::map_lgl(df$`4 months`, is.null),
  "12 months must never be NULL" = !purrr::map_lgl(df$`12 months`, is.null),
  "18 months must never be NULL" = !purrr::map_lgl(df$`18 months`, is.null),
  "24 months must have exactly 18 NULL objects" = sum(purrr::map_lgl(df$`24 months`, is.null)) == 18
)

# 6. Biodomains Validity
stopifnot(
  "Biodomains list elements must not be empty" = purrr::map_int(df$biodomains, length) > 0,
  "Biodomains elements must be valid BIODOMAINS subsets" = unlist(df$biodomains) %in% BIODOMAINS
)

message("✅ All 6 validation suites passed successfully!")
```


# 2. rna_exists_multiple_proteins_results
* 15 pairs of objects (m/f with matching metadata and unique measurements)
* 1 gene : 3 uniprotid
* Some objects don't have '24 months' results
```{r}
ids <- 100:114
pad_number <- "9"

ensembl_id_in_rna_results <- c("ENSMUSG00000000001", "ENSMUSG00000000001", "ENSMUSG00000000001",
                               "ENSMUSG00000000056", "ENSMUSG00000000056", "ENSMUSG00000000056",
                               "ENSMUSG00000000214", "ENSMUSG00000000214", "ENSMUSG00000000214",
                               "ENSMUSG00000000296", "ENSMUSG00000000296", "ENSMUSG00000000296",
                               "ENSMUSG00000000340", "ENSMUSG00000000340", "ENSMUSG00000000340"
                               )
gene_symbol_in_rna_results <- c("Gnai3", "Gnai3", "Gnai3",
                                "Narf", "Narf", "Narf",
                                "Th", "Th", "Th",
                                "Tpd52l1", "Tpd52l1", "Tpd52l1",
                                "Dbt", "Dbt", "Dbt"
                                )

# pattern used to deterministically generate unique_ids, some with isoform suffixes: no suffix (66.7%), -1 (20%), -2 (6.7%), -3 (6.7%)
isoform_pattern <- c(
  rep("", 3), "-1",
  rep("", 3), "-1", "-2",
  rep("", 4), "-1", "-3"
)

object_dict <- tibble(

  ensembl_gene_id = ensembl_id_in_rna_results,
  gene_symbol = gene_symbol_in_rna_results,
  uniprotid = paste0(
    "P", gsub(" ", pad_number, sprintf("%5d", ids)),
    rep(isoform_pattern, length.out = length(ids))
  ),
  unique_id = paste0(ensembl_gene_id, uniprotid),
  display_symbol = paste0(gene_symbol, " (", uniprotid, ")")
)

# Populate fields that match across sexes: biodomains, name, matched_control, model_group, model_type, tissue

# generate link object for name field
name_link_object <- generate_link_object("name", link_url = paste0("/models/", MODEL_NAME), link_text = MODEL_NAME)[["name"]]

object_dict <- object_dict %>%
  mutate(
    biodomains = purrr::map(
      ensembl_gene_id,
      ~ {
        # Extract numeric digits from ensg as double float to prevent integer overflow
        gene_num <- as.numeric(gsub("\\D", "", .x))
        # rows for the same gene always get the same biodomains
        get_deterministic_biodomains(gene_num, BIODOMAINS)
      }
    ),
    name            = list(name_link_object),
    matched_control = MATCHED_CONTROL,
    model_group     = MODEL_GROUP,
    model_type      = MODEL_TYPE,
    tissue          = TISSUE

# Expand to include an object for each sex with identical metadata
  ) %>% crossing(sex = SEXES)


# populate row-specific heatmap data values

# pattern used to deterministically omit 24 month data (15% overall)
pattern_24m <- c(
  FALSE, FALSE,  # Gene Pair 1: Both missing
  FALSE, TRUE,   # Gene Pair 2: Female missing only
  TRUE,  FALSE,  # Gene Pair 3: Male missing only
  rep(TRUE, 24)  # Gene Pairs 4-15: Neither missing
)

# Generate row-specific heatmap data
rna_exists_multiple_proteins_results <- object_dict %>%
  dplyr::mutate(
    has_24_months = rep(pattern_24m, length.out = nrow(object_dict))
  ) %>%
  dplyr::mutate(
    # Populate timepoint data with deterministic values
    `4 months`  = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("4 months", paste0(.x, .y))[["4 months"]]),
    `12 months` = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("12 months", paste0(.x, .y))[["12 months"]]),
    `18 months` = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("18 months", paste0(.x, .y))[["18 months"]]),
    `24 months` = purrr::map2(has_24_months, paste0(ensembl_gene_id, sex), ~ {
      if (.x) generate_heatmap_object("24 months", .y)[["24 months"]] else NULL
    })
  ) %>%
  select(-has_24_months)


rna_exists_multiple_proteins_results
```


## Validation

### output is deterministic
1. Generate first file, then run this to store the output
```{r}
rna_exists_multiple_proteins_results_old <- rna_exists_multiple_proteins_results
```

2. Generate the file again, then run this to compare with the previous output
```{r}
all.equal(rna_exists_multiple_proteins_results, rna_exists_multiple_proteins_results_old)
```

## Validate fields
```{r}

df <- rna_exists_multiple_proteins_results
row_count <- 30

# 1. Total Rows & Sex Distribution
stopifnot(
  "Row count must contain expected #" = nrow(df) == row_count,
  "Must contain expected # Female objects" = sum(df$sex %in% c("Female", "f", "female")) == row_count/2,
  "Must contain expected # Male objects"   = sum(df$sex %in% c("Male", "m", "male")) == row_count/2
)

# 2. Key Uniqueness & Pair Counts
stopifnot(
  "Must have expected unique Ensembl IDs" = length(unique(df$ensembl_gene_id)) == length(unique(ensembl_id_in_rna_results)),
  "Must have expected unique UniProt IDs" = length(unique(df$uniprotid)) == length(unique(ensembl_id_in_rna_results))  * 3,
  "Must have expected unique Gene Symbols"= length(unique(df$gene_symbol)) == length(unique(ensembl_id_in_rna_results)),
  "Gene symbols match rna_results" = setequal(unique(df$gene_symbol), unique(gene_symbol_in_rna_results)),
  "Ensembl IDs match rna_results" = setequal(unique(df$ensembl_gene_id), unique(ensembl_id_in_rna_results))
)

# 3. String Construction Logic
# Compound values
stopifnot(
  "unique_id must equal ensg + uniprot" = df$unique_id == paste0(df$ensembl_gene_id, df$uniprotid),
  "display_symbol must equal gene_symbol + (uniprot)" = df$display_symbol == paste0(df$gene_symbol, " (", df$uniprotid, ")")
)

# Ensembl Gene ID Regex Format Validation
stopifnot(
  "All ensembl_gene_id values must match regex" = grepl(ensembl_regex, df$ensembl_gene_id)
)

# Expected isoforms are present
stopifnot(
  "Must contain at least one '-1' isoform" = any(endsWith(df$uniprotid, "-1")),
  "Must contain at least one '-2' isoform" = any(endsWith(df$uniprotid, "-2")),
  "Must contain at least one '-3' isoform" = any(endsWith(df$uniprotid, "-3"))
)

# 4a. Gene-Level Validation (Metadata across all 6 rows per gene)
gene_validation <- df %>%
  group_by(ensembl_gene_id) %>%
  dplyr::summarise(
    row_count         = n(),
    male_count        = sum(sex %in% c("Male", "m", "male")),
    female_count      = sum(sex %in% c("Female", "f", "female")),
    uniprot_count     = n_distinct(uniprotid),
    same_symbol       = n_distinct(gene_symbol) == 1,
    same_tissue       = n_distinct(tissue) == 1,
    same_matched_ctrl = n_distinct(matched_control) == 1,
    same_biodomains   = all(purrr::map_lgl(biodomains, ~ identical(.x, biodomains[[1]]))),
    .groups = "drop"
  )

# 4b. Isoform-Level Validation (1 M / 1 F pair per UniProt ID)
isoform_validation <- df %>%
  group_by(ensembl_gene_id, uniprotid) %>%
  dplyr::summarise(
    pair_count     = n(),
    has_both_sexes = n_distinct(sex) == 2,
    .groups = "drop"
  )

stopifnot(
  "Every gene must have exactly 6 rows"              = gene_validation$row_count == 6,
  "Every gene must contain 3 Male rows"              = gene_validation$male_count == 3,
  "Every gene must contain 3 Female rows"            = gene_validation$female_count == 3,
  "Every gene must have exactly 3 unique UniProt IDs"= gene_validation$uniprot_count == 3,
  "Gene symbols must match across all 6 rows"        = gene_validation$same_symbol,
  "Tissue values must match across all 6 rows"       = gene_validation$same_tissue,
  "Matched controls must match across all 6 rows"    = gene_validation$same_matched_ctrl,
  "Biodomains lists must match across all 6 rows"    = gene_validation$same_biodomains,
  "Every UniProt ID must appear as an exact pair (2 rows)" = isoform_validation$pair_count == 2,
  "Every UniProt ID pair must contain both sexes"          = isoform_validation$has_both_sexes
)

# 5. Month Data Population & Missingness
stopifnot(
  "4 months must never be NULL"  = !purrr::map_lgl(df$`4 months`, is.null),
  "12 months must never be NULL" = !purrr::map_lgl(df$`12 months`, is.null),
  "18 months must never be NULL" = !purrr::map_lgl(df$`18 months`, is.null),
  "24 months must have exactly 4 NULL objects" = sum(purrr::map_lgl(df$`24 months`, is.null)) == 4
)

# 6. Biodomains Validity
stopifnot(
  "Biodomains list elements must not be empty" = purrr::map_int(df$biodomains, length) > 0,
  "Biodomains elements must be valid BIODOMAINS subsets" = unlist(df$biodomains) %in% BIODOMAINS
)
```


3. no_rna_protein_exists_results
* 15 pairs of objects (m/f with matching metadata and unique measurements)
* 1 gene : 1 uniprotid
* Some objects don't have '24 months' results
```{r}
# Populate gene/protein fields: ensembl_gene_id, gene_symbol, uniprotid, unique_id, display_symbol
ids <- 10:24
pad_number <- "8"

# pattern used to deterministically generate unique_ids, some with isoform suffixes: no suffix (66.7%), -1 (20%), -2 (6.7%), -3 (6.7%)
isoform_pattern <- c(
  rep("", 3), "-1",
  rep("", 3), "-1", "-2",
  rep("", 4), "-1", "-3"
)

object_dict <- tibble(
  ensembl_gene_id = paste0(ENSEMBL_PREFIX, gsub(" ", pad_number, sprintf("%11d", ids))),
  gene_symbol = paste0("GeneP_", ids),
  uniprotid = paste0(
    "P", gsub(" ", pad_number, sprintf("%5d", ids)),
    rep(isoform_pattern, length.out = length(ids))
  ),
  unique_id = paste0(ensembl_gene_id, uniprotid),
  display_symbol = paste0(gene_symbol, " (", uniprotid, ")")
)

# Populate fields that match across sexes: biodomains, name, matched_control, model_group, model_type, tissue

# generate link object for name field
name_link_object <- generate_link_object("name", link_url = paste0("/models/", MODEL_NAME), link_text = MODEL_NAME)[["name"]]

object_dict <- object_dict %>%
  mutate(
    biodomains = purrr::map(
      ensembl_gene_id,
      ~ {
        # Extract numeric digits from ensg as double float to prevent integer overflow
        gene_num <- as.numeric(gsub("\\D", "", .x))
        # rows for the same gene always get the same biodomains
        get_deterministic_biodomains(gene_num, BIODOMAINS)
      }
    ),
    name            = list(name_link_object),
    matched_control = MATCHED_CONTROL,
    model_group     = MODEL_GROUP,
    model_type      = MODEL_TYPE,
    tissue          = TISSUE

# Expand to include an object for each sex with identical metadata
  ) %>% crossing(sex = SEXES)


# populate row-specific heatmap data values

# pattern used to deterministically omit 24 month data (15% overall)
pattern_24m <- c(
  FALSE, FALSE,  # Gene Pair 1: Both missing
  FALSE, TRUE,   # Gene Pair 2: Female missing only
  TRUE,  FALSE,  # Gene Pair 3: Male missing only
  rep(TRUE, 24)  # Gene Pairs 4-15: Neither missing
)

# Generate row-specific heatmap data
no_rna_protein_exists_results <- object_dict %>%
  dplyr::mutate(
    has_24_months = rep(pattern_24m, length.out = nrow(object_dict))
  ) %>%
  dplyr::mutate(
    # Populate timepoint data with deterministic values
    `4 months`  = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("4 months", paste0(.x, .y))[["4 months"]]),
    `12 months` = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("12 months", paste0(.x, .y))[["12 months"]]),
    `18 months` = purrr::map2(ensembl_gene_id, sex, ~ generate_heatmap_object("18 months", paste0(.x, .y))[["18 months"]]),
    `24 months` = purrr::map2(has_24_months, paste0(ensembl_gene_id, sex), ~ {
      if (.x) generate_heatmap_object("24 months", .y)[["24 months"]] else NULL
    })
  ) %>%
  select(-has_24_months)

no_rna_protein_exists_results
```
## Validation

### output is deterministic
1. Generate first file, then run this to store the output
```{r}
no_rna_protein_exists_results_old <- no_rna_protein_exists_results
```

2. Generate the file again, then run this to compare with the previous output
```{r}
all.equal(no_rna_protein_exists_results, no_rna_protein_exists_results_old)
```

## Validate fields
```{r}
df <- no_rna_protein_exists_results
row_count <- 30

# 1. Total Rows & Sex Distribution
stopifnot(
  "Row count must contain expected #" = nrow(df) == row_count,
  "Must contain expected # Female objects" = sum(df$sex %in% c("Female", "f", "female")) == row_count/2,
  "Must contain expected # Male objects"   = sum(df$sex %in% c("Male", "m", "male")) == row_count/2
)

# 2. Key Uniqueness & Pair Counts
stopifnot(
  "Must have exactly 15 unique Ensembl IDs" = length(unique(df$ensembl_gene_id)) == row_count / 2,
  "Must have exactly 15 unique UniProt IDs" = length(unique(df$uniprotid)) == row_count / 2,
  "Must have exactly 15 unique Gene Symbols"= length(unique(df$gene_symbol)) == row_count / 2
)

# 3. String Construction Logic
stopifnot(
  "unique_id must equal ensg + uniprot" = df$unique_id == paste0(df$ensembl_gene_id, df$uniprotid),
  "display_symbol must equal gene_symbol + (uniprot)" = df$display_symbol == paste0(df$gene_symbol, " (", df$uniprotid, ")")
)

# Ensembl Gene ID Regex Format Validation
stopifnot(
  "All ensembl_gene_id values must match pattern ^ENSMUSG\\d{11}$" =
    all(grepl(ensembl_regex, df$ensembl_gene_id))
)

# Expected isoforms are present
stopifnot(
  "Must contain at least one '-1' isoform" = any(endsWith(df$uniprotid, "-1")),
  "Must contain at least one '-2' isoform" = any(endsWith(df$uniprotid, "-2")),
  "Must contain at least one '-3' isoform" = any(endsWith(df$uniprotid, "-3"))
)

# 4. M/F Pair Equality on Shared Metadata
pair_validation <- df %>%
  group_by(ensembl_gene_id) %>%
  dplyr::summarise(
    pair_count      = n(),
    has_both_sexes  = n_distinct(sex) == 2,
    same_uniprot    = n_distinct(uniprotid) == 1,
    same_symbol     = n_distinct(gene_symbol) == 1,
    same_tissue     = n_distinct(tissue) == 1,
    same_matched_ctrl = n_distinct(matched_control) == 1,
    same_biodomains = identical(biodomains[[1]], biodomains[[2]]),
    .groups = "drop"
  )

stopifnot(
  "Every gene must appear as an exact M/F pair"     = pair_validation$pair_count == 2,
  "Every gene pair must contain two distinct sexes" = pair_validation$has_both_sexes,
  "UniProt IDs must match across M/F pair"          = pair_validation$same_uniprot,
  "Gene symbols must match across M/F pair"         = pair_validation$same_symbol,
  "Tissue values must match across M/F pair"        = pair_validation$same_tissue,
  "Biodomains lists must match across M/F pair"     = pair_validation$same_biodomains
)

# 5. Month Data Population & Missingness
stopifnot(
  "4 months must never be NULL"  = !purrr::map_lgl(df$`4 months`, is.null),
  "12 months must never be NULL" = !purrr::map_lgl(df$`12 months`, is.null),
  "18 months must never be NULL" = !purrr::map_lgl(df$`18 months`, is.null),
  "24 months must have exactly 4 NULL objects" = sum(purrr::map_lgl(df$`24 months`, is.null)) == 4
)

# 6. Biodomains Validity
stopifnot(
  "Biodomains list elements must not be empty" = purrr::map_int(df$biodomains, length) > 0,
  "Biodomains elements must be valid BIODOMAINS subsets" = unlist(df$biodomains) %in% BIODOMAINS
)
```

# Assemble the collection

```{r}
# *****************************
# Generate protein_de_aggregate
# *****************************

json_list <- list(
  basic_results,
  rna_exists_multiple_proteins_results,
  no_rna_protein_exists_results
) %>%
  list_rbind() %>%
  purrr::transpose() %>%
  # Since we are using a tibble, we need to do some extra stuff to omit null keys entirely
  purrr::map(~ purrr::discard(.x, is.null))

json_output <- toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
# cat(json_output)

# write json file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

write(json_output, file.path(output_dir, "protein_de_aggregate.json"))
```


# Upload to synapse
*********************************************************
Validate ./output/protein_de_aggregate.json before upload
*********************************************************
```{r}
synLogin()
syn_file <- File(
              file.path(output_dir, "protein_de_aggregate.json"),
              description = "mock protein_de_aggregate.json for Model AD 1.1 development",
              parent = "syn51498091"
            )

res <- synStore(
          syn_file,
          forceVersion = FALSE,
          executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/data_mocks/mock_protein_de_aggregate.rmd"
        )
# print synapse_id and new version number
print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```


-----
TESTS
-----

# sanity check results
```{r}
all_mocks <- rbind(
  basic_results,
  rna_exists_multiple_proteins_results,
  no_rna_protein_exists_results
)

stopifnot(
 nrow(all_mocks) == 120 + 30 + 30,
# ensure no collisions across collections
 length(unique(all_mocks$ensembl_gene_id)) == 60 + 5 + 15,
 length(unique(all_mocks$gene_symbol)) == 60 + 5 + 15,
length(unique(all_mocks$uniprotid)) == 60 + 15 + 15
)

```

# test generate heatmap object
```{r}
object <- generate_heatmap_object("12 months")

json_list <- c(
  object
)

json_output <- toJSON(json_list, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)

```


# test generate link object
```{r}
object <- generate_link_object("name", "/models/LOAD2", "LOAD2")

json_list <- c(
  object
)

json_output <- toJSON(json_list, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)
```
# test get deterministic biodomains
```{r}
biodomains <- get_deterministic_biodomains(345, BIODOMAINS)
biodomains2 <- get_deterministic_biodomains(345, BIODOMAINS)
stopifnot(
  "expect same biodomains if same numeric" = biodomains == biodomains2
)


biodomains3 <- get_deterministic_biodomains(10, BIODOMAINS)
biodomains4 <- get_deterministic_biodomains(10, BIODOMAINS)
stopifnot(
  "expect same biodomains if same numeric" = biodomains3 == biodomains4
)

```
