---
title: "Create xQTL ui_config source file"
related issues: QTL-104
---
This notebook generates the ui_config.json source file (syn77129966) containing the data required to construct Comparison Tool interfaces for the xQTL Explorer.

Interfaces that are currently handled by this notebook:
* eQTL CT

This notebook must be updated when the Explorer is updated to:
* Add a new CT interface
* Add or modify a column of an existing CT interface
* Add or modify a CT dropdown menu or dropdown value
* Add or modify a filterset or filter value


Notes:
* `query_param_key`: Values must be plural, camelCase, and should intelligibly align with the related display values presented to users on the page:
   - filter.name (filter panels)
   - filter.short_name (filter chiclets)
   - column.name (column header)


## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}
if (!require("jsonlite", quietly = TRUE)) { install.packages("jsonlite") }
if (!require("purrr")) { install.packages("purrr") }
if (!require("stringr")) { install.packages("stringr") }

library(synapser)
library(purrr)
library(jsonlite)
library(stringr)
```


## Variables and functions
Sources the required variables and functions.
```{r}
# *******************************
# STATIC ROW COUNTS (MG-384)
# *******************************
# we don't need row_count values CT loading indicators since we've introduced server-side pagination
eqtl_row_count = NA

# *********
# FUNCTIONS
# *********

# Builds a ui_config object from a dataframe containing column info
# - eqtl_columns: a csv-compatible dataframe containing a row of values per CT column:
#    - name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden, column_width
# eqtl_filters: a dataframe containing a row of values per CT filterset:
#    - name, data_key, short_name, query_param_key
# eqtl_filter_values: A list of lists of filter values; value lists must be in the same order as eqtl_filters
#   - list(c(eqtl_filters[1]_value list here), c(eqtl_filters[2]_value list here), ...)
build_eqtl_object <- function(page, eqtl_columns, eqtl_filters, eqtl_filter_values) {
  columns <- build_columns(eqtl_columns)
  filters <- build_filters(eqtl_filters, eqtl_filter_values)

  # Assemble the ui_config object
  list(
    page = page,
    dropdowns = list(),
    row_count = eqtl_row_count,
    columns = columns,
    filters = filters
  )
}


build_columns <- function(page_column_data) {

  columns <- purrr::pmap(page_column_data, build_column)
  names(columns) <- NULL
  columns
}

# Builds a column object
# - name: The column display name
#   -- Specify NA for hidden columns; if you specify a different value, it will be ignored
# - type: The column type
# - tooltip: The tooltip display value for the column name
#   -- Specify NA for column names that need no explanatory tooltip, and for hidden columns
# - sort_tooltip: The tooltip display value for the column sort indicator
#   -- This key is always omitted for columns with the primary type, and for hidden columns
#   -- Specify NA to use the default sort_tooltip value for any other column: "Sort by <name> value"
# - link_url: The optional default target URL for columns with a link_* type
#   -- Specify NA for columns with other types
# - link_text: The optional default link display text for columns with a link_* type
#   -- Specify NA for columns with other types
# - is_exported: Whether this column should be included in CSV data exports
# - is_hidden: Whether this column should be displayed in the CT table
# - column_width: Optional override that specifies the column's width, in pixels
#   -- By default, the column width is automatically calculated
#   -- Specify a numeric value to specify a fixed width for columns with variable content length / auto widths
build_column <- function(name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden, column_width = NA) {

  # Use the default sort_tooltip if one isn't specified, for non-hidden columns only
  if (is_hidden == TRUE) { sort_tooltip = NA }
  else { sort_tooltip = if (!is.na(sort_tooltip)) sort_tooltip else paste("Sort by", name, "value") }

  column <- list(
    name = name,
    type = type,
    data_key = data_key,
    tooltip = tooltip,
    sort_tooltip = sort_tooltip,
    link_url = link_url ,
    link_text = link_text,
    is_exported = is_exported,
    is_hidden = is_hidden,
    column_width = column_width
  )

  # Omit keys with NA values
  clean_column <- purrr::discard(column, ~ is.null(.x) | all(is.na(.x)))
  clean_column
}

# Builds a list of filter definitions
# - filter_defs: A dataframe of filterset info:
#   - name, data_field, short_name, query_param_key, filter_values
# - filter_vals: A list of lists specifying the filter values for each filterset
build_filters <- function(filter_defs, filter_values) {

  # Construct list of filters
  filters <- Map(function(name, data_key, short_name, query_param_key, filter_values) {

    filter <- list(
      name = name,
      data_key = data_key,
      short_name = short_name,
      query_param_key = query_param_key,
      values = filter_values
      )

      # Omit keys with NA values
      filter <- discard(filter, ~ is.null(.x) | all(is.na(.x)))

    }, filter_defs$name, filter_defs$data_key, filter_defs$short_name, filter_defs$query_param_key, filter_values)

  # Remove names to ensure JSON is an array, not object
  names(filters) <- NULL
  filters
}


# Vector to store all downloaded synIds for Synapse provenance (populated via download_file function)
data_sources <- character()

# Download a file from Synapse by synId and read it into a dataframe
# - synId: The synapseId of the file to download
# - name: The name to assign to the resulting dataframe or vector
# - type: The type of file being downloaded
#   -- Supports: tsv, csv, json, txt
download_file <- function(synId, df_name, type) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("./", "input")
  id_parts <- strsplit(synId, '.', fixed = TRUE)
  id_value <- id_parts[[1]][1]
  version_value <- id_parts[[1]][2]

  if (is.na(version_value)) {
    version_value <- NULL
  }

  cat(paste0("Downloading ", synId, "..."))
  file <- synGet(id_value, version = version_value, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId, "\n- Modified on ", file$properties$modifiedOn, "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  if (type == 'txt') {
    name <- readLines(path)
  }
  else {
    if (type == 'json') { data <- fromJSON(path) }
    if (type == 'csv' || type == 'table') { data <- read.csv(path, na.strings = c("", "NA")) }
    if (type == 'tsv') { data <- read.csv(path, sep = "\t") }
    name <- as.data.frame(data)
  }
  name
}

```

# Generate ui_config.json

```{r}
# download source files from synapse
columns <- download_file("syn77143572", columns ,"csv")
filters <- download_file("syn77143748", filters ,"csv")
labels <- download_file("syn77171904", labels, "csv")

# extract filter values from display label mapping file
cell_class_values <- str_sort(unique(labels$cell_class))
cell_subclass_values <- str_sort(labels$cell_subclass)
filter_vals <- list(
  c('rs12345', 'rs67890'), # temp placeholder
  c('geneA', 'geneB'),     # temp placeholder
  cell_class_values,
  cell_subclass_values
  )


json_list <- list(
  build_eqtl_object("eQTLs", columns, filters, filter_vals)
)

json_output <- toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
# cat(json_output)

# write json file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

write(json_output, file.path(output_dir, "ui_config.json"))
```


# Upload to synapse
**********************************************
Validate ./output/ui_config.json before upload
**********************************************
```{r}
synLogin()
syn_file <- File(
              file.path(output_dir, "ui_config.json"),
              description = "ui_config.json for xQTL Explorer",
              parent = "syn75312280"
            )

res <- synStore(
          syn_file,
          forceVersion = FALSE,
          used = data_sources,
          executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/xqtl/notebooks/xqtl_ui_config.rmd"
        )
# print synapse_id and new version number
print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```


-----
TESTS
-----

## Test download_file
```{r}
download_file("syn77140300", eqtl_columns, "csv")
```