---
title: "AG-1974: OpenTargets Drug Data"
---
This notebook pulls drug data from the OpenTargets GraphQL API and cleans it for use by downstream transforms:
- nomianted_drugs
- drug_info

## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
- purrr
- ghql
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}
if (!require("jsonlite", quietly = TRUE)) { install.packages("jsonlite") }
if (!require("purrr")) { install.packages("purrr") }
if (!require("ghql")) { install.packages("ghql") }

library(synapser)
library(purrr)
library(jsonlite)
library(ghql)
```


## Set up clients, variables and functions
Setup the required graphQL client and nominated drugs source file
```{r}
# ***************************
# OPEN TARGETS GRAPHQL CLIENT
# ***************************

url <- "https://api.platform.opentargets.org/api/v4/graphql"
con <- GraphqlClient$new(url = url)

# **************************
# PROVENANCE & LOCAL STORAGE
# **************************

# Vector to store all downloaded synIds for Synapse provenance (populated via download_csv function)
data_sources <- character()

# Local file storage
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

# *********
# FUNCTIONS
# *********

# Combines the provided lists of values to produce a set of unique values.
# - name: The canonical name of the drug, to removed from the list of alias values
# - ...: A variable number of lists of values to combine
resolve_drug_names <- function(name, ...) {
  combined <- unlist(list(...))
  combined <- combined[!is.na(combined) & combined != ""]

  # Remove the canonical name
  filtered <- combined[tolower(combined) != tolower(name)]

  # Case-insensitive deduplication, keeping original casing of the first occurrence found
  final_list <- filtered[!duplicated(tolower(filtered))]

  # Return as a list (or character(0) if empty for clean JSON)
  return(if(length(final_list) > 0) final_list else list())
}

# Download a csv file from Synapse by synId and read it into a dataframe
# - synId: synapse ID of the file to download
# - df_name: the name to use for the resulting dataframe
download_csv <- function(synId, df_name) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("./", "input")
  id_parts <- strsplit(synId, '.', fixed = TRUE)
  id_value <- id_parts[[1]][1]
  version_value <- id_parts[[1]][2]

  if (is.na(version_value)) {
    version_value <- NULL
  }

  cat(paste0("Downloading ", synId, "..."))

  file <- synGet(id_value, version = version_value, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId, "\n- Modified on ", file$properties$modifiedOn, "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  as.data.frame(read.csv(path))
}

# ***************
# DRUG TARGET IDs
# ***************

# Get the list of chembl_ids from harmonized_drugs.csv
drug_targets <- download_csv("syn73677240.6")
chembl_ids <- drug_targets$chembl_id
```

# Download and adapt OpenTargets Drug data for nominated drugs
```{r}
# construct the GraphQL query
drug_query <- Query$new()
variables = list(chembl_ids = chembl_ids)

drug_query$query('drug_info', 'query getDrugs($chembl_ids: [String!]!){
  drugs (chemblIds: $chembl_ids){
    id
    description
    drugType
    maximumClinicalTrialPhase
    yearOfFirstApproval
    mechanismsOfAction {
      rows {
        mechanismOfAction
      }
    }
    linkedTargets {
      rows {
        id
      }
    }
    name
    crossReferences {
      source
      ids
    }
    synonyms
    tradeNames
  }
}')

#Execute the GraphQL query
result <- con$exec(drug_query$queries$drug_info, variables)

# Uncomment to write response data to local file
#write(prettify(result), file.path(output_dir, "drug_info_response"))

# Read the JSON response into a dataframe
response_df <- fromJSON(result, flatten = TRUE)
drug_info <- response_df$data$drugs
drug_info

# clean up nested row structures
drug_info <- drug_info %>%
  mutate(
mechanismsOfAction = map(mechanismsOfAction.rows, ~ unlist(.x$mechanismOfAction)),
    linkedTargets = map(linkedTargets.rows, ~ unlist(.x$id))
  ) %>%
  # have to use - instead of ! to negate columns with . characters
  select(-mechanismsOfAction.rows, -linkedTargets.rows)

# Extract DrugBank ID from crossReferences
drugbank_mapping <- drug_info %>%
  select(id, crossReferences) %>%
  unnest(crossReferences) %>%
  filter(source == "drugbank") %>%
  unnest(ids) %>%
  select(id, drug_bank_id = ids)

drug_info <- drug_info %>%
  left_join(drugbank_mapping, by = "id") %>%
  select(-crossReferences)

# manually resolve missing DrugBank IDs
manual_fixes <- tribble(
  ~id,             ~drug_bank_id_fix,
  "CHEMBL4435170",   "DB16650",
  "CHEMBL4594217",   "DB08907",
  "CHEMBL4084119",   "DB06655",
  "CHEMBL4788758",   "DB19277",
  "CHEMBL545437",    "DB05395"
)

# Populate the missing DrugBank IDs
drug_info <- drug_info %>%
  left_join(manual_fixes, by = "id") %>%
  mutate(drug_bank_id = coalesce(drug_bank_id, drug_bank_id_fix)) %>%
  select(-drug_bank_id_fix)

# Clean up values and align colnames
drug_info <- drug_info %>%

  # Collapse synonyms and tradeNames into a single set (case-insensitive)
  mutate(aliases = pmap(
    .l = list(name, synonyms, tradeNames),
    .f = resolve_drug_names
  )) %>%
  select(-tradeNames, -synonyms, -name) %>%

  # If maximumClinicalTrialPhase == -1, convert to NA to avoid magic numbers
  mutate(maximumClinicalTrialPhase = na_if(maximumClinicalTrialPhase, -1)) %>%

  # rename columns
  dplyr::rename(chembl_id = id) %>%
  dplyr::rename(drug_type = drugType) %>%
  dplyr::rename(maximum_clinical_trial_phase = maximumClinicalTrialPhase) %>%
  dplyr::rename(year_of_first_approval = yearOfFirstApproval) %>%
  dplyr::rename(mechanisms_of_action = mechanismsOfAction) %>%
  dplyr::rename(linked_targets = linkedTargets)

drug_info
```

# Generate source file and upload to Synapse
```{r}
json_output <- toJSON(drug_info, pretty = TRUE, auto_unbox = FALSE)

# write json file
write(json_output, file.path(output_dir, "opentargets_drug_data.json"))

# BREAK HERE FOR LOCAL VALIDATION

# upload to synapse
synLogin()
syn_file <- File(
              file.path(output_dir, "opentargets_drug_data.json"),
              description = "Open Targets Drug data for Agora",
              parent = "syn7525089"
            )

res <- synStore(
          syn_file,
          forceVersion = FALSE,
          used = data_sources,
          executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/AG-1974:_OpenTarget_ Drug_Data.rmd"
        )
# print synapse_id and new version number
print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))
```

