---
title: "AG-1974: OpenTargets Drug Metadata"
---
This notebook pulls drug metadata from the OpenTargets GraphQL API and prepares it for use by downstream transforms:
- nominated_drugs
- drug_info

## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
- purrr
- ghql
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}
if (!require("jsonlite", quietly = TRUE)) { install.packages("jsonlite") }
if (!require("purrr")) { install.packages("purrr") }
if (!require("ghql")) { install.packages("ghql") }

library(synapser)
library(purrr)
library(jsonlite)
library(ghql)
```



## Set up clients, variables and functions
Setup the required graphQL client and get nominated drug ids
```{r}
# ***************************
# OPEN TARGETS GRAPHQL CLIENT
# ***************************

url <- "https://api.platform.opentargets.org/api/v4/graphql"
con <- GraphqlClient$new(url = url)

# **************************
# PROVENANCE & LOCAL STORAGE
# **************************

# Vector to store all downloaded synIds for Synapse provenance (populated via download_csv function)
data_sources <- character()

# Local file storage
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

# *********
# FUNCTIONS
# *********

# Combines the provided lists of values to produce a set of unique values.
# - name: The canonical name of the drug, to removed from the list of alias values
# - ...: A variable number of lists of values to combine
resolve_drug_names <- function(name, ...) {
  combined <- unlist(list(...))
  combined <- combined[!is.na(combined) & combined != ""]

  # Remove the canonical name
  filtered <- combined[tolower(combined) != tolower(name)]

  # Case-insensitive deduplication, keeping original casing of the first occurrence found
  final_list <- filtered[!duplicated(tolower(filtered))]

  # Return as a list (or character(0) if empty for clean JSON)
  return(if(length(final_list) > 0) final_list else list())
}

# Download a csv file from Synapse by synId and read it into a dataframe
# - synId: synapse ID of the file to download
# - df_name: the name to use for the resulting dataframe
download_csv <- function(synId) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("./", "input")
  cat(paste0("Downloading ", synId, "..."))

  file <- synGet(synId, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId,
      "\n- Modified on ", format(as.POSIXct(file$properties$modifiedOn, format="%Y-%m-%dT%H:%M:%OS"), "%b %d %H:%M:%S %Y"),
      "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  as.data.frame(read.csv(path))
}

# ***************
# DRUG TARGET IDs
# ***************

# Get the list of chembl_ids from harmonized_drugs.csv
drug_targets <- download_csv("syn73677240") # TODO pin this version once finalized?
chembl_ids <- unique(drug_targets$chembl_id) # 67 ids
```

# Download and adapt OpenTargets drug metadata for nominated drugs
```{r}
# construct the GraphQL query
drug_query <- Query$new()
variables = list(chembl_ids = chembl_ids)

drug_query$query('drug_info', 'query getDrugs($chembl_ids: [String!]!){
  drugs (chemblIds: $chembl_ids){
    id
    description
    drugType
    maximumClinicalTrialPhase
    yearOfFirstApproval
    mechanismsOfAction {
      rows {
        mechanismOfAction
      }
    }
    linkedTargets {
      rows {
        id
      }
    }
    name
    crossReferences {
      source
      ids
    }
    synonyms
    tradeNames
  }
}')

#Execute the GraphQL query
result <- con$exec(drug_query$queries$drug_info, variables)

# Uncomment to write response data to local file
#write(prettify(result), file.path(output_dir, "drug_info_response"))

# Read the JSON response into a dataframe
response_df <- fromJSON(result, flatten = TRUE)
drug_info <- response_df$data$drugs
# drug_info

# clean up nested row structures
drug_info <- drug_info %>%
  mutate(
mechanismsOfAction = map(mechanismsOfAction.rows, ~ unlist(.x$mechanismOfAction)),
    linkedTargets = map(linkedTargets.rows, ~ unlist(.x$id))
  ) %>%
  # have to use - instead of ! to negate columns with . characters
  select(-mechanismsOfAction.rows, -linkedTargets.rows)

# Extract DrugBank ID from crossReferences
drugbank_mapping <- drug_info %>%
  select(id, crossReferences) %>%
  unnest(crossReferences) %>%
  filter(source == "drugbank") %>%
  unnest(ids) %>%
  select(id, drug_bank_id = ids)

drug_info <- drug_info %>%
  left_join(drugbank_mapping, by = "id") %>%
  select(-crossReferences)

# missing values were manually resolved from DrugBank
drugbank_manual_fixes <- tribble(
  ~id,             ~drug_bank_id_fix,
  "CHEMBL4435170",   "DB16650",
  "CHEMBL4594217",   "DB08907",
  "CHEMBL4084119",   "DB06655",
  "CHEMBL4788758",   "DB19277",
  "CHEMBL545437",    "DB05395"
)

# Populate the missing DrugBank IDs
drug_info <- drug_info %>%
  left_join(drugbank_manual_fixes, by = "id") %>%
  mutate(drug_bank_id = coalesce(drug_bank_id, drug_bank_id_fix)) %>%
  select(-drug_bank_id_fix)

# Clean up values and align colnames
drug_info <- drug_info %>%

  # Collapse synonyms and tradeNames into a single set (case-insensitive)
  mutate(aliases = pmap(
    .l = list(name, synonyms, tradeNames),
    .f = resolve_drug_names
  )) %>%

  select(-tradeNames, -synonyms, -name) %>%

  # If maximumClinicalTrialPhase == -1, convert to NA to avoid magic numbers
  mutate(maximumClinicalTrialPhase = na_if(maximumClinicalTrialPhase, -1)) %>%

  # collapse ensembl_gene_id and mechanisms_of_action lists to unique values
  mutate(linkedTargets = map(linkedTargets, unique)) %>%
  mutate(mechanismsOfAction = map(mechanismsOfAction, unique)) %>%

  # rename columns
  dplyr::rename(chembl_id = id) %>%
  dplyr::rename(modality = drugType) %>%
  dplyr::rename(maximum_clinical_trial_phase = maximumClinicalTrialPhase) %>%
  dplyr::rename(year_of_first_approval = yearOfFirstApproval) %>%
  dplyr::rename(mechanisms_of_action = mechanismsOfAction) %>%
  dplyr::rename(linked_targets = linkedTargets)

  # Add minimal rows for chembl_ids not available on OT
  manual_entries <- data.frame(
    chembl_id = c("CHEMBL449520", "CHEMBL3970532", "CHEMBL293762",
                  "CHEMBL71851", "CHEMBL338802", "CHEMBL51934"),
    maximum_clinical_trial_phase = c(NA, NA, NA, NA, 4, 2),
    modality = c("Small molecule", "Small molecule", "Small molecule",
               "Small molecule", "Small molecule", "Small molecule"),
    year_of_first_approval = c(NA, NA, NA, NA, 1941, NA),
    stringsAsFactors = FALSE
  )
  drug_info <- bind_rows(drug_info, manual_entries)

  # Ensure empty list fields serialize as empty lists in JSON
  list_cols <- c("mechanisms_of_action", "linked_targets", "aliases")
  for (col in list_cols) {
    # Ensure the column is a list
    if (!is.list(drug_info[[col]])) {
      drug_info[[col]] <- as.list(drug_info[[col]])
    }

    # Find indices that are NULL or empty, and overwrite with list(character(0))
    is_empty <- sapply(drug_info[[col]], function(x) is.null(x) || length(x) == 0 || all(is.na(x)))
    drug_info[[col]][is_empty] <- list(character(0))
}

drug_info # 65 rows - missing 2 are the ones with 2 chembl_ids
```

# Generate source file and upload to Synapse
```{r}
json_output <- toJSON(drug_info, pretty = TRUE, auto_unbox = FALSE)

# write json file
write(json_output, file.path(output_dir, "opentargets_drug_metadata.json"))

# BREAK HERE FOR LOCAL VALIDATION

# upload to synapse
synLogin()
syn_file <- File(
              file.path(output_dir, "opentargets_drug_metadata.json"),
              description = "Open Targets Drug data for Agora",
              parent = "syn7525089"
            )

res <- synStore(
          syn_file,
          forceVersion = FALSE,
          used = data_sources,
          executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/AG-1974_OpenTargets_Drug_Metadata.rmd"
        )
# print synapse_id and new version number
print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))
```

