---
title: "Agora ui_config source file"
---
This notebook generates the ui_config.json source file (syn73677126) containing the data required to construct Comparison Tool interfaces for Agora.

Interfaces that are currently handled by this notebook:
* Nominated Targets
* Nominated Drugs

This notebook must be updated when the Explorer is updated to:
* Add a new CT interface
* Add or modify a column of an existing CT interface
* Add or modify a CT dropdown menu or dropdown value
* Add or modify a filterset or filter value


Notes:
* `query_param_key`: Values must be camelCase, should be plural, and should intelligibly align with the related display
values presented to users on the page.
  Example:
        column.name: ""Pharos Class"
        filter.name: "Pharos Class", <--- the filter's display name in the Filter panel
        filter.short_name: "Pharos", <-- the filter's short display name in the "applied filter" chiclet
        filter.query_param_key: "pharosClasses", <-- the filter's query parameter key is (plural, camelCase)

## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}
if (!require("jsonlite", quietly = TRUE)) { install.packages("jsonlite") }
if (!require("purrr")) { install.packages("purrr") }

library(synapser)
library(purrr)
library(jsonlite)
```


## Variables and functions
Sources the required variables and functions.
```{r}
# *********
# FUNCTIONS
# *********

# Builds a ui_config object from a dataframe containing column info
# - page_column_data: a csv-compatible dataframe containing a row of values per CT column:
#    - name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden
build_object <- function(page_column_data) {

  columns <- purrr::pmap(page_column_data, build_column)
  names(columns) <- NULL
  columns
}

# Builds a column object
# - name: The column display name
#   -- Specify NA for hidden columns; if you specify a different value, it will be ignored
# - type: The column type
# - tooltip: The tooltip display value for the column name
#   -- Specify NA for column names that need no explanatory tooltip, and for hidden columns
# - sort_tooltip: The tooltip display value for the column sort indicator
#   -- This key is always omitted for columns with the primary type, and for hidden columns
#   -- Specify NA to use the default sort_tooltip value for any other column: "Sort by <name> value"
# - link_url: The optional default target URL for columns with a link_* type
#   -- Specify NA for columns with other types
# - link_text: The optional default link display text for columns with a link_* type
#   -- Specify NA for columns with other types
# - is_exported: Whether this column should be included in CSV data exports
# - is_hidden: Whether this column should be displayed in the CT table
# - column_width: Optional override that specifies the column's width, in pixels
#   -- Specify NA to use the default behavior of automatically calculating the column width
build_column <- function(name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden, column_width) {

  # Use the default sort_tooltip if one isn't specified, for non-hidden columns only
  if (is_hidden == TRUE) { sort_tooltip = NA }
  else { sort_tooltip = if (!is.na(sort_tooltip)) sort_tooltip else paste("Sort by", name, "value") }

  column <- list(
    name = name,
    type = type,
    data_key = data_key,
    tooltip = tooltip,
    sort_tooltip = sort_tooltip,
    link_url = link_url ,
    link_text = link_text,
    is_exported = is_exported,
    is_hidden = is_hidden,
    column_width = column_width
  )

  # Omit keys with NA values
  clean_column <- purrr::discard(column, ~ is.null(.x) | all(is.na(.x)))
  clean_column
}

# Builds a list of filter definitions
# - filter_defs: A dataframe of filterset info:
#   - name, data_field, short_name, query_param_key, filter_values
# - filter_vals: A list of lists specifying the filter values for each filterset
build_filters <- function(filter_defs, filter_values) {

  # Construct list of filters
  filters <- Map(function(name, data_key, short_name, query_param_key, filter_values) {

    filter <- list(
      name = name,
      data_key = data_key,
      short_name = short_name,
      query_param_key = query_param_key,
      values = filter_values
      )

      # Omit keys with NA values
      filter <- discard(filter, ~ is.null(.x) | all(is.na(.x)))

    }, filter_defs$name, filter_defs$data_key, filter_defs$short_name, filter_defs$query_param_key, filter_values)

  # Remove names to ensure JSON is an array, not object
  names(filters) <- NULL
  filters
}


# *****************
# Nominated Targets
# *****************
target_columns <- data.frame(
  name = c("Gene Symbol", NA, "Nominations",
           "First Nominated", "Nominating Teams", "Cohorts",
           "Data Used", "Programs", "Pharos Class"
  ),
  type = c("primary", "text", "numeric",
           "numeric", "text", "text",
           "text","text", "text"
  ),
  data_key = c("hgnc_symbol", "ensembl_gene_id", "total_nominations",
               "initial_nomination", "nominating_teams", "cohort_studies",
               "input_data", "programs", "pharos_class"),
  tooltip = c(rep(NA, times = 9)),
  sort_tooltip = rep(NA, times = 9),
  link_url = rep(NA, times = 9),
  link_text = rep(NA, times = 9),
  is_exported = c(rep(TRUE, times = 9)),
  is_hidden = c(FALSE, TRUE, rep(FALSE, times = 7)),
  column_width = c(NA, NA, NA, NA, 300, NA, NA, NA, NA),
  stringsAsFactors = FALSE
)

# Nominated Target Filters
# ----------------------
target_filters <- data.frame(
  name = c('Cohort', 'Data Used', 'First Nominated', 'Nominating Team',
           'Nominations', 'Pharos Class', 'Program'),
  data_key = c('cohort_studies', 'input_data', 'initial_nomination', 'nominating_teams',
               'total_nominations', 'pharos_class', 'programs'),
  short_name = c('Cohort', 'Data', 'First', 'Team',
                 'Nominations', 'Pharos', 'Program'),
  query_param_key = c('cohorts', 'data', 'firstNominations', 'teams',
                      'nominations', 'pharosClasses', 'programs'),
  stringsAsFactors = FALSE
)

# Nominated Target Filter Values
# ------------------------------
targets_data_cohort_filter_values <- c(
  "ABA", "ACT", "AD Mouse Models", "AD-BXD mice", "ADNI", "AIBL", "APP Mice",
  "Banner", "BLSA", "Cardiovascular Health Study", "Diversity outbred mice",
  "Emory", "Framingham Heart Study", "FreshMicro", "GSE197305", "GSE33000",
  "GSE44772", "GSE48350", "GSE5281", "HBI_scRNAseq", "HBTRC",
  "Health and Retirement Study", "https://doi.org/10.1002/ajmg.b.30607",
  "https://doi.org/10.1016/j.celrep.2020.107908", "https://doi.org/10.1016/j.nbd.2019.104540",
  "https://doi.org/10.1016/j.neulet.2013.06.061", "https://doi.org/10.1038/s41588-018-0311-9",
  "https://doi.org/10.1038/s41588-019-0358-2", "https://doi.org/10.1038/tp.2015.55",
  "https://doi.org/10.3233/jad-2008-15105", "Kronos",
  "Late Onset Alzheimer's Disease Family Study", "Mayo", "MC-BrAD", "MC-CAA",
  "MC_snRNA", "MSBB", "ROSMAP", "Rush", "syn20933797", "syn52525880",
  "TAUAPPms", "VirusResilience"
)
targets_data_filter_values <- c("Behavior", "Clinical", "Genetics", "Metabolomics", "Pathway Modeling", "Phenomics", "Protein", "Risk Scores", "RNA")
targets_first_nom_filter_values <- c(2018:2025)
targets_team_filter_values <- c("ASU", "Chang Lab", "Columbia-Rush", "Duke", "Duke BARU", "Emory", "Emory-Sage-SGC", "Harvard-MIT", "IUSM-Purdue", "JAX-VUMC-UW Resilience", "Longo Lab", "Mayo", "Mayo-UFL-ISB", "MSSM - Roussos Lab", "MSSM - Zhang Lab")
targets_pharos_filter_values <- c('Tbio', 'Tchem', 'Tclin', 'Tdark')
targets_program_filter_values <- c("AMP-AD", "Community", "Resilience-AD", "TREAT-AD")
targets_total_nom_filter_values <- c(1:5)

target_filter_values <- list(
  targets_data_cohort_filter_values,
  targets_data_filter_values,
  targets_first_nom_filter_values,
  targets_team_filter_values,
  targets_total_nom_filter_values,
  targets_pharos_filter_values,
  targets_program_filter_values
)

# Build Nominated Targets ui_config
# ---------------------------------
build_targets_object <- function() {
  columns <- build_object(target_columns)
  filters <- build_filters(target_filters, target_filter_values)

  # Assemble the ui_config object
  list(
    page = "Nominated Targets",
    dropdowns = list(),
    row_count = NA,
    columns = columns,
    filters = filters
  )
}


# ***************
# Nominated Drugs
# ***************
# name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden
#
drug_columns <- data.frame(
  name = c("Drug", "Chembl ID", "Nominations",
           "Combined with", "First Nominated", "Principal Investigators",
           "Programs", "Modality", "First Approved",
           "Max Clinical Trial Phase"),
  type = c("primary", "text", "numeric",
           "text", "numeric", "text",
           "text", "text", "numeric",
           "text"),
  data_key = c("common_name", "chembl_id", "total_nominations",
               "combined_with", "initial_nomination", "principal_investigators",
               "programs", "modality", "year_of_first_approval",
               "maximum_clinical_trial_phase"),
  tooltip = rep(NA, times = 10),
  sort_tooltip =  rep(NA, times = 10),
  link_url =  rep(NA, times = 10),
  link_text =  rep(NA, times = 10),
  is_exported = rep(TRUE, times = 10),
  is_hidden = rep(FALSE, times = 10),
  column_width = rep(NA, times = 10),
  stringsAsFactors = FALSE
)

# AG-2042: Hide programs column, at least until it has >1 value
drug_columns[drug_columns$name == "Programs", "is_hidden"] <- TRUE
drug_columns[drug_columns$name == "Chembl ID", "is_hidden"] <- TRUE

# Nominated Drug Filters
# ----------------------
drug_filters <- data.frame(
  name = c('Max Clinical Trial Phase', 'Modality', 'Nominating PI', 'Nominations'),
  data_key = c('maximum_clinical_trial_phase', 'modality', 'principal_investigators', 'total_nominations'),
  short_name = c('Phase', 'Modality', 'PI', 'Nominations'),
  query_param_key = c('phase', 'modalities', 'nominatingPis', 'nominations'),
  stringsAsFactors = FALSE
)

# Nominated Drug Filter Values
# ----------------------------
drug_modality_filter_values <- c(
  "Small molecule", "Protein"
)

# These are in alpha order by last name
drug_pi_filter_values <- c(
 "Mark Albers", "Feixiong Cheng", "Yadong Huang",
 "Jan Krumsiek", "Marina Sirota", "Lei Xie"
)
drug_total_nom_filter_values <- c(1, 2)
drug_max_clinical_trial_phase_filter_values <- c("Preclinical", "Phase I", "Phase II", "Phase III", "Phase IV", "Unknown")

drug_filter_values <- list(
  drug_max_clinical_trial_phase_filter_values,
  drug_modality_filter_values,
  drug_pi_filter_values,
  drug_total_nom_filter_values
)

# Build Nominated Drugs ui_config
# ---------------------------------
build_drugs_object <- function() {
  columns <- build_object(drug_columns)
  filters <- build_filters(drug_filters, drug_filter_values)

  # Assemble the ui_config object
  list(
    page = "Nominated Drugs",
    dropdowns = list(),
    row_count = NA,
    columns = columns,
    filters = filters
  )
}

```

# Generate and upload ui_config

```{r}
# *********************************************
# Generate ui_config.json and upload to Synapse
# *********************************************
json_list <- c(
  list(build_targets_object()),
  list(build_drugs_object())
)

json_output <- toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
#cat(json_output)

# write json file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

write(json_output, file.path(output_dir, "ui_config.json"))

# upload to synapse
synLogin()
syn_file <- File(
              file.path(output_dir, "ui_config.json"),
              description = "ui_config.json for Agora.",
              parent = "syn7525089"
            )

res <- synStore(
          syn_file,
          forceVersion = FALSE,
          executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/agora/notebooks/preprocessing/AG-2005_ui_config.rmd"
        )
# print synapse_id and new version number
print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))
```


-----
TESTS
-----

## Test build_column
```{r}
col_def <- build_column("name_value", "type_value", "data_key_value", "tooltip_value", "sort_tooltip_value", "link_url_value", "link_text_value", TRUE, FALSE)

print(col_def)

json_output <- toJSON(col_def, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)

```

## Test build_object
```{r}
test_input <- data.frame(
  name = c(NA, "text_name", "link_external_name", "link_internal_name"),
  type = c("primary", "text", "link_external", "link_internal" ),
  data_key = c("key1", "key2", "key3", "key4"),
  tooltip = c(NA, "tooltip2", NA, "tooltip3"),
  sort_tooltip = c(NA, "sort_tooltip1", NA, "sort_tooltip2"),
  link_url = c(NA, NA, NA, "https://test/url/here"),
  link_text = c(NA, NA, "Results", "View Strain"),
  is_exported = c(TRUE, TRUE, TRUE, FALSE),
  is_hidden = c(TRUE, FALSE, TRUE, FALSE),
  stringsAsFactors = FALSE
)

coldefs <- build_object(test_input)

json_list <- c(
  coldefs
)

json_output <- toJSON(json_list, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)

```

# tests
```{r}
json_list <-  list(build_targets_object())

toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
```