Create ui_info.json source file (MG-190)

This notebook generates a collection of JSON objects that represent The Model AD Explorer's Comparison Tool table configurations.

Comparison Tool pages and dropdown menus that are currently supported by this notebook:
* Model Overview
* Gene Expression
* Disease Correlation

ADDING NEW PAGES
1. Update the `pages` vector to include the title of the new CT page
2. Define a vector of menu options for each dropdown menu on the new page
3. Define a vector of values for each static column on the new page
4. Define a vector of values shared by the dynamic columns on the new page (typically all but `name` and possibly `tooltip`); see Disease Correlation for an example that adds column-type-specific tooltips to the dynamic columns
5. Define a vector of `name` values for each dynamic column on the new page; see Gene Expression for an example that uses a multiple vectors of names to generate unique sets of dynamic columns based on dropdown selection
6. Define a vector of filterset display names and filterset display values
7. Create a `build_<page>_objects` function that generates the required columns and filters data
8. Update the `Combine and Output` section to invoke the new build function


## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
- tidyverse
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}
if (!require("jsonlite", quietly = TRUE)) {
  install.packages("jsonlite")
}
if (!require("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}


library(synapser)
library(jsonlite)
library(tidyverse)
```

## Synapse helper functions
```{r}
# MODEL-AD 2.0 harmonized source file directory
# ADP Backend > Files > 2.ConsortiumStudies > MODEL-AD Data Explorer > Model AD Explorer 2.0 Source Files
synapse_folder <- 'syn51498054'

# Vector to store all downloaded synIds for Synapse provenance (populated via download_file function)
data_sources <- character()

# Download a file from Synapse by synId and read it into a dataframe
download_file <- function(synId, df_name, type) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("./", "input")
  id_parts <- strsplit(synId, '.', fixed = TRUE)
  id_value <- id_parts[[1]][1]
  version_value <- id_parts[[1]][2]

  if (is.na(version_value)) {
    version_value <- NULL
  }

  cat(paste0("Downloading ", synId, "..."))

  file <- synGet(id_value, version = version_value, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId, "\n- Modified on ", file$properties$modifiedOn, "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  if (type == 'json') { data <- fromJSON(path) }
  if (type == 'csv'  { data <- read.csv(path) }

  as.data.frame(data)
}

```


## Download source data from Synapse
Note that each file's synId and version is added to the data_sources vector via the download_file function.
The populated data_sources vector is used to populate Synapse provenance when the generated file is uploaded.
```{r}
# TODO pin source file versions prior to finalizing MVP data release

# model_names, model_types, model_centers
model_info <- download_file('syn61378590', type = "csv")

# modified_genes
model_allele_info <- download_file('syn64618791', type = "csv")
human_transgene_allele_map <- download_file('syn64846805', type = "csv")

# biodomains
biodomains_source <- download_file('syn66351272', type = 'csv')

model_allele_info
human_transgene_allele_map
biodomains_source
```

## Extract Modified Gene filter values
```{r}
# Include all human genes in the transgene map
human_genes <- unique(human_transgene_allele_map$gene_symbol)

# Include mouse genes only when not overridden by the transgene map
join <- model_allele_info %>%
  left_join(human_transgene_allele_map, by = join_by('mgi_allele_id')) %>% # not pretty, but it works here
  select(model, gene, gene_symbol)

mouse_genes <- unique(join[is.na(join$gene_symbol),]$gene)
```

## Create model_overview.json
```{r}
# ----------------
# Shared functions
# ----------------

# Generates a CT column definition with base fields
# column_name: the name of the CT column
# column_info: a vector of the column's type, tooltip, and sort_tooltip values
make_column <- function(name, metadata, override_tooltip = NULL) {
  names(metadata) <- c("data_type", "tooltip", "sort_tooltip")
  tooltip_value <- if (!is.null(override_tooltip)) override_tooltip else metadata["tooltip"]

  list(
    name = name,
    type = metadata["data_type"],
    tooltip = tooltip_value,
    sort_tooltip = metadata["sort_tooltip"]
  )
}

# Generates a CT filterset definition
# filterset_name: The display name of the filterset
# filter_on_field: the CT data field the filterset targets
# filter_values: The filter values
make_filter <- function(filterset_name, filter_on_field, filter_values) {
    list(
      name = filterset_name,
      field = filter_on_field,
      values = filter_values
  )
}

# --------------------
# Shared filter values
# --------------------

# Filter values, ordered alphabetically
model_names <- sort(model_info$model)
model_types <- sort(unique(model_info$model_type))
model_centers <- sort(unique(model_info$contributing_group))
modified_genes <- sort(c(mouse_genes, human_genes))
biodomains <- sort(unique(biodomains_source$Biodomain))
sexes <-  c("Female", "Male")

# ------------------------------
# Gene Expression
# ------------------------------
# Page dropdowns
ge_menu1 <- c("RNA - DIFFERENTIAL EXPRESSION")
ge_menu2 <- c("Tissue - Hemibrain", "Tissue - Cerebral Cortex", "Tissue - Hippocampus")
ge_menu3 <- c("Sex - Females & Males", "Sex - Females", "Sex - Males")

# Static and dynamic column info
# - static columns in every view of the page
# - dynamic columns vary depending on dropdown menu selection
ge_static_row <- c("Model", "text", "Mouse Model", "Sort by Model value")
ge_dynamic_row <- c("heat_map", "", "Sort by log2 fold change value")
ge_dynamic_names_jax <- c("4 months", "6 months", "8 months", "12 months", "24 months")
ge_dynamic_names_uci <- c("2 months","4 months", "8 months", "12 months", "18 months", "22 months")

# Filterset display names and target data fields
ge_filter_definitions <- data.frame(
  name = c('Biological Domain', 'Model Type', 'Mouse Model'),
  field = c('biodomains', 'model_type', 'name'),
  stringsAsFactors = FALSE
)

# Filterset filter value lists
ge_filter_values <- list(
  biodomains,
  model_types,
  model_names
)

# Builds Gene Expression objects based on the specified page, column, and filter info
build_ge_objects <- function() {

  # Generate filterdefs
  filters <- Map(function(name, field, values) {
    list(
      name = name,
      field = field,
      values = values
    )
  }, ge_filter_definitions$name, ge_filter_definitions$field, ge_filter_values)

  # Generate every possible combination of dropdown options
  combos <- expand.grid(menu1 = ge_menu1, menu2 = ge_menu2, menu3 = ge_menu3, stringsAsFactors = FALSE)

  # Generate a JSON object for each combo
  lapply(seq_len(nrow(combos)), function(i) {
    row <- combos[i, ]

    # Use the Tissue value to select the appropriate list of dynamic column names
    dynamic_names <- if (row$menu2 == "Tissue - Hemibrain") ge_dynamic_names_jax else ge_dynamic_names_uci

    # Generate coldefs for static & dynamic columns
    dynamic_columns <- lapply(dynamic_names, function(name) make_column(name, ge_dynamic_row))
    static_column <- make_column(ge_static_row[1], ge_static_row[-1])
    columns <- c(list(static_column), dynamic_columns)

    # Remove names to ensure JSON is an array, not object
    names(filters) <- NULL

    # Populate the object with page, dropdowns, coldefs, and filterdefs
    list(
      page = "Gene Expression",
      dropdowns = c(row$menu1, row$menu2, row$menu3),
      columns = columns,
      filters = filters
    )
  })
}

# ------------------------------
# Disease Correlation
# ------------------------------
# Page dropdowns
dc_menu1 <- c("CONSENSUS NETWORK MODULES")
dc_menu2 <- c(
  "Consensus Cluster A - ECM Organization",
  "Consensus Cluster B - Immune System",
  "Consensus Cluster C - Neuronal System",
  "Consensus Cluster D - Cell Cycle, NMD",
  "Consensus Cluster E - Organelle Biogenesis, Cellular Stress Response"
)

# Static and dynamic column info
# - static columns in every view of the page
# - dynamic columns vary depending on dropdown menu selection
dc_static_row1 <- c("Age", "text", "", "Sort by Age value")
dc_static_row2 <- c("Sex", "text", "", "Sort by Sex value")
dc_dynamic_row <- c("heat_map", "", "Sort by correlation value")

dc_dynamic_names_A <- c("IFG", "PHG", "TCX")
dc_dynamic_names_B <- c("CBE", "DLPFC", "PHG", "TCX", "FP", "IFG", "STG")
dc_dynamic_names_C <- c("CBE", "DLPFC", "FP", "IFG", "PHG", "STG", "TCX")
dc_dynamic_names_D <- c("CBE", "DLPFC", "FP", "IFG", "PHG", "STG", "TCX")
dc_dynamic_names_E <- c("CBE", "DLPFC", "PHG", "STG", "TCX", "FP")

dc_ages <- c('4 months', '8 months', '12 months')

# Brain region expansion values for tooltips
dc_tooltip_map <- c(
  CBE = "Cerebellum",
  DLPFC = "Dorsolateral Prefrontal Cortex",
  FP = "Frontal Pole",
  IFG = "Inferior Frontal Gyrus",
  PHG = "Parahippocampal Gyrus",
  STG = "Superior Frontal Gyrus",
  TCX = "Temporal Cortex"
)

# Returns a list of dynamic colnames
# * cluster_label: The Consensus Cluster dropdown option to get dynamic colnames for
get_dc_dynamic_names <- function(cluster_label) {
  if (grepl("Cluster A", cluster_label)) return(dc_dynamic_names_A)
  if (grepl("Cluster B", cluster_label)) return(dc_dynamic_names_B)
  if (grepl("Cluster C", cluster_label)) return(dc_dynamic_names_C)
  if (grepl("Cluster D", cluster_label)) return(dc_dynamic_names_D)
  if (grepl("Cluster E", cluster_label)) return(dc_dynamic_names_E)
  character(0)
}

# Filterset display names and target data fields
dc_filter_definitions <- data.frame(
  name = c('Age', 'Model Type', 'Modified Gene', 'Mouse Model', 'Sex'),
  field = c( 'age', 'model_type', 'modified_genes', 'name', 'sex'),
  stringsAsFactors = FALSE
)

# Filterset filter value lists
dc_filter_values <- list(
  dc_ages,
  model_types,
  modified_genes,
  model_names,
  sexes
)

# Builds Disease Correlation objects based on the specified page, column, and filter info
build_dc_objects <- function() {

  # Generate filterdefs
  filters <- Map(function(name, field, values) {
    list(
      name = name,
      field = field,
      values = values
    )
  }, dc_filter_definitions$name, dc_filter_definitions$field, dc_filter_values)

  # Remove names to ensure JSON is an array, not an object
  names(filters) <- NULL

  # Generate objects for every possible combination of dropdown options
  combos <- expand.grid(menu1 = dc_menu1, menu2 = dc_menu2, stringsAsFactors = FALSE)

  # Generate a JSON object for each combo
  lapply(seq_len(nrow(combos)), function(i) {

    # Generate coldefs for static & dynamic columns
    row <- combos[i, ]
    dynamic_names <- get_dc_dynamic_names(row$menu2)
    dynamic_columns <- lapply(dynamic_names, function(name) {
      tooltip_text <- dc_tooltip_map[[name]]
      make_column(name, dc_dynamic_row, override_tooltip = tooltip_text)
    })

    static1 <- make_column(dc_static_row1[1], dc_static_row1[-1])
    static2 <- make_column(dc_static_row2[1], dc_static_row2[-1])
    columns <- c(list(static1, static2), dynamic_columns)

    # Populate the object with page, dropdowns, coldefs, and filterdefs
    list(
      page = "Disease Correlation",
      dropdowns = c(row$menu1, row$menu2),
      columns = columns,
      filters = filters
    )
  })
}

# ------------------------------
# Model Overview (single static object, no menus)
# ------------------------------
# Static column info
mo_static_definitions <- data.frame(
  name = c(
    "Model Type", "Matched Control", "Gene Expression", "Disease Correlation",
    "Pathology", "Biomarkers", "Study Data", "JAX Strain", "Center"
  ),
  type = c(
    "text", "text",
    "link_internal", "link_internal", "link_internal", "link_internal",
    "link_external", "link_external", "link_external"
  ),
  stringsAsFactors = FALSE
)

# Filterset display names and target data fields
mo_filter_definitions <- data.frame(
  name = c(
    'Available Data', 'Contributing Center', 'Model Type', 'Modified Gene'
    ),
  field = c(
    'available_data', 'center', 'model_type', 'modified_genes'
  ),
  stringsAsFactors = FALSE
)

# Filterset filter value lists
mo_result_types <- c('Biomarkers', 'Disease Correlation', 'Gene Expression', 'Pathology')

mo_filter_values <- list(
  mo_result_types,
  model_centers,
  model_types,
  modified_genes
)

# Builds the Model Overview object based on the specified page, column, and filter info
build_mo_object <- function() {

  # Generate filterdefs
  filters <- Map(function(name, field, values) {
    list(
      name = name,
      field = field,
      values = values
    )
  }, mo_filter_definitions$name, mo_filter_definitions$field, mo_filter_values)

  # Remove names to ensure JSON is an array, not an object
  names(filters) <- NULL

   # Add coldefs
  columns <- apply(mo_static_definitions, 1, function(row) {
    name <- row["name"]
   type <- row["type"]

    metadata <- c(type, "", paste("Sort by", name, "value"))
    column <- make_column(name, metadata)

    # For link type columns, handle link_text and link_url fields
    # - Define values here if they apply to every CT row in the column
    # - Define values that vary by CT row in the data payload
    if (column$type %in% c("link_internal", "link_external")) {
      if (name %in% c("Gene Expression", "Disease Correlation", "Pathology", "Biomarkers")) {
        column$link_text <- "Results"
        # Omit row-specific link_url
      } else if (name == "Study Data") {
        column$link_text <- "View Data"
        # Omit row-specific link_url
      } else if (name == "JAX Strain") {
        column$link_text <- "View Strain"
        # Omit row-specific link_url
      } else if (name == "Center") {
        # Omit row-specific link_text
        column$link_url <- "https://www.model-ad.org/"
      }
    }
    column
  })

  # Populate the object with page, dropdowns, coldefs, and filterdefs
  list(
    page = "Model Overview",
    dropdowns = list(),
    columns = columns,
    filters = filters
  )
}

# ------------------------------
# Combine and Output
# ------------------------------
json_list <- c(
  build_ge_objects(),
  build_dc_objects(),
  list(build_mo_object())  # wrap in list to match structure
)

json_output <- toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
# cat(json_output)

# Optionally write to file for debugging
write(json_output, "./output/ui_config.json")

```

## Write json file and upload to synapse
```{r}
# write csv file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}
write(json_output, file.path(output_dir, "ui_config.json"))

# upload to synapse
syn_file <- File(file.path(output_dir, "ui_config.json"),
                 parent = synapse_folder)

res <- synStore(syn_file, forceVersion = FALSE,
                used = data_sources,
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-190_Create_ui_info_source_file.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```
