
# Modify Disease Correlation source (MG-264)

This notebook downloads the correlation data that was exported from the legacy explorer (syn65467849.1),
 aligns the model_name values with our other data, then writes the updated file back to the same synID.


## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- tidyverse
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}

if (!require("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

library(synapser)
library(tidyverse)
```

## Utility functions & variables
Run this to initialize shared variables and functions.
```{r}
# MODEL-AD 2.0 harmonized source file directory
# ADP Backend > Files > 2.ConsortiumStudies > MODEL-AD Data Explorer > Model AD Explorer 2.0 Source Files
synapse_folder <- 'syn51498054'

# Vector to store all downloaded synIds for Synapse provenance (populated via download_file function)
data_sources <- character()

# Download a file from Synapse by synId and read it into a dataframe
download_file <- function(synId, df_name, type) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("./", "input")
  id_parts <- strsplit(synId, '.', fixed = TRUE)
  id_value <- id_parts[[1]][1]
  version_value <- id_parts[[1]][2]

  if (is.na(version_value)) {
    version_value <- NULL
  }

  cat(paste0("Downloading ", synId, "..."))


  file <- synGet(id_value, version = version_value, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId, "\n- Modified on ", file$properties$modifiedOn, "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  if (type == 'f') {
    data <- read_feather(path)
  } else {
    if (type == 'json') { data <- fromJSON(path) }
    if (type == 'csv' || type == 'table') { data <- read.csv(path) }
    if (type == 'tsv') { data <- read.csv(path, sep = "\t") }
    if (type == 'rda') { data <- load(path)}
    if (type == 'rds') {data <- readRDS(path)}
    df_name <- as.data.frame(data)
  }
}
```


## Download data file from Synapse
Note that each file's synId and version is added to the data_sources vector via the download_file function.
The populated data_sources vector is used to populate Synapse provenance when the harmonized csv is uploaded.
```{r}
disease_correlation_source <- download_file('syn65467849.1', type = "csv")
```
## Update Cluster values (MG-445)
Update the Cluster labels to the new display format, and fix 'Biogenesis' typo from legacy explorer data
```{r}
disease_correlation_final <- disease_correlation_source %>% mutate(Cluster = case_when(
  Cluster == "Consensus Cluster A (ECM organization)" ~ "ECM Organization - Consensus Cluster A",
  Cluster == "Consensus Cluster B (Immune system)" ~ "Immune System - Consensus Cluster B",
  Cluster == "Consensus Cluster C (Neuronal system)" ~ "Neuronal System - Consensus Cluster C",
  Cluster == "Consensus Cluster D (Cell Cycle, NMD)" ~ "Cell Cycle, NMD Consensus Cluster D",
  Cluster == "Consensus Cluster E (Organelle Biogensis, Cellular stress response)" ~ "Organelle Biogenesis, Cellular Stress Response - Consensus Cluster E"
))
```

## Update 5xFAD model_name to include center info
Since only IU/Jax/Pitt contributed disease_correlation data, we can assign all 5xFAD results to that center.

```{r}
disease_correlation_final$Mouse.Model[disease_correlation_final$Mouse.Model == '5xFAD'] <- '5xFAD (IU/Jax/Pitt)'
```


## Write csv file and upload to synapse
```{r}
# write csv file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}
write.csv(disease_correlation_final, file.path(output_dir, "disease_correlation_results.csv"), row.names = FALSE)

# upload to synapse
syn_file <- File(file.path(output_dir, "disease_correlation_results.csv"),
                 parent = synapse_folder)

res <- synStore(syn_file, forceVersion = FALSE,
                used = data_sources,
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-261_Modify_disease_correlation_source_file.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```