# Create Pathology source file (MG-40)

This notebook generates a single harmonized source file (syn61357279) containing the results data and display values required to surface Pathology evidence in the Model-AD Explorer.

Models that are currently handled by this notebook:
* 3xTG-AD
* 5xFAD
* Trem2-R47H_NSS
* Abca7*V1599M

NOTE: synapser's dependency chain loads plyr, which masks dplyr's summarise(),
mutate(), and rename(). All calls to those functions use dplyr:: explicitly
to prevent silent use of plyr's implementations.

ADDING NEW MODELS
As data becomes available for additional models, the data for those models will need to be incorporated into the harmonized source file.

To incorporate each new model into this notebook:

Prerequisites:
* Genotype display name mappings for the new model have been added to rnaseq_genotype_label_map.csv (syn69014401)

1. 'Download data and metadata files from Synapse' section:
- Add a section to download the new model's metadata and data files

2. 'Generate model-specific intermediate files' section
- Add a code block that generates an intermediate results file for the new model
- You may need to modify the transform_results function to include new unit and type display value mappings

3. 'Sanity check nrow in generated files' section
- Add validation for the new model's intermediate file

4. 'Join source files to metadata to populate age, tissue, and genotype values' section
- Update to join metadata to the new model's intermediate file

5. 'Merge files, write csv file, and upload to synapse' section
- Add the new model's intermediate file to the bind_rows call on the first line


ADDING NEW MEASURES
As additional measurement types become available, those measurement types will need to be incorporated into the harmonized source file.

To incorporate a new measurement type into the notebook:

1. 'Utility functions & variables' section
- Update stain_keeps to ensure that the required rows aren't filtered out
- Add a display name for the new measure
- Add a display value for the new measure's units, if not already defined

2. 'Download data and metadata files from Synapse' section
- Add the data file(s) containing the new measurement to the appropriate model-specific section(s)
- Adjust the metadata file version(s) as required to bring in metadata for the new measurement

3. 'Generate model-specific intermediate files' section
- For each model, add the new stain-display name mapping
--include units in the mappings when multiple measurements are being captured from a single source row

4. 'Sanity check nrow in generated files' section
- Add a sanity check for the new measure for each model that has it


## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- tidyverse
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}

if (!require("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

library(synapser)
library(tidyverse)
```

## Utility functions & variables
Run this to initialize shared variables and functions.
```{r}
# *******
# FILTERS
# *******

# Measures (stain values) that we want to include
stain_keeps <- c("GFAP", "S100B", "LAMP1", "IBA1", "ThioS", "HT7", "AT8")

# ages that we want to display results for
age_keeps <- c(4, 12, 18)

# column specs for intermediate per-model files
# For result files without individualID column: 3xTG-AD, NSS, Abca7
keeps_with_pool_join_field <- c('specimenID', 'model', 'type', 'measurement', 'units')
# For result files with individualID column: 5xFAD
keeps_with_join_fields <- c('individualID', 'specimenID', 'model', 'type', 'measurement', 'units')

# column spec for final harmonized output file
keeps <- c('individual_id', 'model', 'type', 'measurement', 'units', 'age_death', 'tissue', 'sex', 'genotype')

# **************
# DISPLAY LABELS
# **************

# Model display names (copied from rnaseq_genotype_label_map.csv) -- there isn't
# a good way to get these values dynamically.
threex_display_name <- '3xTg-AD'
fivex_display_name <- '5xFAD (UCI)'
abca7_display_name <- 'Abca7*V1599M'
nss_display_name <- 'Trem2-R47H_NSS'

# Pathology display names
IBA1_display_name <- 'Microglia Cell Density (IBA1)'
GFAP_display_name <- 'Astrocyte Cell Density (GFAP)'
S100B_display_name <- 'Astrocyte Cell Density (S100B)'
LAMP1_display_name <- 'Dystrophic Neurites (LAMP1)'
AT8_display_name <- 'Phospho-tau (AT8)'
HT7_display_name <- 'Tau (HT7)'
ThioS_density_display_name <- 'Plaque Density (Thio-S)'
ThioS_size_display_name <- 'Plaque Size (Thio-S)'

# Pathology display units
objectDensity_display_name <- '# objects / square mm'
averageObjectVolume_display_name <- 'mean object volume (cubic mm)'
totalObjectVolume_display_name <- 'total object volume (cubic mm)'
meanObjectArea_display_name <- 'mean object area (square mm)'
areaPercent_display_name <- '%'

# *********
# FUNCTIONS
# *********

# Vector to store all downloaded synIds for Synapse provenance (populated via download_csv function)
data_sources <- character()

# Download a csv file from Synapse by synId and read it into a dataframe
# - synId: synapse ID of the file to download
# - df_name: the name to use for the resulting dataframe
download_csv <- function(synId, df_name) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("./", "input")
  id_parts <- strsplit(synId, '.', fixed = TRUE)
  id_value <- id_parts[[1]][1]
  version_value <- id_parts[[1]][2]

  if (is.na(version_value)) {
    version_value <- NULL
  }

  cat(paste0("Downloading ", synId, "..."))

  file <- synGet(id_value, version = version_value, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId, "\n- Modified on ", file$properties$modifiedOn, "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  df_name <- read_csv(path, show_col_types = FALSE, na = c("", "NA", "N/A"), name_repair = ~make.names(., unique = TRUE)) %>%
    select(where(~!all(is.na(.))))
}


# Transforms source data into the desired format, converts type and unit values to standardized display labels,
# and drops unneeded columns
#  - results: A dataframe containing result rows
#  - measurement_columns: A vector containing the colname(s) of the results column(s) that contain the targeted measure(s)
#  - model_name: the model display name; canonical display names are defined in model_info.csv [syn61378590]
#  - fields_to_keep: the appropriate 'column specs for intermediate per-model files' from above
transform_results <- function(results, measurement_columns, model_name, fields_to_keep) {
  results %>% pivot_longer(
                cols = all_of(measurement_columns),
                names_to = "units",
                values_to = "measurement",
                values_transform = list(measurement = as.numeric)

        # Populate model column
        ) %>% add_column(
                model = model_name

        # Convert various unit colnames to consistent unit display values
        ) %>% dplyr::mutate(

                units = case_match(
                        units,
                        c('objectDensity', 'X.objects..sqmm') ~ objectDensity_display_name,
                        c('mean.object.area..sqmm.', 'mean.object.area..squm.') ~ meanObjectArea_display_name,
                        'averageObjectVolume' ~ averageObjectVolume_display_name,
                        'totalObjectVolume' ~ totalObjectVolume_display_name,
                        'Area..' ~ areaPercent_display_name
                )

        # Convert stain values to type display names; tolower() to avoids case mismatches due to variation across source files
        ) %>% dplyr::mutate(
                type = case_when(
                        tolower(stain) == 'iba1' & units == objectDensity_display_name ~ IBA1_display_name,
                        tolower(stain) == 'gfap'  & units == objectDensity_display_name ~  GFAP_display_name,
                        tolower(stain) == 's100b' & units == objectDensity_display_name ~ S100B_display_name,
                        tolower(stain) == 'lamp1' ~ LAMP1_display_name,
                        tolower(stain) == 'at8' ~ AT8_display_name,
                        tolower(stain) == 'ht7' ~ HT7_display_name,
                        tolower(stain) == 'thios' & units == meanObjectArea_display_name ~ ThioS_size_display_name, # 5xFAD
                        tolower(stain) == 'thios' & units == averageObjectVolume_display_name ~ ThioS_size_display_name, # 3xTg-AD
                        tolower(stain) == 'thios' & units == objectDensity_display_name ~ ThioS_density_display_name,
                        .default = 'UNMAPPED_DROPME'
                )

        # drop any unmapped rows
        ) %>% filter(
                type != 'UNMAPPED_DROPME'

        # Drop unused columns
        ) %>% select(
                all_of(fields_to_keep)
        )
}


# Join a data file to a pair of metadata files to get required age, sex, ageDeath, and genotype values
# - df: A dataframe containing result rows
# - individual_metadata: A dataframe containing individual-level metadata
# - biospecimen_metadata: A dataframe containing specimen metadata
# This function prints out the number of rows before and after filtering results by age, which is useful for validation.
# This function will fail if it identifies metadata inconsistencies for pooled samples.
join_to_metadata <- function(results, individual_metadata, biospecimen_metadata) {

  # Ensure that all individualID columns are chr type
  if ("individualID" %in% colnames(results)) {
    results <- results %>%
      dplyr::mutate(individualID = as.character(individualID))
  }

  individual_metadata <- individual_metadata %>%
    dplyr::select(individualID, sex, ageDeath, genotype) %>%
    dplyr::mutate(individualID = as.character(individualID))

  biospecimen_metadata <- biospecimen_metadata %>%
    dplyr::select(individualID, specimenID, tissue) %>%
    dplyr::mutate(individualID = as.character(individualID))

  # Join metadata files by individualID
  metadata <- left_join(biospecimen_metadata, individual_metadata, by = 'individualID')

  # Figure out which join columns are in the results dataframe
  join_by <- if(all(c('specimenID', 'individualID') %in% colnames(results)) &&
              all(c('specimenID', 'individualID') %in% colnames(metadata))) {
    c('specimenID', 'individualID')
  } else {
    c('specimenID')
  }

  # *** DATA INTEGRITY CHECK ***
  # Check if any specimenID (pool) maps to multiple sexes, genotypes, or ages
  joined_temp <- results %>%
    left_join(metadata, by = join_by, relationship = "many-to-many")

  mismatches <- joined_temp %>%
    group_by(specimenID) %>%
    dplyr::summarise(
      n_sex = n_distinct(sex),
      n_genotype = n_distinct(genotype),
      n_age = n_distinct(ageDeath),
      .groups = 'drop'
    ) %>%
    filter(n_sex > 1 | n_genotype > 1 | n_age > 1)

  if (nrow(mismatches) > 0) {
    # Print the offending specimens
    cat("FATAL: Metadata inconsistency found in pooled specimens:")
    print(mismatches)
    stop("Script halted. Pooled specimens must have uniform sex, genotype, and ageDeath.")
  }
  # *** END DATA INTEGRITY CHECK ***

  # Join results to metadata; prevent creating duplicate data rows for 'Pool-X' specimenIDs by matching only first metadata row
  results %>% left_join(metadata, by=join_by, multiple="first"
            ) %>% left_join(genotype_label_map, by = c('model', 'genotype')

            # If specimenID starts with "Pool", use specimenID. Otherwise, keep individualID.
            ) %>% dplyr::mutate(individualID = if_else(
                grepl("^Pool", specimenID),
                as.character(specimenID),
                as.character(individualID)
            )

            # Correct genotypes to align with target display values
            ) %>% dplyr::mutate(genotype = display_label

            # Print out nrow of results before age filtering
            ) %>% { print(nrow(.)) ; .

            # Filter out results for unwanted ages
            } %>% filter(ageDeath %in% age_keeps

            # Print out nrow of results after age filtering
            ) %>% { print(nrow(.)) ; .

            # Convert all individualID columns to string type
            } %>% dplyr::mutate(individualID = as.character(individualID)

            # Rename columns
            ) %>% dplyr::rename(individual_id = individualID,
                         age_death = ageDeath

            # Drop unneeded columns
            ) %>% select(all_of(keeps))
}
```


## Download data and metadata files from Synapse
Note that each file's synId and version is added to the data_sources vector via the download_csv function.
The populated data_sources vector is used to populate Synapse provenance when the harmonized csv is uploaded.
```{r}
# Genotype -> Display label map (rnaseq_genotype_label_map.csv)
genotype_label_map <- download_csv('syn69014401.17') %>%
  # Model name fix, plus change "5xFAD (UCI)" -> "5xFAD"
  dplyr::mutate(model = model_group,
         display_label = str_replace(display_label, " \\(UCI\\)", ""))

# 3xTG-AD
# metadata
threex_biospecimen <- download_csv('syn23532198.3')
threex_individual <- download_csv('syn23532199.4') %>%
  # genotype fix
  dplyr::mutate(genotype = str_replace(genotype, "_homozygous", "_carrier"),
         genotype = str_replace(genotype, "3XTg", "3xTg"))
# data
threex_pathology_source <- download_csv('syn26601504.1')

# 5xFAD
# metadata
fivex_biospecimen <- download_csv('syn18876530.10')
fivex_individual <- download_csv('syn18880070.14') %>%
  # genotype fix
  dplyr::mutate(genotype = str_replace(genotype, "_hemizygous", "_carrier"))
# data
fivex_gfap_s100b_source <- download_csv('syn22049816.3')
fivex_lamp1_source <- download_csv('syn22049817.2')
fivex_iba1_thios_source <- download_csv('syn22049825.4')

#NSS
# metadata
nss_biospecimen <- download_csv('syn29568452.3')
nss_individual <- download_csv('syn27147690.3') %>%
  # genotype fix
  dplyr::mutate(genotype = str_replace(genotype, ",", ";"))
# data
nss_pathology_source <- download_csv('syn42131930.1')

#Abca7
# metadata
abca7_biospecimen <- download_csv('syn30859390.3')
abca7_individual <- download_csv('syn30859394.2') %>%
  # genotype fix
  dplyr::mutate(genotype = str_replace(genotype, ",", ";"))
# data
abca7_pathology_source <- download_csv('syn53127289.1')
```


## Generate model-specific files & join to metadata
File-measurement-unit and display value mappings for legacy models are from the v1 data spreadsheet here:
https://docs.google.com/spreadsheets/d/1B3ZMReC7nR2CGgpFNJuxJTj__lF2h6dt/edit?gid=1108411940#gid=1108411940

### 3xTG-AD
```{r}
threex_final <- threex_pathology_source %>%

        # Filter out rows with unused 'Stain' values
        filter(stain %in% stain_keeps

        # Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
        ) %>% transform_results(c('objectDensity', 'averageObjectVolume', 'totalObjectVolume'), threex_display_name, keeps_with_pool_join_field

        # Drop rows with NA measurements
        ) %>% filter(!is.na(measurement)

        # Join results to metadata
        ) %>% join_to_metadata(threex_individual, threex_biospecimen)
```


```{r}
### 5xFAD
fivex_source_keeps <- c("IndividualID", "SpecimenID", "Stain")

# 5xFAD colnames aren't consistent with other models
fivex_rename_cols <- function (results) {
  results %>% dplyr::rename(c('stain' = 'Stain',
                       'individualID' = 'IndividualID',
                       'specimenID' = 'SpecimenID'))
}

# Filter each source file to select desired rows by 'Stain' value, drop unneeded columns, and rename the remaining columns
fivex_gfap_s100b <- fivex_gfap_s100b_source[fivex_gfap_s100b_source$Stain %in% c('GFAP', 'S100B'),] %>%
  select(all_of(c(fivex_source_keeps, "X.objects..sqmm")))  %>%
  fivex_rename_cols()

fivex_lamp1 <- fivex_lamp1_source[fivex_lamp1_source$Stain == 'LAMP1',] %>%
  select(all_of(c(fivex_source_keeps, "Area.."))) %>%
  fivex_rename_cols()

fivex_iba1 <- fivex_iba1_thios_source[fivex_iba1_thios_source$Stain == 'Iba1',] %>%
  select(all_of(c(fivex_source_keeps, "X.objects..sqmm"))) %>%
  fivex_rename_cols()

fivex_thios <- fivex_iba1_thios_source[fivex_iba1_thios_source$Stain %in% 'ThioS',] %>%
  select(all_of(c(fivex_source_keeps, "X.objects..sqmm", "mean.object.area..squm."))) %>%
  fivex_rename_cols()

# Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
fivex_gfap_s100b_pivot <- transform_results(fivex_gfap_s100b, c('X.objects..sqmm'), fivex_display_name, keeps_with_join_fields)
fivex_lamp1_pivot <- transform_results(fivex_lamp1, c('Area..'), fivex_display_name, keeps_with_join_fields)
fivex_iba1_pivot <- transform_results(fivex_iba1, c('X.objects..sqmm'), fivex_display_name, keeps_with_join_fields)
fivex_thios_pivot <- transform_results(fivex_thios, c('X.objects..sqmm', 'mean.object.area..squm.'), fivex_display_name, keeps_with_join_fields)

# Merge all results into a single data frame
fivex_final <- bind_rows(fivex_gfap_s100b_pivot, fivex_lamp1_pivot, fivex_iba1_pivot, fivex_thios_pivot

    # MG-12: drop rows with "N/A" specimenIDs
    ) %>% filter(
          !is.na(specimenID)

    # MG-268: Replace N/A measurement values with 0s
    ) %>% dplyr::mutate(
            measurement = ifelse(is.na(measurement), 0, measurement)

    # Join results to metadata
    ) %>% join_to_metadata(fivex_individual, fivex_biospecimen)
```

### Trem2-R47H_NSS
```{r}
nss_final <- nss_pathology_source %>%

        # Filter out rows with unused 'Stain' values
        filter(stain %in% stain_keeps

        # Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
        ) %>% transform_results(c('objectDensity', 'averageObjectVolume', 'totalObjectVolume'), nss_display_name, keeps_with_pool_join_field

        # Drop rows with NA measurements
        ) %>% filter(!is.na(measurement)

        # Join results to metadata
        ) %>% join_to_metadata(nss_individual, nss_biospecimen)
```

### Abca7*V1599M
```{r}
abca7_final <- abca7_pathology_source %>%

        # Filter out rows with unused 'Stain' values
        filter(stain %in% stain_keeps

        # Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
        ) %>% transform_results(c('objectDensity', 'averageObjectVolume', 'totalObjectVolume'), abca7_display_name, keeps_with_pool_join_field

        # Drop rows with NA measurements
        ) %>% filter(!is.na(measurement)

        # Drop LAMP1 rows with measurements other than total object volume (cubic mm)
        ) %>% filter(!(type == LAMP1_display_name & (units == objectDensity_display_name | units == averageObjectVolume_display_name))

        # Join results to metadata
        ) %>% join_to_metadata(abca7_individual, abca7_biospecimen)
```
*************


## Sanity check generated files
Checks that ensure that the processed data includes the expected number of results per model, per result type:
1. Source data vs processed data: nrow(filtered_source_data) == nrow(processed_data)
2. Source data vs expected rows: nrow(nrow(processed_data)) == nrow(legacy_explorer_data)
   - Expected row values are calculated in the 'Immunohisto expected row counts' notebook
3. There should be no NA values: all values with NA measurements should be dropped, and there shouldn't be any NA values introduced from the `left_join()` to metadata.


```{r}
# *******
# 3xTG-AD
# *******
# IBA1
threex_iba1 <- threex_pathology_source[threex_pathology_source$stain %in% 'IBA1',]
stopifnot(nrow(threex_final[threex_final$type == IBA1_display_name,]) == nrow(threex_iba1))
stopifnot(nrow(threex_iba1) == 144) # IBA1 :  3xTg-AD + B6129  ==  144

# GFAP
threex_gfap <- threex_pathology_source[threex_pathology_source$stain %in% 'GFAP',]
stopifnot(nrow(threex_final[threex_final$type == GFAP_display_name,]) == nrow(threex_gfap))
stopifnot(nrow(threex_gfap) == 144) #  GFAP :  3xTg-AD + B6129  ==  144

# S100B
threex_s100b <- threex_pathology_source[threex_pathology_source$stain %in% 'S100B',]
stopifnot(nrow(threex_final[threex_final$type == S100B_display_name,]) == nrow(threex_s100b))
stopifnot(nrow(threex_s100b) == 144) #  S100B :  3xTg-AD + B6129  ==  144

# LAMP1
threex_lamp1 <- threex_pathology_source[threex_pathology_source$stain %in% 'LAMP1',]
stopifnot(nrow(threex_final[threex_final$type == LAMP1_display_name,]) == nrow(threex_lamp1))
stopifnot(nrow(threex_lamp1) == 142) # LAMP1 :  3xTg-AD + B6129  ==  142

# AT8
threex_at8 <- threex_pathology_source[threex_pathology_source$stain %in% 'AT8',]
stopifnot(nrow(threex_final[threex_final$type == AT8_display_name,]) == nrow(threex_at8))
stopifnot(nrow(threex_at8) == 72) # AT8 :  3xTg-AD + B6129  ==  72

# HT7
threex_ht7 <- threex_pathology_source[threex_pathology_source$stain %in% 'HT7',]
stopifnot(nrow(threex_final[threex_final$type == HT7_display_name,]) == nrow(threex_ht7))
stopifnot(nrow(threex_ht7) == 72) # HT7 :  3xTg-AD + B6129  ==  72

# ThioS size & density
threex_thioss <- threex_pathology_source[threex_pathology_source$stain %in% 'ThioS',]
stopifnot(nrow(threex_final[threex_final$type == ThioS_size_display_name,]) == nrow(threex_thioss))
stopifnot(nrow(threex_final[threex_final$type == ThioS_density_display_name,]) == nrow(threex_thioss))
stopifnot(nrow(threex_thioss) == 144) #  ThioS size :  3xTg-AD + B6129  ==  144

# Check all fields were populated properly -- there should be no NAs
stopifnot(!any(is.na(threex_final)))

# *****
# 5xFAD
# *****
# IBA1
# Correct for 56 rows removed via age filtering
fivex_ibas <- nrow(fivex_iba1_thios_source[fivex_iba1_thios_source$Stain %in% 'Iba1',]) - 56
stopifnot(nrow(fivex_final[fivex_final$type == IBA1_display_name,]) == fivex_ibas)
stopifnot(nrow(fivex_ibas) == 280) # IBA1 :  5xFAD + C57BL/6J ==  280 rows expected

# GFAP
# Correct for 54 rows removed via age filtering
fivex_gfap <- nrow(fivex_gfap_s100b_source[fivex_gfap_s100b_source$Stain %in% 'GFAP',]) - 54
stopifnot(nrow(fivex_final[fivex_final$type == GFAP_display_name,]) == fivex_gfap)
stopifnot(nrow(fivex_gfap) == 215) # GFAP :  5xFAD + C57BL/6J  ==  215 rows expected

# S100B
# Correct for 54 rows removed via age filtering
fivex_s100bs <- nrow(fivex_gfap_s100b_source[fivex_gfap_s100b_source$Stain %in% 'S100B',]) - 54
stopifnot(nrow(fivex_final[fivex_final$type == S100B_display_name,]) == fivex_s100bs)
stopifnot(nrow(fivex_s100bs) == 215) #  S100B :  5xFAD + C57BL/6J  ==  215  rows expected

# LAMP1
# Correct for 56 rows removed via age filtering
fivex_lamp1 <- nrow(fivex_lamp1_source[fivex_lamp1_source$Stain %in% 'LAMP1',]) - 56
stopifnot(nrow(fivex_final[fivex_final$type == LAMP1_display_name,]) == fivex_lamp1)
stopifnot(nrow(fivex_lamp1) == 264) # LAMP1 :  5xFAD + C57BL/6J  ==  264  rows expected

# ThioS size & density
fivex_thios <- fivex_iba1_thios_source[fivex_iba1_thios_source$Stain %in% 'ThioS',]
fivex_thios <-  filter(fivex_thios, !is.na(SpecimenID)) # Remove rows with NA specimenID
# Correct for 56 rows removed via age filtering
fivex_thios <- nrow(fivex_thios) - 56
stopifnot(nrow(fivex_final[fivex_final$type == ThioS_size_display_name,]) ==fivex_thios)
stopifnot(nrow(fivex_final[fivex_final$type == ThioS_density_display_name,]) == fivex_thios)
stopifnot(nrow(fivex_thios) == 278) #  ThioS (size == density):  5xFAD + C57BL/6J  ==  278 rows expected

# Check all fields were populated properly -- there should be no NAs
stopifnot(!any(is.na(fivex_final)))

# **************
# Trem2-R47H_NSS
# **************
# IBA1
nss_iba1 <- nss_pathology_source[nss_pathology_source$stain %in% 'IBA1',]
stopifnot(nrow(nss_final[nss_final$type == IBA1_display_name,]) == nrow(nss_iba1))
# results not in legacy explorer

# GFAP
nss_gfap <- nss_pathology_source[nss_pathology_source$stain %in% 'GFAP',]
stopifnot(nrow(nss_final[nss_final$type == GFAP_display_name,]) == nrow(nss_gfap))
# results not in legacy explorer

# S100B
nss_s100b <- nss_pathology_source[nss_pathology_source$stain %in% 'S100B',]
stopifnot(nrow(nss_final[nss_final$type == S100B_display_name,]) == nrow(nss_s100b))
# results not in legacy explorer

# LAMP1
nss_lamp1 <- nss_pathology_source[nss_pathology_source$stain %in% 'LAMP1',]
stopifnot(nrow(nss_final[nss_final$type == LAMP1_display_name,]) == nrow(nss_lamp1))
# results not in legacy explorer

# ThioS size & density
nss_thios <- nss_pathology_source[nss_pathology_source$stain %in% 'ThioS',]
# Remove source rows with NA measurements
nss_thioss <-  filter(nss_thios, !is.na(nss_thios$averageObjectVolume))
nss_thiosd <-  filter(nss_thios, !is.na(nss_thios$objectDensity))
stopifnot(nrow(nss_final[nss_final$type == ThioS_size_display_name,]) == nrow(nss_thioss))
stopifnot(nrow(nss_final[nss_final$type == ThioS_density_display_name,]) == nrow(nss_thiosd))
# results not in legacy explorer

# Check all fields were populated properly -- there should be no NAs
stopifnot(!any(is.na(nss_final)))


# ************
# Abca7*V1599M
# ************
# IBA1
abca7_iba1 <- abca7_pathology_source[abca7_pathology_source$stain %in% 'IBA1',]
stopifnot(nrow(abca7_final[abca7_final$type == IBA1_display_name,]) == nrow(abca7_iba1))
# results not in legacy explorer

# GFAP
abca7_gfap <- abca7_pathology_source[abca7_pathology_source$stain %in% 'GFAP',]
abca7_gfap <-  filter(abca7_gfap, !is.na(abca7_gfap$objectDensity)) # Remove source rows with NA measurements
stopifnot(nrow(abca7_final[abca7_final$type == GFAP_display_name,]) == nrow(abca7_gfap))
# results not in legacy explorer

# S100B
abca7_s100b <- abca7_pathology_source[abca7_pathology_source$stain %in% 'S100B',]
stopifnot(nrow(abca7_final[abca7_final$type == S100B_display_name,]) == nrow(abca7_s100b))
# results not in legacy explorer

# LAMP1
abca7_lamp1 <- abca7_pathology_source[abca7_pathology_source$stain %in% 'LAMP1',]
stopifnot(nrow(abca7_final[abca7_final$type == LAMP1_display_name,]) == nrow(abca7_lamp1))
# results not in legacy explorer

# ThioS size & density
abca7_thios <- abca7_pathology_source[abca7_pathology_source$stain %in% 'ThioS',]
stopifnot(nrow(abca7_final[abca7_final$type == ThioS_size_display_name,]) == nrow(abca7_thios))
stopifnot(nrow(abca7_final[abca7_final$type == ThioS_density_display_name,]) == nrow(abca7_thios))
# results not in legacy explorer

# Check all fields were populated properly -- there should be no NAs
stopifnot(!any(is.na(abca7_final)))
```


## Merge model-specific results files & write csv file to synapse
```{r}
pathology_final <- bind_rows(
  threex_final,
  fivex_final,
  nss_final,
  abca7_final)

# MODEL-AD 2.0 harmonized source file directory
# ADP Backend > Files > 2.ConsortiumStudies > MODEL-AD Data Explorer > Model AD Explorer 2.0 Source Files
synapse_folder <- 'syn51498054'

# write csv file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}
write.csv(pathology_final, file.path(output_dir, "pathology.csv"), row.names = FALSE)

# BREAK HERE FOR LOCAL VERIFICATION

# upload to synapse
syn_file <- File(file.path(output_dir, "pathology.csv"),
                 description = "Harmonized Pathology data for Model-AD Explorer 2.0.",
                 parent = synapse_folder)

res <- synStore(syn_file, forceVersion = FALSE,
                used = data_sources,
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-40_Create_harmonized_pathology_source_file.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```



# compare generated file to latest version in synapse
```{r}
file_synID <- "syn61357279"
input_dir <- file.path("./", "input")

old_file <- download_csv(file_synID)
  path <- old_file$path

# verify changes are expected
identical(old_file, pathology_final)
all.equal(old_file, pathology_final)
```
