---
title: "MG-402: Create ui_config source file"
---
This notebook generates the ui_config.json source file (syn66527739) containing the data required to construct Comparison Tool interfaces for the Model-AD Explorer MVP.

Interfaces that are currently handled by this notebook:
* Differential Expression
* Disease Correlation
* Model Overview

This notebook must be updated when the Explorer is updated to:
* Add a new CT interface
* Add or modify a column of an existing CT interface
* Add or modify a CT dropdown menu or dropdown value
* Add or modify a filterset or filter value


Notes:
* `query_param_key`: Values must be plural, camelCase, and should intelligibly align with the related display values presented to users on the page:
   - filter.name (filter panels)
   - filter.short_name (filter chiclets)
   - column.name (column header)


## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}
if (!require("jsonlite", quietly = TRUE)) { install.packages("jsonlite") }
if (!require("purrr")) { install.packages("purrr") }

library(synapser)
library(purrr)
library(jsonlite)
```


## Variables and functions
Sources the required variables and functions.
```{r}
# *******************************
# STATIC ROW COUNTS (MG-384)
# *******************************
# we don't need row_count values for ModelAD CT loading indicators since we've introduced server-side pagination
gene_expression_row_count = NA
disease_correlation_row_count = NA
model_overview_row_count = NA

# *************
# FILTER VALUES
# *************
biodomain_filter_vals <- c('Apoptosis',
                           'APP Metabolism',
                           'Autophagy',
                           'Cell Cycle',
                           'DNA Repair',
                           'Endolysosome',
                           'Epigenetic',
                           'Immune Response',
                           'Lipid Metabolism',
                           'Metal Binding and Homeostasis',
                           'Mitochondrial Metabolism',
                           'Myelination',
                           'Oxidative Stress',
                           'Proteostasis',
                           'RNA Spliceosome',
                           'Structural Stabilization',
                           'Synapse',
                           'Tau Homeostasis',
                           'Vasculature')
model_type_filter_vals <- c('Familial AD',
                            'Late Onset AD')
sex_filter_vals <- c('Female', 'Male')


# *********
# FUNCTIONS
# *********

# Builds a ui_config object from a dataframe containing column info
# Currently used only by Model Overview.
# - page_column_data: a csv-compatible dataframe containing a row of values per CT column:
#    - name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden
build_object <- function(page_column_data) {

  columns <- purrr::pmap(page_column_data, build_column)
  names(columns) <- NULL
  columns
}

# Builds a column object
# - name: The column display name
#   -- Specify NA for hidden columns; if you specify a different value, it will be ignored
# - type: The column type
# - tooltip: The tooltip display value for the column name
#   -- Specify NA for column names that need no explanatory tooltip, and for hidden columns
# - sort_tooltip: The tooltip display value for the column sort indicator
#   -- This key is always omitted for columns with the primary type, and for hidden columns
#   -- Specify NA to use the default sort_tooltip value for any other column: "Sort by <name> value"
# - link_url: The optional default target URL for columns with a link_* type
#   -- Specify NA for columns with other types
# - link_text: The optional default link display text for columns with a link_* type
#   -- Specify NA for columns with other types
# - is_exported: Whether this column should be included in CSV data exports
# - is_hidden: Whether this column should be displayed in the CT table
build_column <- function(name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden) {

  # Use the default sort_tooltip if one isn't specified, for non-hidden columns only
  if (is_hidden == TRUE) { sort_tooltip = NA }
  else { sort_tooltip = if (!is.na(sort_tooltip)) sort_tooltip else paste("Sort by", name, "value") }

  column <- list(
    name = name,
    type = type,
    data_key = data_key,
    tooltip = tooltip,
    sort_tooltip = sort_tooltip,
    link_url = link_url ,
    link_text = link_text,
    is_exported = is_exported,
    is_hidden = is_hidden
  )

  # Omit keys with NA values
  clean_column <- purrr::discard(column, ~ is.null(.x) | all(is.na(.x)))
  clean_column
}

# Builds a list of filter definitions
# - filter_defs: A dataframe of filterset info:
#   - name, data_field, short_name, query_param_key, filter_values
# - filter_vals: A list of lists specifying the filter values for each filterset
build_filters <- function(filter_defs, filter_values) {

  # Construct list of filters
  filters <- Map(function(name, data_key, short_name, query_param_key, filter_values) {

    filter <- list(
      name = name,
      data_key = data_key,
      short_name = short_name,
      query_param_key = query_param_key,
      values = filter_values
      )

      # Omit keys with NA values
      filter <- discard(filter, ~ is.null(.x) | all(is.na(.x)))

    }, filter_defs$name, filter_defs$data_key, filter_defs$short_name, filter_defs$query_param_key, filter_values)

  # Remove names to ensure JSON is an array, not object
  names(filters) <- NULL
  filters
}


# ***************
# Differential Expression
# ***************

# Differential Expression Dropdown Menus
# ------------------------------
ge_menu1 <- c("RNA - DIFFERENTIAL EXPRESSION")
ge_menu2 <- c("Tissue - Hemibrain", "Tissue - Cerebral Cortex", "Tissue - Hippocampus")

# Differential Expression Columns
# -------------------------
# name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden
#
# Static columns are the same across all views
ge_primary_column <- list("Gene", "primary", "gene_symbol", "Mouse Gene", NA, NA, NA, TRUE, FALSE)
ge_model_column <- list("Model", "link_internal", "name", "Mouse Model", "Sort by Model value", NA, NA, TRUE, FALSE)
ge_sex_column <- list("Sex", "text", "sex", NA, "Sort by Sex value", NA, NA, TRUE, FALSE)
ge_matched_control_column <- list("Control", "text", "matched_control", NA, "Sort by Control value", NA, NA, TRUE, FALSE)

# Hidden static columns
ge_tissue_column <- list(NA, "text", "tissue", NA, NA, NA, NA, TRUE, TRUE)
ge_biodomains_column <- list(NA, "text", "biodomains", NA, NA, NA, NA, TRUE, TRUE)
ge_model_group_column <- list(NA, "text", "model_group", NA, NA, NA, NA, FALSE, TRUE)
ge_model_type_column <- list(NA, "text", "model_type", NA, NA, NA, NA, TRUE, TRUE)
ge_ensembl_gene_id_column <- list(NA, "text", "ensembl_gene_id", NA, NA, NA, NA, TRUE, TRUE)


# Variable heatmap columns have the same values across all views for these fields
ge_dynamic_column_type <- "heat_map"
ge_dynamic_column_sort_tooltip <- "Sort by log2 fold change value"
ge_dynamic_column_is_exported <- TRUE
ge_dynamic_column_is_hidden <- FALSE

# Variable heatmap column names vary by contributing center and by tissue, since each center sequences different tissues
ge_dynamic_names_jax <- c("4 months", "12 months")
ge_dynamic_names_uci <- c("4 months", "12 months", "18 months")

# Differential Expression Filters
# -----------------------
ge_filters <- data.frame(
  name = c('Biological Domain', 'Model Type', 'Mouse Model', 'Sex'),
  data_key = c('biodomains', 'model_type', 'name', 'sex'),
  short_name = c('Biodomain', 'Type', 'Model', 'Sex'),
  query_param_key = c('biodomains', 'modelTypes', 'models', 'sex'),
  stringsAsFactors = FALSE
)

# Differential Expression Filter values
# -----------------------------
# MG-750: Only include model filter values for models with results for the selected tissue

# Hemibrain: All JAX models
ge_model_name_filter_vals_hemibrain <- c(
                            '5xFAD (IU/Jax/Pitt)',
                            'APOE4',
                            'LOAD1',
                            'Trem2R47H')

# Cortex: 5xFAD (UCI); need to wrap single value in a list for JSON serialization
ge_model_name_filter_vals_cerebral_cortex <- list('5xFAD (UCI)')

# Hippocampus: All UCI models
ge_model_name_filter_vals_hippocampus <- c(
                            '3xTg-AD',
                            '5xFAD (UCI)',
                            'Abca7*V1599M',
                            'Abca7*V1599M.5xFAD',
                            'Trem2-R47H_NSS',
                            'Trem2-R47H_NSS.5xFAD')

# Generates Differential Expression ui_config objects
build_ge_objects <- function() {

  # Static columns are always the same
  primary <- do.call(build_column, ge_primary_column)
  ensembl_gene_id  <- do.call(build_column, ge_ensembl_gene_id_column)
  tissue <- do.call(build_column, ge_tissue_column)
  model <- do.call(build_column, ge_model_column)
  sex <- do.call(build_column, ge_sex_column)
  matched_control <- do.call(build_column, ge_matched_control_column)
  biodomains <- do.call(build_column, ge_biodomains_column)
  model_group <- do.call(build_column, ge_model_group_column)
  model_type <- do.call(build_column, ge_model_type_column)


  # Build a ui_config for each possible combination of dropdown menu options
  dropdown_combos <- expand.grid(menu1 = ge_menu1, menu2 = ge_menu2, stringsAsFactors = FALSE)
  lapply(seq_len(nrow(dropdown_combos)), function(i) {
    combo <- dropdown_combos[i, ]

    # Dynamic columns vary by tissue dropdown
    dynamic_names <- if (grepl("Hemibrain", combo$menu2, ignore.case = TRUE)) ge_dynamic_names_jax else ge_dynamic_names_uci
    dynamic_columns <- lapply(dynamic_names,function(name) {
                           build_column(name, ge_dynamic_column_type, name, NA, ge_dynamic_column_sort_tooltip,
                                        NA, NA, ge_dynamic_column_is_exported, ge_dynamic_column_is_hidden)
    })

    # Dynamic model filter values vary by tissue dropdown
    dynamic_model_filters <- if (grepl("Hemibrain", combo$menu2, ignore.case = TRUE)) {
      ge_model_name_filter_vals_hemibrain
    } else if (grepl("Cerebral Cortex", combo$menu2, ignore.case = TRUE)) {
      ge_model_name_filter_vals_cerebral_cortex
    } else if (grepl("Hippocampus", combo$menu2, ignore.case = TRUE)) {
      ge_model_name_filter_vals_hippocampus
    } else {
      stop() # fail if we are missing required filter values
    }

    # Construct the filter values list for this specific combo
    ge_filter_values <- list(
      biodomain_filter_vals,
      model_type_filter_vals,
      dynamic_model_filters,
      sex_filter_vals
    )

    # The column ordering is mirrored when a CSV is generated
    columns <- c(list(primary, ensembl_gene_id, tissue, model, sex, matched_control), dynamic_columns, c(list(biodomains, model_type, model_group)))
    filters <- build_filters(ge_filters, ge_filter_values)

    list(
      page = "Differential Expression",
      dropdowns = c(combo$menu1, combo$menu2, combo$menu3),
      row_count = gene_expression_row_count,
      columns = columns,
      filters = filters
    )
  })
}


# *******************
# Disease Correlation
# *******************

# Disease Correlation Dropdowns Menus
# -----------------------------------
dc_menu1 <- c("CONSENSUS NETWORK MODULES")
dc_menu2 <- c(
  "Immune System - Consensus Cluster B",
  "Neuronal System - Consensus Cluster C",
  "Organelle Biogenesis, Cellular Stress Response - Consensus Cluster E",
  "Cell Cycle, NMD Consensus Cluster D",
  "ECM Organization - Consensus Cluster A"
)

# Disease Correlation Columns
# ---------------------------
# name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden
#
# All views have the same static column values
dc_primary_column <- list("Model", "primary", "name", "Mouse Model", NA, NA, NA, TRUE, FALSE)
dc_age_column <- list("Age", "text", "age", NA, "Sort by Age value", NA, NA, TRUE, FALSE)
dc_sex_column <- list("Sex", "text", "sex", NA, "Sort by Sex value", NA, NA, TRUE, FALSE)

# Hidden static columns
dc_cluster_column <- list(NA, "text", "cluster", NA, NA, NA, NA, TRUE, TRUE)
dc_model_type_column <- list(NA, "text", "model_type", NA, NA, NA, NA, TRUE, TRUE)
dc_matched_control_column <- list(NA, "text", "matched_control", NA, NA, NA, NA, TRUE, TRUE)
dc_modified_genes_column <- list(NA, "text", "modified_genes", NA, NA, NA, NA, TRUE, TRUE)

# Variable heatmap columns have the same values across all views for these fields
dc_dynamic_column_type <- "heat_map"
dc_dynamic_column_sort_tooltip <- "Sort by correlation value"
dc_dynamic_column_is_exported <- TRUE
dc_dynamic_column_is_hidden <- FALSE

# Dynamic column names vary by consensus cluster option
dc_dynamic_names_A <- c("IFG module", "PHG module", "TCX module")
dc_dynamic_names_B <- c("CBE module", "DLPFC module", "PHG module", "TCX module", "FP module", "IFG module", "STG module")
dc_dynamic_names_C <- c("CBE module", "DLPFC module", "FP module", "IFG module", "PHG module", "STG module", "TCX module")
dc_dynamic_names_D <- c("CBE module", "DLPFC module", "FP module", "IFG module", "PHG module", "STG module", "TCX module")
dc_dynamic_names_E <- c("CBE module", "DLPFC module", "PHG module", "STG module", "TCX module", "FP module")

get_dc_dynamic_names <- function(cluster_label) {
  if (grepl("Cluster A", cluster_label)) return(dc_dynamic_names_A)
  if (grepl("Cluster B", cluster_label)) return(dc_dynamic_names_B)
  if (grepl("Cluster C", cluster_label)) return(dc_dynamic_names_C)
  if (grepl("Cluster D", cluster_label)) return(dc_dynamic_names_D)
  if (grepl("Cluster E", cluster_label)) return(dc_dynamic_names_E)
  character(0)
}

# Brain region acronym expansions for tooltips
dc_tooltip_map <- c(
  "CBE module" = "Human gene module: Cerebellum",
  "DLPFC module" = "Human gene module: Dorsolateral Prefrontal Cortex",
  "FP module" = "Human gene module: Frontal Pole",
  "IFG module" = "Human gene module: Inferior Frontal Gyrus",
  "PHG module" = "Human gene module: Parahippocampal Gyrus",
  "STG module" = "Human gene module: Superior Frontal Gyrus",
  "TCX module" = "Human gene module: Temporal Cortex"
)

# Disease Correlation Filters
# ------------------------------
dc_filters <- data.frame(
  name = c('Age', 'Model Type', 'Modified Gene', 'Mouse Model', 'Sex'),
  data_key = c('age', 'model_type', 'modified_genes', 'name', 'sex'),
  short_name = c(NA, 'Type', 'Gene', 'Model', NA),
  query_param_key = c('ages', 'modelTypes', 'modifiedGenes', 'models', 'sexes'),
  stringsAsFactors = FALSE
)

# Disease Correlation Filter values
# ---------------------------------
# Only include model.name filter values for models with Disease Correlation results
dc_model_name_filter_vals <- c('5xFAD (IU/Jax/Pitt)',
                            'APOE4',
                            'LOAD1',
                            'LOAD1.Abca7A1527G',
                            'LOAD1.Ceacam1KO',
                            'LOAD1.Clasp2L163P',
                            'LOAD1.CR1long',
                            'LOAD1.Il1rapExon2KO',
                            'LOAD1.MthfrC677T',
                            'LOAD1.Snx1D465N',
                            'Trem2R47H')
# Only include modified_gene filter values for genes modified in models with Disease Correlation results
dc_modified_gene_filter_vals <- c('Abca7',
                               'APOE',
                               'APP',
                               'Ceacam1',
                               'Clasp2',
                               'CR1',
                               'CR2',
                               'Il1rap',
                               'Mthfr',
                               "PSEN1",
                               'Snx1',
                               'Trem2')
dc_filter_values <- list(
  c('4 months', '8 months', '12 months'),
  model_type_filter_vals,
  dc_modified_gene_filter_vals,
  dc_model_name_filter_vals,
  sex_filter_vals
)

# Generates Disease Correlation ui_config objects for each possible combination of dropdown menus
build_dc_objects <- function() {

  # Static columns are always the same
  primary <- do.call(build_column, dc_primary_column)
  age <- do.call(build_column, dc_age_column)
  sex <- do.call(build_column, dc_sex_column)
  cluster <- do.call(build_column, dc_cluster_column)
  model_type <- do.call(build_column, dc_model_type_column)
  matched_control <- do.call(build_column, dc_matched_control_column)
  modified_genes <- do.call(build_column, dc_modified_genes_column)

  # Build a ui_config for each possible combination of dropdown menu options
  dropdown_combos <- expand.grid(menu1 = dc_menu1, menu2 = dc_menu2, stringsAsFactors = FALSE)
  lapply(seq_len(nrow(dropdown_combos)), function(i) {
    combo <- dropdown_combos[i, ]

    # Dynamic columns vary by consensus cluster option
    dynamic_names <- get_dc_dynamic_names(combo$menu2)
    dynamic_columns <- lapply(dynamic_names, function(name) {
      tooltip_text <- dc_tooltip_map[[name]]

      # The data_key values for the disease_correlation dataset are exact matches to the name value
      build_column(name, dc_dynamic_column_type, data_key = sub(" module", "", name),
                   tooltip_text, dc_dynamic_column_sort_tooltip, NA,
                   NA, dc_dynamic_column_is_exported, dc_dynamic_column_is_hidden
    )})

    # The column ordering is mirrored when a CSV is generated
    columns <- c(list(primary, cluster, age, sex), dynamic_columns, c(list(model_type, matched_control, modified_genes)))
    filters <- build_filters(dc_filters, dc_filter_values)

    list(
      page = "Disease Correlation",
      dropdowns = c(combo$menu1, combo$menu2),
      row_count = disease_correlation_row_count,
      columns = columns,
      filters = filters
    )
  })
}


# **************
# Model Overview
# **************

# ********************
# Mouse Model Overview
# ********************
# name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden
#
mo_columns <- data.frame(
  name = c("Model", "Model Type", "Matched Controls",
           "Transcriptomics", "Disease Correlation", "Biomarkers",
           "Pathology", "Study Data", "JAX Strain",
           "Center", NA, NA
  ),
  type = c("primary", "text", "text",
    "link_internal", "link_internal", "link_internal",
    "link_internal","link_external", "link_external",
    "text", "text", "text"
  ),
  data_key = c("name", "model_type", "matched_controls",
               "transcriptomics", "disease_correlation", "biomarkers",
               "pathology", "study_data", "jax_strain",
               "center", "modified_genes", "available_data"),
  tooltip = c("Mouse Model",  rep(NA, times = 11)),
  sort_tooltip = rep(NA, times = 12),
  link_url = rep(NA, times = 12),
  link_text = c(NA, NA, NA, "Results", "Results", "Results", "Results", "View Data", "View Strain", NA, NA, NA),
  is_exported = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE),
  is_hidden = c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE),
  stringsAsFactors = FALSE
)

# Model Overview Filters
# ----------------------
mo_filters <- data.frame(
  name = c('Available Data', 'Contributing Center', 'Model Type', 'Modified Gene'),
  data_key = c('available_data', 'center', 'model_type', 'modified_genes'),
  short_name = c('Data', 'Center', 'Type', 'Gene'),
  query_param_key = c('availableData', 'centers', 'modelTypes', 'modifiedGenes'),
  stringsAsFactors = FALSE
)

# Model Overview Filter Values
# ----------------------------
# Include modified_gene filter values for genes modified in any included model
# - Mouse gene symbols use initial caps, e.g. Psen1
# - For human transgenes introduced into mouse models use all caps, e.g. PSEN1
mo_modified_gene_filter_vals <- c('Abca7',
                               'APOE',
                               'APP',
                               'Ceacam1',
                               'Clasp2',
                               'CR1',
                               'CR2',
                               'Il1rap',
                               'MAPT',
                               'Mthfr',
                               'Plcg2',
                               'Psen1',
                               'PSEN1',
                               'Snx1',
                               'Trem2')

mo_filter_values <- list(
  c('Biomarkers', 'Disease Correlation', 'Pathology', 'Transcriptomics'),
  c('IU/Jax/Pitt', 'UCI'),
  model_type_filter_vals,
  mo_modified_gene_filter_vals
)

# Generates Model Overview ui_config objects
build_mo_object <- function() {
  columns <- build_object(mo_columns)
  filters <- build_filters(mo_filters, mo_filter_values)

  # Assemble the ui_config object
  # note that the page value does not actually control the CT's display title
  # so this is not updated to 'Mouse Model Overview' even though the display title will be updated
  list(
    page = "Model Overview",
    dropdowns = list(),
    row_count = model_overview_row_count,
    columns = columns,
    filters = filters
  )
}


# ********************
# Marmo Model Overview
# ********************

marmo_ad_adkp_study_url <- "https://adknowledgeportal.synapse.org/Explore/Studies/DetailsPage/StudyDetails?Study=syn61849889"

# name, type, data_key, tooltip, sort_tooltip, link_url, link_text, is_exported, is_hidden
#
marmo_mo_columns <- data.frame(
  name = c("Model", "Model Type", "Plasma Biomarkers", "Study Data", NA, NA),
  type = c("primary", "text", "link_internal", "link_external", "text", "text"),
  data_key = c("name", "model_type", "biomarkers", "study_data", "modified_genes", "available_data"),
  tooltip = c("Marmoset Model",  rep(NA, times = 5)),
  sort_tooltip = rep(NA, times = 6),
  link_url = c(NA, NA, NA, marmo_ad_adkp_study_url, NA, NA),
  link_text = c(NA, NA, "Results", "View Data", NA, NA),
  is_exported = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE),
  is_hidden = c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE),
  stringsAsFactors = FALSE
)

# Marmo Model Overview Filters
# ----------------------
marmo_mo_filters <- data.frame(
  name = c('Available Data', 'Model Type', 'Modified Gene'),
  data_key = c('available_data', 'model_type', 'modified_genes'),
  short_name = c('Data', 'Type', 'Gene'),
  query_param_key = c('availableData', 'modelTypes', 'modifiedGenes'),
  stringsAsFactors = FALSE
)

# Marmo Model Overview Filter Values
# ----------------------------
# Include modified_gene filter values for genes modified in any included model
# Note: like human gene symbols, marmoset gene symbols use all caps
marmo_mo_modified_gene_filter_vals <- c('PSEN1')

marmo_mo_filter_values <- list(
  c('Plasma Biomarkers'),
  model_type_filter_vals,
  marmo_mo_modified_gene_filter_vals
)

# Generates Model Overview ui_config objects
build_marmo_mo_object <- function() {
  columns <- build_object(marmo_mo_columns)
  filters <- build_filters(marmo_mo_filters, marmo_mo_filter_values)

  # Assemble the ui_config object
  list(
    page = "Marmoset Model Overview",
    dropdowns = list(),
    row_count = model_overview_row_count,
    columns = columns,
    filters = filters
  )
}

```

# Generate and upload ui_config

```{r}
# *********************************************
# Generate ui_config.json and upload to Synapse
# *********************************************
json_list <- c(
  build_ge_objects(),
  build_dc_objects(),
  list(build_mo_object()),
  list(build_marmo_mo_object())
)

json_output <- toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
# cat(json_output)

# write json file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

write(json_output, file.path(output_dir, "ui_config.json"))

# BREAK HERE FOR LOCAL VALIDATION - generated file is in ADT notebooks/output

# upload to synapse
synLogin()
syn_file <- File(
              file.path(output_dir, "ui_config.json"),
              description = "ui_config.json for Model-AD Explorer 2.0.",
              parent = "syn51498054"
            )

res <- synStore(
          syn_file,
          forceVersion = FALSE,
          executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-402_Create ui_config source file.rmd"
        )
# print synapse_id and new version number
print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```


-----
TESTS
-----

## Test build_column
```{r}
col_def <- build_column("name_value", "type_value", "data_key_value", "tooltip_value", "sort_tooltip_value", "link_url_value", "link_text_value", TRUE, FALSE)

print(col_def)

json_output <- toJSON(col_def, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)

```

## Test build_object
```{r}
test_input <- data.frame(
  name = c(NA, "text_name", "link_external_name", "link_internal_name"),
  type = c("primary", "text", "link_external", "link_internal" ),
  data_key = c("key1", "key2", "key3", "key4"),
  tooltip = c(NA, "tooltip2", NA, "tooltip3"),
  sort_tooltip = c(NA, "sort_tooltip1", NA, "sort_tooltip2"),
  link_url = c(NA, NA, NA, "https://test/url/here"),
  link_text = c(NA, NA, "Results", "View Strain"),
  is_exported = c(TRUE, TRUE, TRUE, FALSE),
  is_hidden = c(TRUE, FALSE, TRUE, FALSE),
  stringsAsFactors = FALSE
)

coldefs <- build_object(test_input)

json_list <- c(
  coldefs
)

json_output <- toJSON(json_list, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)

```

# test build_mo_object
```{r}
json_list <-  list(build_mo_object())

toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
```

# test build_dc_objects
```{r}
json_list <-  build_dc_objects()

toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
```


# test build_ge_objects
```{r}
json_list <-  build_ge_objects()

toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
```