---
title: "MG-402: Create ui_config source file"
---
This notebook generates the ui_config.json source file (syn66527739) containing the data required to construct Comparison Tool interfaces for the Model-AD Explorer MVP.

Interfaces that are currently handled by this notebook:
* Gene Expression
* Disease Correlation
* Model Overview

This notebook must be updated when the Explorer is updated to:
* Add a new CT interface
* Add or modify a column of an existing CT interface
* Add or modify a CT dropdown menu or dropdown value
* Add or modify a filterset or filter value

## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}

if (!require("jsonlite", quietly = TRUE)) {
  install.packages("jsonlite")
}

library(synapser)
library(jsonlite)
```



```{r}

# *******************************
# STATIC COLUMN DEFINITION FIELDS
# *******************************
static_colnames <- c("name", "data_type", "tooltip", "sort_tooltip")


# *************
# FILTER VALUES
# *************
biodomain_filter_vals <- c('Apoptosis',
                           'APP Metabolism',
                           'Autophagy',
                           'Cell Cycle',
                           'DNA Repair',
                           'Endolysosome',
                           'Epigenetic',
                           'Immune Response',
                           'Lipid Metabolism',
                           'Metal Binding and Homeostasis',
                           'Mitochondrial Metabolism',
                           'Myelination',
                           'Oxidative Stress',
                           'Proteostasis',
                           'RNA Spliceosome',
                           'Structural Stabilization',
                           'Synapse',
                           'Tau Homeostasis',
                           'Vasculature')
model_type_filter_vals <- c('Familial AD',
                            'Late Onset AD')
model_name_filter_vals <- c('3xTg-AD',
                            '5xFAD (IU/Jax/Pitt)',
                            '5xFAD (UCI)',
                            'Abca7*V1599M',
                            'APOE4',
                            'LOAD1',
                            'LOAD1.Abca7A1527G',
                            'LOAD1.Ceacam1KO',
                            'LOAD1.Clasp2L163P',
                            'LOAD1.CR1long',
                            'LOAD1.Il1rapExon2KO',
                            'LOAD1.MthfrC677T',
                            'LOAD1.Plcg2M28L',
                            'LOAD1.Snx1D465N',
                            'Trem2-R47H_NSS',
                            'Trem2R47H')
modified_gene_filter_vals <- c('Abca7',
                               'APOE',
                               'App',
                               'APP',
                               'Ceacam1',
                               'Clasp2',
                               'CR1',
                               'CR2',
                               'Il1rap',
                               'MAPT',
                               'Mthfr',
                               'Plcg2',
                               'Psen1',
                               'PSEN1',
                               'Snx1',
                               'Trem2')
sex_filter_vals <- c('Female', 'Male')

# ****************
# SHARED FUNCTIONS
# ****************

# Builds a column definition object
# - name: The column display name
# - metadata: An object specifying the column's data_type, tooltip value, and sort_tooltip value
# - override_tooltip: If specified, this value is used in place of metadata.tooltip
make_column <- function(name, metadata, override_tooltip = NULL) {
  names(metadata) <- static_colnames[-1]
  tooltip_value <- if (!is.null(override_tooltip)) override_tooltip else metadata["tooltip"]

  list(
    name = name,
    type = metadata["data_type"],
    tooltip = tooltip_value,
    sort_tooltip = metadata["sort_tooltip"]
  )
}


# Builds a list of filter definitions
# - filter_defs: A dataframe of filterset names and linked data fields
# -filter_vals: A list of lists specifying the filter values for each filterset
make_filters <- function(filter_defs, filter_vals) {

  # Construct list of filters
  filters <- Map(function(name, field, values) {
    list(
      name = name,
      field = field,
      values = values
      )
    }, filter_defs$name, filter_defs$field, filter_vals)

  # Remove names to ensure JSON is an array, not object
  names(filters) <- NULL

  filters
}

# ***************
# Gene Expression
# ***************

# Gene Expression Dropdown Menus
# ------------------------------
ge_menu1 <- c("RNA - DIFFERENTIAL EXPRESSION")
ge_menu2 <- c("Tissue - Hemibrain", "Tissue - Cerebral Cortex", "Tissue - Hippocampus")
ge_menu3 <- c("Sex - Females & Males", "Sex - Females", "Sex - Males")

# Gene Expression Columns
# -------------------------
ge_static_row <- c("Model", "text", "Mouse Model", "Sort by Model value")
ge_dynamic_row <- c("heat_map", "", "Sort by log2 fold change value")
ge_dynamic_names_jax <- c("4 months", "12 months")
ge_dynamic_names_uci <- c("4 months", "12 months", "18 months")


# Gene Expression Filters
# -----------------------
ge_filter_defs <- data.frame(
  name = c('Biological Domain', 'Model Type', 'Mouse Model'),
  field = c('biodomains', 'model_type', 'name'),
  stringsAsFactors = FALSE
)

# Gene Expression Filter values
# -----------------------------
ge_filter_values <- list(
  biodomain_filter_vals,
  model_type_filter_vals,
  model_name_filter_vals
)

# Generates Gene Expression ui_config objects
build_ge_objects <- function() {
  combos <- expand.grid(menu1 = ge_menu1, menu2 = ge_menu2, menu3 = ge_menu3, stringsAsFactors = FALSE)

  lapply(seq_len(nrow(combos)), function(i) {

    # Build column_defs
    row <- combos[i, ]
    static_column <- make_column(ge_static_row[1], ge_static_row[-1])

    # Age columns vary by contributing center; use Tissue value as a proxy, since values are unique per center
    dynamic_names <- if (row$menu2 == "Tissue - Hemibrain") ge_dynamic_names_jax else ge_dynamic_names_uci
    dynamic_columns <- lapply(dynamic_names, function(name) make_column(name, ge_dynamic_row))

    columns <- c(list(static_column), dynamic_columns)

    # Build filter_defs
    filters <- make_filters(ge_filter_defs, ge_filter_values)

    list(
      page = "Gene Expression",
      dropdowns = c(row$menu1, row$menu2, row$menu3),
      columns = columns,
      filters = filters
    )
  })
}


# *******************
# Disease Correlation
# *******************

# Disease Correlation Dropdowns Menus
# -----------------------------------
dc_menu1 <- c("CONSENSUS NETWORK MODULES")
dc_menu2 <- c(
  "Consensus Cluster A - ECM Organization",
  "Consensus Cluster B - Immune System",
  "Consensus Cluster C - Neuronal System",
  "Consensus Cluster D - Cell Cycle, NMD",
  "Consensus Cluster E - Organelle Biogenesis, Cellular Stress Response"
)

# Disease Correlation Columns
# ---------------------------
dc_static_row1 <- c("Age", "text", "", "Sort by Age value")
dc_static_row2 <- c("Sex", "text", "", "Sort by Sex value")
dc_dynamic_row <- c("heat_map", "", "Sort by correlation value")

dc_dynamic_names_A <- c("IFG", "PHG", "TCX")
dc_dynamic_names_B <- c("CBE", "DLPFC", "PHG", "TCX", "FP", "IFG", "STG")
dc_dynamic_names_C <- c("CBE", "DLPFC", "FP", "IFG", "PHG", "STG", "TCX")
dc_dynamic_names_D <- c("CBE", "DLPFC", "FP", "IFG", "PHG", "STG", "TCX")
dc_dynamic_names_E <- c("CBE", "DLPFC", "PHG", "STG", "TCX", "FP")

dc_tooltip_map <- c(
  CBE = "Cerebellum",
  DLPFC = "Dorsolateral Prefrontal Cortex",
  FP = "Frontal Pole",
  IFG = "Inferior Frontal Gyrus",
  PHG = "Parahippocampal Gyrus",
  STG = "Superior Frontal Gyrus",
  TCX = "Temporal Cortex"
)

get_dc_dynamic_names <- function(cluster_label) {
  if (grepl("Cluster A", cluster_label)) return(dc_dynamic_names_A)
  if (grepl("Cluster B", cluster_label)) return(dc_dynamic_names_B)
  if (grepl("Cluster C", cluster_label)) return(dc_dynamic_names_C)
  if (grepl("Cluster D", cluster_label)) return(dc_dynamic_names_D)
  if (grepl("Cluster E", cluster_label)) return(dc_dynamic_names_E)
  character(0)
}

# Disease Correlation Filters
# ------------------------------
dc_filter_defs <- data.frame(
  name = c('Age', 'Model Type', 'Modified Gene', 'Mouse Model', 'Sex'),
  field = c('age', 'model_type', 'modified_genes', 'name', 'sex'),
  stringsAsFactors = FALSE
)

# Disease Correlation Filter values
# ---------------------------------
dc_filter_values <- list(
  c('4 months', '8 months', '12 months'),
  model_type_filter_vals,
  modified_gene_filter_vals,
  model_name_filter_vals,
  sex_filter_vals
)

# Generates Disease Correlation ui_config objects
build_dc_objects <- function() {
  combos <- expand.grid(menu1 = dc_menu1, menu2 = dc_menu2, stringsAsFactors = FALSE)

  lapply(seq_len(nrow(combos)), function(i) {
    row <- combos[i, ]
    dynamic_names <- get_dc_dynamic_names(row$menu2)
    dynamic_columns <- lapply(dynamic_names, function(name) {
      tooltip_text <- dc_tooltip_map[[name]]
      make_column(name, dc_dynamic_row, override_tooltip = tooltip_text)
    })

    static1 <- make_column(dc_static_row1[1], dc_static_row1[-1])
    static2 <- make_column(dc_static_row2[1], dc_static_row2[-1])
    columns <- c(list(static1, static2), dynamic_columns)

    filters <- make_filters(dc_filter_defs, dc_filter_values)

    list(
      page = "Disease Correlation",
      dropdowns = c(row$menu1, row$menu2),
      columns = columns,
      filters = filters
    )
  })
}


# **************
# Model Overview
# **************

# Model Overview Columns
# ----------------------
mo_columns <- data.frame(
  name = c(
    "Model Type", "Matched Control", "Gene Expression", "Disease Correlation",
    "Pathology", "Biomarkers", "Study Data", "JAX Strain", "Center"
  ),
  type = c(
    "text", "text",
    "link_internal", "link_internal", "link_internal", "link_internal",
    "link_external", "link_external", "link_external"
  ),
  stringsAsFactors = FALSE
)

# Model Overview Filters
# ----------------------
mo_filters <- data.frame(
  name = c('Available Data', 'Contributing Center', 'Model Type', 'Modified Gene'),
  field = c('available_data', 'center', 'model_type', 'modified_genes'),
  stringsAsFactors = FALSE
)

# Model Overview Filter Values
# ----------------------------
mo_filter_values <- list(
  c('Biomarkers', 'Disease Correlation', 'Gene Expression', 'Pathology'),
  c('IU/Jax/Pitt', 'UCI'),
  model_type_filter_vals,
  model_name_filter_vals
)

# Generates Model Overview ui_config objects
build_mo_object <- function() {
  columns <- apply(mo_columns, 1, function(row) {
    name <- row["name"]
    type <- row["type"]

    metadata <- c(type, "", paste("Sort by", name, "value"))
    column <- make_column(name, metadata)

    if (column$type %in% c("link_internal", "link_external")) {
      if (name %in% c("Gene Expression", "Disease Correlation", "Pathology", "Biomarkers")) {
        column$link_text <- "Results"
      } else if (name == "Study Data") {
        column$link_text <- "View Data"
      } else if (name == "JAX Strain") {
        column$link_text <- "View Strain"
      } else if (name == "Center") {
        # Omit link_text, add link_url instead
        column$link_url <- "https://www.model-ad.org/"
      }
    }
    column
  })

  # Construct list of filters
  filters <- Map(function(name, field, values) {
    list(
      name = name,
      field = field,
      values = values
      )
    }, mo_filters$name, mo_filters$field, mo_filter_values)

  # Remove names to ensure JSON is an array, not object
  names(filters) <- NULL

  # Assemble the ui_config object
  list(
    page = "Model Overview",
    dropdowns = list(),
    columns = columns,
    filters = filters
  )
}

# *********************************************
# Generate ui_config.json and upload to Synapse
# *********************************************
json_list <- c(
  build_ge_objects(),
  build_dc_objects(),
  list(build_mo_object())
)

json_output <- toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
cat(json_output)

# write json file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

write(json_output, file.path(output_dir, "ui_config.json"))

# upload to synapse
syn_file <- File(file.path(output_dir, "ui_config.json"),
                 description = "ui_config.json for Model-AD Explorer 2.0.",
                 parent = "syn51498054")

res <- synStore(syn_file, forceVersion = FALSE,
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-402_Create ui_config source file.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```