---
title: "MG-402: Create ui_config source file"
---
This notebook generates the ui_config.json source file (syn66527739) containing the data required to construct Comparison Tool interfaces for the Model-AD Explorer MVP.

Interfaces that are currently handled by this notebook:
* Gene Expression
* Disease Correlation
* Model Overview

This notebook must be updated when the Explorer is updated to:
* Add a new CT interface
* Add or modify a column of an existing CT interface
* Add or modify a CT dropdown menu or dropdown value
* Add or modify a filterset or filter value

## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- jsonlite
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}

if (!require("jsonlite", quietly = TRUE)) {
  install.packages("jsonlite")
}

library(synapser)
library(jsonlite)
```


## Variables and functions
Sources the required variables and functions.
```{r}
# *******************************
# STATIC ROW COUNTS (MG-384)
# *******************************
gene_expression_row_count = "over 200000" # THIS IS NOT THE REAL VALUE
disease_correlation_row_count = "1800" # 1800 source rows/total results, 300 CT rows, 60 rows/cluster, variable results per row
model_overview_row_count = "16"

# *************
# FILTER VALUES
# *************
biodomain_filter_vals <- c('Apoptosis',
                           'APP Metabolism',
                           'Autophagy',
                           'Cell Cycle',
                           'DNA Repair',
                           'Endolysosome',
                           'Epigenetic',
                           'Immune Response',
                           'Lipid Metabolism',
                           'Metal Binding and Homeostasis',
                           'Mitochondrial Metabolism',
                           'Myelination',
                           'Oxidative Stress',
                           'Proteostasis',
                           'RNA Spliceosome',
                           'Structural Stabilization',
                           'Synapse',
                           'Tau Homeostasis',
                           'Vasculature')
model_type_filter_vals <- c('Familial AD',
                            'Late Onset AD')
model_name_filter_vals <- c('3xTg-AD',
                            '5xFAD (IU/Jax/Pitt)',
                            '5xFAD (UCI)',
                            'Abca7*V1599M',
                            'APOE4',
                            'LOAD1',
                            'LOAD1.Abca7A1527G',
                            'LOAD1.Ceacam1KO',
                            'LOAD1.Clasp2L163P',
                            'LOAD1.CR1long',
                            'LOAD1.Il1rapExon2KO',
                            'LOAD1.MthfrC677T',
                            'LOAD1.Plcg2M28L',
                            'LOAD1.Snx1D465N',
                            'Trem2-R47H_NSS',
                            'Trem2R47H')
modified_gene_filter_vals <- c('Abca7',
                               'APOE',
                               'App',
                               'APP',
                               'Ceacam1',
                               'Clasp2',
                               'CR1',
                               'CR2',
                               'Il1rap',
                               'MAPT',
                               'Mthfr',
                               'Plcg2',
                               'Psen1',
                               'PSEN1',
                               'Snx1',
                               'Trem2')
sex_filter_vals <- c('Female', 'Male')


# *********
# FUNCTIONS
# *********

# Builds a ui_config object from a dataframe containing column info
# Currently used only by Model Overview.
# - page_column_data: a csv-compatible dataframe containing a row of values per CT column:
#    - name, type, data_key, tooltip, sort_tooltip, link_url, link_text
build_object <- function(page_column_data) {

  columns <- apply(page_column_data, 1, function(row) {

    name <- if (!is.null(row["name"])) row["name"] else ""
    type <- row["type"]
    data_key <- row["data_key"]
    tooltip <- row["tooltip"]
    sort_tooltip <- row["sort_tooltip"]
    link_url = row["link_url"]
    link_text = row["link_text"]

    build_column(name, type, data_key, tooltip, sort_tooltip, link_url, link_text)
  })
  columns
}

# Builds a column object
# - name: The column display name
#   -- Specify NA for columns with the primary type
# - type: The column type
# - tooltip: The tooltip display value for the column name
#   -- Specify NA for column names that need no explanatory tooltip
# - sort_tooltip: The tooltip display value for the column sort indicator
#   -- Specify NA to use the default sort_tooltip value: "Sort by <name> value"
# - link_url: The optional default target URL for columns with a link_* type
#   -- Specify NA for columns with other types
# - link_text: The optional default link display text for columns with a link_* type
#   -- Specify NA for columns with other types
build_column <- function(name, type, data_key, tooltip, sort_tooltip, link_url, link_text) {

  # Use the default sort_tooltip if one isn't specified, for non-primary columns only
  if (type == "primary") { sort_tooltip = NA }
  else { sort_tooltip = if (!is.na(sort_tooltip)) sort_tooltip else paste("Sort by", name, "value") }

  column <- list(
    name = name,
    type = type,
    data_key = data_key,
    tooltip = tooltip,
    sort_tooltip = sort_tooltip,
    link_url = link_url ,
    link_text = link_text
  )

  # Omit keys with NA values
  clean_column <- discard(column, ~ is.null(.x) | all(is.na(.x)))
  clean_column
}

# Builds a list of filter definitions
# - filter_defs: A dataframe of filterset info
#   -- name, data_field, short_name, filter_values
# - filter_vals: A list of lists specifying the filter values for each filterset
build_filters <- function(filter_defs, filter_values) {

  # Construct list of filters
  filters <- Map(function(name, data_key, short_name, filter_values) {

    filter <- list(
      name = name,
      data_key = data_key,
      short_name = short_name,
      values = filter_values
      )

      # Omit keys with NA values
      filter <- discard(filter, ~ is.null(.x) | all(is.na(.x)))

    }, filter_defs$name, filter_defs$data_key, filter_defs$short_name, filter_values)

  # Remove names to ensure JSON is an array, not object
  names(filters) <- NULL
  filters
}


# ***************
# Gene Expression
# ***************

# Gene Expression Dropdown Menus
# ------------------------------
ge_menu1 <- c("RNA - DIFFERENTIAL EXPRESSION")
ge_menu2 <- c("Tissue - Hemibrain", "Tissue - Cerebral Cortex", "Tissue - Hippocampus")
ge_menu3 <- c("Sex - Females & Males", "Sex - Females", "Sex - Males")

# Gene Expression Columns
# -------------------------
# Static columns are the same across all views
ge_primary_column <- list(NA, "primary", "gene_symbol", NA, NA, NA, NA)
ge_model_column <- list("Model", "link_internal", "name", "Mouse Model", "Sort by Model value", NA, NA)
ge_matched_control_column <- list("Control", "text", "matched_control", NA, "Sort by Control value", NA, NA)

# Dynamic columns have the same type and tooltip across all views
ge_dynamic_column_type <- "heat_map"
ge_dynamic_column_sort_tooltip <- "Sort by log2 fold change value"

# Dynamic column names vary by contributing center and by tissue, since each center sequences different tissues
ge_dynamic_names_jax <- c("4 months", "12 months")
ge_dynamic_names_uci <- c("4 months", "12 months", "18 months")


# Gene Expression Filters
# -----------------------
ge_filters <- data.frame(
  name = c('Biological Domain', 'Model Type', 'Mouse Model'),
  data_key = c('biodomains', 'model_type', 'name'),
  short_name = c('Biodomain', 'Type', 'Model'),
  stringsAsFactors = FALSE
)

# Gene Expression Filter values
# -----------------------------
ge_filter_values <- list(
  biodomain_filter_vals,
  model_type_filter_vals,
  model_name_filter_vals
)

# Generates Gene Expression ui_config objects
build_ge_objects <- function() {

  # Static columns are always the same
  primary <- do.call(build_column, ge_primary_column)
  model <- do.call(build_column, ge_model_column)
  matched_control <- do.call(build_column, ge_matched_control_column)

  # Dynamic columns vary by tissue dropdown
  dropdown_combos <- expand.grid(menu1 = ge_menu1, menu2 = ge_menu2, menu3 = ge_menu3, stringsAsFactors = FALSE)
  lapply(seq_len(nrow(dropdown_combos)), function(i) {
    combo <- dropdown_combos[i, ]
    dynamic_names <- if (combo$menu2 == "Tissue - Hemibrain") ge_dynamic_names_jax else ge_dynamic_names_uci
    dynamic_columns <- lapply(dynamic_names,function(name) {
                           build_column(name, ge_dynamic_column_type, name, NA, ge_dynamic_column_sort_tooltip, NA, NA)
    })

    columns <- c(list(primary, model, matched_control), dynamic_columns)
    filters <- build_filters(ge_filters, ge_filter_values)

    list(
      page = "Gene Expression",
      dropdowns = c(combo$menu1, combo$menu2, combo$menu3),
      row_count = gene_expression_row_count,
      columns = columns,
      filters = filters
    )
  })
}


# *******************
# Disease Correlation
# *******************

# Disease Correlation Dropdowns Menus
# -----------------------------------
dc_menu1 <- c("CONSENSUS NETWORK MODULES")
dc_menu2 <- c(
  "Consensus Cluster A - ECM Organization",
  "Consensus Cluster B - Immune System",
  "Consensus Cluster C - Neuronal System",
  "Consensus Cluster D - Cell Cycle, NMD",
  "Consensus Cluster E - Organelle Biogenesis, Cellular Stress Response"
)

# Disease Correlation Columns
# ---------------------------
# All views have the same static column values
dc_primary_column <- list(NA, "primary", "name", NA, NA, NA, NA)
dc_age_column <- list("Age", "text", "age", NA, "Sort by Age value", NA, NA)
dc_sex_column <- list("Sex", "text", "sex", NA, "Sort by Sex value", NA, NA)

# All views have the same type and tooltip for dynamic columns
dc_dynamic_column_type <- "heat_map"
dc_dynamic_column_sort_tooltip <- "Sort by correlation value"

# Dynamic column names vary by consensus cluster option
dc_dynamic_names_A <- c("IFG", "PHG", "TCX")
dc_dynamic_names_B <- c("CBE", "DLPFC", "PHG", "TCX", "FP", "IFG", "STG")
dc_dynamic_names_C <- c("CBE", "DLPFC", "FP", "IFG", "PHG", "STG", "TCX")
dc_dynamic_names_D <- c("CBE", "DLPFC", "FP", "IFG", "PHG", "STG", "TCX")
dc_dynamic_names_E <- c("CBE", "DLPFC", "PHG", "STG", "TCX", "FP")

get_dc_dynamic_names <- function(cluster_label) {
  if (grepl("Cluster A", cluster_label)) return(dc_dynamic_names_A)
  if (grepl("Cluster B", cluster_label)) return(dc_dynamic_names_B)
  if (grepl("Cluster C", cluster_label)) return(dc_dynamic_names_C)
  if (grepl("Cluster D", cluster_label)) return(dc_dynamic_names_D)
  if (grepl("Cluster E", cluster_label)) return(dc_dynamic_names_E)
  character(0)
}

# Brain region acronym expansions for tooltips
dc_tooltip_map <- c(
  CBE = "Human gene module: Cerebellum",
  DLPFC = "Human gene module: Dorsolateral Prefrontal Cortex",
  FP = "Human gene module: Frontal Pole",
  IFG = "Human gene module: Inferior Frontal Gyrus",
  PHG = "Human gene module: Parahippocampal Gyrus",
  STG = "Human gene module: Superior Frontal Gyrus",
  TCX = "Human gene module: Temporal Cortex"
)

# Disease Correlation Filters
# ------------------------------
dc_filters <- data.frame(
  name = c('Age', 'Model Type', 'Modified Gene', 'Mouse Model', 'Sex'),
  data_key = c('age', 'model_type', 'modified_genes', 'name', 'sex'),
  short_name = c(NA, 'Type', 'Gene', 'Model', NA),
  stringsAsFactors = FALSE
)

# Disease Correlation Filter values
# ---------------------------------
dc_filter_values <- list(
  c('4 months', '8 months', '12 months'),
  model_type_filter_vals,
  modified_gene_filter_vals,
  model_name_filter_vals,
  sex_filter_vals
)

# Generates Disease Correlation ui_config objects for each possible combination of dropdown menus
build_dc_objects <- function() {

  # Static columns are always the same
  primary <- do.call(build_column, dc_primary_column)
  age <- do.call(build_column, dc_age_column)
  sex <- do.call(build_column, dc_sex_column)

  # Dynamic columns vary by consensus cluster option
  dropdown_combos <- expand.grid(menu1 = dc_menu1, menu2 = dc_menu2, stringsAsFactors = FALSE)
  lapply(seq_len(nrow(dropdown_combos)), function(i) {
    combo <- dropdown_combos[i, ]
    dynamic_names <- get_dc_dynamic_names(combo$menu2)
    dynamic_columns <- lapply(dynamic_names, function(name) {
      tooltip_text <- dc_tooltip_map[[name]]
      # The data_key values for the disease_correlation dataset are exact matches to the name value
      build_column(name, dc_dynamic_column_type, data_key = name, tooltip_text, dc_dynamic_column_sort_tooltip, NA, NA)
    })


    columns <- c(list(primary, age, sex), dynamic_columns)
    filters <- build_filters(dc_filters, dc_filter_values)

    list(
      page = "Disease Correlation",
      dropdowns = c(combo$menu1, combo$menu2),
      row_count = disease_correlation_row_count,
      columns = columns,
      filters = filters
    )
  })
}


# **************
# Model Overview
# **************
mo_columns <- data.frame(
  name = c(NA, "Model Type", "Matched Controls",
           "Gene Expression", "Disease Correlation", "Pathology",
           "Biomarkers", "Study Data", "JAX Strain", "Center"
  ),
  type = c("primary", "text", "text",
    "link_internal", "link_internal", "link_internal",
    "link_internal","link_external", "link_external", "link_external"
  ),
  data_key = c("name", "model_type", "matched_controls", "gene_expression",
                 "disease_correlation", "pathology", "biomarkers",
                 "study_data", "jax_strain", "center"),
  tooltip = rep(NA, times = 10),
  sort_tooltip = rep(NA, times = 10),
  link_url = c(NA, NA, NA, NA, NA, NA, NA, NA, NA, NA),
  link_text = c(NA, NA, NA, "Results", "Results", "Results", "Results", "View Data", "View Strain", NA),
  stringsAsFactors = FALSE
)

# Model Overview Filters
# ----------------------
mo_filters <- data.frame(
  name = c('Available Data', 'Contributing Center', 'Model Type', 'Modified Gene'),
  data_key = c('available_data', 'center', 'model_type', 'modified_genes'),
  short_name = c('Data', 'Center', 'Type', 'Gene'),
  stringsAsFactors = FALSE
)

# Model Overview Filter Values
# ----------------------------
mo_filter_values <- list(
  c('Biomarkers', 'Disease Correlation', 'Gene Expression', 'Pathology'),
  c('IU/Jax/Pitt', 'UCI'),
  model_type_filter_vals,
  model_name_filter_vals
)

# Generates Model Overview ui_config objects
build_mo_object <- function() {
  columns <- build_object(mo_columns)
  filters <- build_filters(mo_filters, mo_filter_values)

  # Assemble the ui_config object
  list(
    page = "Model Overview",
    dropdowns = list(),
    row_count = model_overview_row_count,
    columns = columns,
    filters = filters
  )
}



```

# Generate and upload ui_config

```{r}
# *********************************************
# Generate ui_config.json and upload to Synapse
# *********************************************
json_list <- c(
  build_ge_objects(),
  build_dc_objects(),
  list(build_mo_object())
)

json_output <- toJSON(json_list, pretty = TRUE, auto_unbox = TRUE)
# cat(json_output)

# write json file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

write(json_output, file.path(output_dir, "ui_config.json"))

# upload to synapse
syn_file <- File(
              file.path(output_dir, "ui_config.json"),
              description = "ui_config.json for Model-AD Explorer 2.0.",
              parent = "syn51498054"
            )

res <- synStore(
          syn_file,
          forceVersion = FALSE,
          executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-402_Create ui_config source file.rmd"
        )
# print synapse_id and new version number
print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```


## Test build_object
```{r}
test_input <- data.frame(
  name = c(NA, "text_name", "link_external_name", "link_internal_name"),
  type = c("primary", "text", "link_external", "link_internal" ),
  data_key = c("key1", "key2", "key3", "key4"),
  tooltip = c(NA, "tooltip2", NA, "tooltip3"),
  sort_tooltip = c(NA, "sort_tooltip1", NA, "sort_tooltip2"),
  link_url = c(NA, NA, NA, "https://test/url/here"),
  link_text = c(NA, NA, "Results", "View Strain"),
  stringsAsFactors = FALSE
)

coldefs <- build_object(test_input)

json_list <- c(
  coldefs
)

json_output <- toJSON(json_list, pretty = TRUE, na = NULL, null = NULL, auto_unbox = TRUE)
cat(json_output)

```