
# Create Pathology source file (MG-40)

This notebook generates a single harmonized source file (syn61357279) containing the results data and display values required to surface Pathology evidence in the Model-AD Explorer.

Models that are currently handled by this notebook:
* 3xTG-AD
* 5xFAD
* Trem2-R47H_NSS
* Abca7*V1599M

ADDING NEW MODELS
As data becomes available for additional models, the data for those models will need to be incorporated into the harmonized source file.

To incorporate each new model into this notebook:

1. 'Download data and metadata files from Synapse' section:
- Add a section to download the new model's metadata and data files

2. 'Generate model-specific intermediate files' section
- Add a code block that generates an intermediate results file for the new model

3. 'Sanity check nrow in generated files' section
- Add validation for the new model's intermediate file

4. 'Join source files to metadata to populate age, tissue, and genotype values' section
- Update to join metadata to the new model's intermediate file

5. 'Merge files, write csv file, and upload to synapse' section
- Add the new model's intermediate file to the bind_rows call on the first line
- Add genotype-genotype display name mappings for the new model


ADDING NEW MEASURES
As additional measurement types become available, those measurement types will need to be incorporated into the harmonized source file.

To incorporate a new measurement type into the notebook:

1. 'Utility functions & variables' section
- Update stain_keeps to ensure that the required rows aren't filtered out
- Add a display name for the new measure
- Add a display value for the new measure's units, if not already defined

2. 'Download data and metadata files from Synapse' section
- Add the data file(s) containing the new measurement to the appropriate model-specific section(s)
- Adjust the metadata file version(s) as required to bring in metadata for the new measurement

3. 'Generate model-specific intermediate files' section
- For each model, add the new stain-display name mapping
--include units in the mappings when multiple measurements are being captured from a single source row

4. 'Sanity check nrow in generated files' section
- Add a sanity check for the new measure for each model that has it


## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- tidyverse
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "http://cran.fhcrc.org"))
}

if (!require("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

library(synapser)
library(tidyverse)
```

## Utility functions & variables
Run this to initialize shared variables and functions.
```{r}
# MODEL-AD 2.0 harmonized source file directory
# ADP Backend > Files > 2.ConsortiumStudies > MODEL-AD Data Explorer > Model AD Explorer 2.0 Source Files
synapse_folder <- 'syn51498054'

# Measures (stain values) that we want to include
stain_keeps <- c("GFAP", "S100B", "LAMP1", "IBA1", "ThioS", "HT7", "AT8")

# Harmonized column specs for intermediate per-model files
keeps_with_pool_join_field <- c('specimenID', 'model', 'type', 'measurement', 'units') # 3xTG-AD, NSS
keeps_with_join_fields <- c('individualID', 'specimenID', 'model', 'type', 'measurement', 'units') # 5xFAD

# Harmonized column spec for final output file
keeps <- c('model', 'type', 'measurement', 'units', 'ageDeath', 'tissue', 'sex', 'genotype')

# Pathology display names
IBA1_display_name <- 'Microglia Cell Density (IBA1)'
GFAP_display_name <- 'Astrocyte Cell Density (GFAP)'
S100B_display_name <- 'Astrocyte Cell Density (S100B)'
LAMP1_display_name <- 'Dystrophic Neurites (LAMP1)'
AT8_display_name <- 'Phospho-tau (AT8)'
HT7_display_name <- 'Tau (HT7)'
ThioS_density_display_name <- 'Plaque Density (Thio-S)'
ThioS_size_display_name <- 'Plaque Size (Thio-S)'


# Pathology display units
objectDensity_display_name <- '# objects / square mm'
averageObjectVolume_display_name <- 'mean object volume (cubic mm)'
totalObjectVolume_display_name <- 'total object volume (cubic mm)'
meanObjectArea_display_name <- 'mean object area (square mm)'
areaPercent_display_name <- '%'


# Vector to store all downloaded synIds for Synapse provenance (populated via download_file function)
data_sources <- character()

# Download a file from Synapse by synId and read it into a dataframe
download_file <- function(synId, df_name, type) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("./", "input")
  id_parts <- strsplit(synId, '.', fixed = TRUE)
  id_value <- id_parts[[1]][1]
  version_value <- id_parts[[1]][2]

  if (is.na(version_value)) {
    version_value <- NULL
  }

  cat(paste0("Downloading ", synId, "..."))


  file <- synGet(id_value, version = version_value, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId, "\n- Modified on ", file$properties$modifiedOn, "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  if (type == 'f') {
    data <- read_feather(path)
  } else {
    if (type == 'json') { data <- fromJSON(path) }
    if (type == 'csv' || type == 'table') { data <- read.csv(path) }
    if (type == 'tsv') { data <- read.csv(path, sep = "\t") }
    if (type == 'rda') { data <- load(path)}
    if (type == 'rds') {data <- readRDS(path)}
    df_name <- as.data.frame(data)
  }
}

# Join a data file to a pair of metadata files to get required age, sex, ageDeath, and genotype values
join_to_metadata <- function(individual_metadata, biospecimen_metadata, df) {

  individual_metadata <- subset(individual_metadata, select=c(individualID, sex, ageDeath, genotype))
  biospecimen_metadata <- subset(biospecimen_metadata, select=c(individualID, specimenID, tissue))

  # Join biospecimen metadata to individual metadata
  merged_metadata <- merge(individual_metadata, biospecimen_metadata, by = 'individualID')

  # Prevent creating duplicate data rows for 'Pool-X' specimenIDs by matching only first metadata row
  df_out <- df %>% left_join(merged_metadata, by='specimenID', multiple="first")

  # make sure we didn't change the number of data rows
  stopifnot(nrow(df) == nrow(df_out))

  df_out
}
```


## Download data and metadata files from Synapse
Note that each file's synId and version is added to the data_sources vector via the download_file function.
The populated data_sources vector is used to populate Synapse provenance when the harmonized csv is uploaded.
```{r}
# 3xTG-AD
# metadata
threex_biospecimen <- download_file('syn23532198.3', type = "csv")
threex_individual <- download_file('syn23532199.4', type = "csv")
# data
threex_pathology_source <- download_file('syn26601504.1', type = "csv")

# 5xFAD
# metadata
fivex_biospecimen <- download_file('syn18876530.10', type = "csv")
fivex_individual <- download_file('syn18880070.14', type = "csv")
# data
fivex_gfap_s100b_source <- download_file('syn22049816.3', type = "csv")
fivex_lamp1_source <- download_file('syn22049817.2', type = "csv")
fivex_iba1_thios_source <- download_file('syn22049825.4', type = "csv")

#NSS
# metadata
nss_biospecimen <- download_file('syn29568452.3', type = "csv")
nss_individual <- download_file('syn27147690.3', type = "csv")
# data
nss_pathology_source <- download_file('syn42131930.1', type = "csv")

#Abca7
# metadata
abca7_biospecimen <- download_file('syn30859390.2', type = "csv")
abca7_individual <- download_file('syn30859394.2', type = "csv")
# data
abca7_pathology_source <- download_file('syn53127289.1', type = "csv")
```


## Generate model-specific intermediate files
File-measurement-unit and display value mappings for legacy models are from the v1 data spreadsheet here:
https://docs.google.com/spreadsheets/d/1B3ZMReC7nR2CGgpFNJuxJTj__lF2h6dt/edit?gid=1108411940#gid=1108411940

### 3xTG-AD
```{r}
threex_pathology <- threex_pathology_source %>%

        # Filter out rows with unused 'Stain' values
        filter(
                stain %in% stain_keeps

                # Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
        ) %>% pivot_longer(
        cols = c(objectDensity, averageObjectVolume, totalObjectVolume),
        names_to = "units",
        values_to = "measurement",
        values_transform = list(measurement = as.numeric)

        # Convert units to display values
) %>% mutate(
        units = case_match(
                units,
                'objectDensity' ~ objectDensity_display_name,
                'averageObjectVolume' ~ averageObjectVolume_display_name,
                'totalObjectVolume' ~ totalObjectVolume_display_name
        )

        # Map 'stain' values to 'type' display names
) %>% mutate(
        type = case_when(
                stain == 'IBA1' ~ IBA1_display_name,
                stain == 'GFAP' ~ GFAP_display_name,
                stain == 'S100B' ~ S100B_display_name,
                stain == 'LAMP1' ~ LAMP1_display_name,
                stain == 'AT8' ~ AT8_display_name,
                stain == 'HT7' ~ HT7_display_name,
                stain == 'ThioS' & units == averageObjectVolume_display_name ~ ThioS_size_display_name,
                stain == 'ThioS' & units == objectDensity_display_name ~ ThioS_density_display_name,
                # Drops ThioS totalObjectVolume measurements
                .default = 'UNMAPPED_DROPME'
        )

        # drop rows for unused ThioS stain/unit combinations
) %>% filter(
        type != 'UNMAPPED_DROPME'

        # Drop rows with NA measurements
) %>% filter(
        !is.na(measurement)

        # Populate model column
) %>% add_column(
        model = '3xTG-AD'

        # Drop unwanted columns
) %>% select(
        all_of(keeps_with_pool_join_field)
)

threex_pathology
```


### 5xFAD
```{r}
# Filter each source to remove unused rows by 'Stain' value
fivex_gfap_s100b <- fivex_gfap_s100b_source[fivex_gfap_s100b_source$Stain %in% c('GFAP', 'S100B'),]
fivex_lamp1 <- fivex_lamp1_source[fivex_lamp1_source$Stain == 'LAMP1',]
fivex_iba1_thios <- fivex_iba1_thios_source[fivex_iba1_thios_source$Stain %in% c('Iba1', 'ThioS'),]

# Pivot targeted measurement columns; capture colname as 'unit' and value as numeric 'measurement'
# GFAP/S100B source file
fivex_gfap_s100b_pivot <- fivex_gfap_s100b %>%
        pivot_longer(
                cols = c('X.objects..sqmm', 'mean.object.area..sqmm.'),
                names_to = "units",
                values_to = "measurement",
                values_transform = list(measurement = as.numeric)

                # Rename columns for consistency
        ) %>% rename(c('type' = 'Stain',
                       'individualID' = 'IndividualID',
                       'specimenID' = 'SpecimenID')

                     # Populate model column
) %>% add_column(
        model = '5xFAD'

        # Drop unused columns
) %>% select(
        all_of(keeps_with_join_fields)
)


# LAMP1 source file
fivex_lamp1_pivot <- fivex_lamp1 %>%
        pivot_longer(
                cols = c('Area..'),
                names_to = "units",
                values_to = "measurement",
                values_transform = list(measurement = as.numeric),

                # Rename columns for consistency
        ) %>% rename(c('type' = 'Stain',
                       'individualID' = 'IndividualID',
                       'specimenID' = 'SpecimenID')

                     # Populate model column
) %>% add_column(
        model = '5xFAD'

        # Drop unused columns
) %>% select(
        all_of(keeps_with_join_fields)
)


# IBA1/ThioS source file
fivex_iba1_thios_pivot <- fivex_iba1_thios %>%
        pivot_longer(
                cols = c('X.objects..sqmm', 'mean.object.area..squm.'),
                names_to = "units",
                values_to = "measurement",
                values_transform = list(measurement = as.numeric),

                # Rename columns for consistency
        ) %>% rename(c('type' = 'Stain',
                       'individualID' = 'IndividualID',
                       'specimenID' = 'SpecimenID')

                     # Populate model column
) %>% add_column(
        model = '5xFAD'

        # Drop unused columns
) %>% select(
        all_of(keeps_with_join_fields)
)


# Merge the three sources into a single data frame
fivex_pathology <- bind_rows(fivex_gfap_s100b_pivot, fivex_lamp1_pivot, fivex_iba1_thios_pivot

                             # Convert units to display values
) %>% mutate(
        units = case_match(
                units,
                'X.objects..sqmm' ~ objectDensity_display_name,
                'mean.object.area..sqmm.' ~ meanObjectArea_display_name,
                'Area..' ~ areaPercent_display_name,
                'X.objects..sqmm' ~ objectDensity_display_name,
                'mean.object.area..squm.' ~ meanObjectArea_display_name
        )

        # Map stain values to type display names
) %>% mutate(
        type = case_when(
                type == 'Iba1' ~ IBA1_display_name,
                type == 'GFAP' ~ GFAP_display_name,
                type == 'S100B' ~ S100B_display_name,
                type == 'LAMP1' ~ LAMP1_display_name,
                type == 'AT8' ~ AT8_display_name,
                type == 'HT7' ~ HT7_display_name,
                type == 'ThioS' & units == meanObjectArea_display_name ~ ThioS_size_display_name,
                type == 'ThioS' & units == objectDensity_display_name ~ ThioS_density_display_name,
                # Drops ThioS totalObjectVolume measurements
                .default = 'UNMAPPED_DROPME'
        )

        # drop rows for unused ThioS stain/unit combinations
) %>% filter(
        type != 'UNMAPPED_DROPME'

        # Drop rows with NA measurements
) %>% filter(
        !is.na(measurement)

        # Drop unwanted columns
) %>% select(
        all_of(keeps_with_join_fields)
)

fivex_pathology
```



### Trem2-R47H_NSS
```{r}
nss_pathology <- nss_pathology_source %>%

        # Filter out rows with unused 'Stain' values
        filter(
                stain %in% stain_keeps

                # Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
        ) %>% pivot_longer(
        cols = c(objectDensity, averageObjectVolume, totalObjectVolume),
        names_to = "units",
        values_to = "measurement",
        values_transform = list(measurement = as.numeric)

        # Convert units to display values
) %>% mutate(
        units = case_match(
                units,
                'objectDensity' ~ objectDensity_display_name,
                'averageObjectVolume' ~ averageObjectVolume_display_name,
                'totalObjectVolume' ~ totalObjectVolume_display_name
        )

        # Map 'stain' values to 'type' display names
) %>% mutate(
        type = case_when(
                stain == 'IBA1' & units == objectDensity_display_name ~ IBA1_display_name,
                stain == 'GFAP' & units == objectDensity_display_name ~ GFAP_display_name,
                stain == 'S100B' & units == objectDensity_display_name ~ S100B_display_name,
                stain == 'LAMP1' ~ LAMP1_display_name,
                stain == 'AT8' ~ AT8_display_name,
                stain == 'HT7' ~ HT7_display_name,
                stain == 'ThioS' & units == averageObjectVolume_display_name ~ ThioS_size_display_name,
                stain == 'ThioS' & units == objectDensity_display_name ~ ThioS_density_display_name,
                # Set default value for rows with unused measurements
                .default = 'UNMAPPED_DROPME'
        )

        # drop rows for unused stain/unit combinations
) %>% filter(
        type != 'UNMAPPED_DROPME'

        # Drop rows with NA measurements
) %>% filter(
        !is.na(measurement)

        # Populate model column
) %>% add_column(
        model = 'Trem2-R47H_NSS'

        # Drop unwanted columns
) %>% select(
        all_of(keeps_with_pool_join_field)
)

nss_pathology

```

### Abca7*V1599M
```{r}
abca7_pathology <- abca7_pathology_source %>%

        # Filter out rows with unused 'Stain' values
        filter(
                stain %in% stain_keeps

                # Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
        ) %>% pivot_longer(
        cols = c(objectDensity, averageObjectVolume, totalObjectVolume),
        names_to = "units",
        values_to = "measurement",
        values_transform = list(measurement = as.numeric)

        # Convert units to display values
) %>% mutate(
        units = case_match(
                units,
                'objectDensity' ~ objectDensity_display_name,
                'averageObjectVolume' ~ averageObjectVolume_display_name,
                'totalObjectVolume' ~ totalObjectVolume_display_name
        )

        # Map 'stain' values to 'type' display names
) %>% mutate(
        type = case_when(
                stain == 'IBA1' & units == objectDensity_display_name ~ IBA1_display_name,
                stain == 'GFAP' & units == objectDensity_display_name ~ GFAP_display_name,
                stain == 'S100B' & units == objectDensity_display_name ~ S100B_display_name,
                stain == 'LAMP1' & units == totalObjectVolume_display_name ~ LAMP1_display_name,
                stain == 'AT8' ~ AT8_display_name,
                stain == 'HT7' ~ HT7_display_name,
                stain == 'ThioS' & units == averageObjectVolume_display_name ~ ThioS_size_display_name,
                stain == 'ThioS' & units == objectDensity_display_name ~ ThioS_density_display_name,
                # Set default value for rows with unused measurements
                .default = 'UNMAPPED_DROPME'
        )

        # drop rows for unused stain/unit combinations
) %>% filter(
        type != 'UNMAPPED_DROPME'

        # Drop rows with NA measurements
) %>% filter(
        !is.na(measurement)

        # Populate model column
) %>% add_column(
        model = 'Abca7*V1599M'

        # Drop unwanted columns
) %>% select(
        all_of(keeps_with_pool_join_field)
)

abca7_pathology

```


## Sanity check nrow in generated files
```{r}
# 3xTG-AD
# IBA1 Correct for "N/A" values in targeted source column: objectDensity
threex_iba1 <- threex_pathology_source[threex_pathology_source$stain == 'IBA1',]
threex_iba1_nas <- sum(is.na(threex_iba1$objectDensity))
stopifnot(nrow(threex_pathology[threex_pathology$type == IBA1_display_name,]) == nrow(threex_iba1) - threex_iba1_nas)

# GFAP Correct for "N/A" values in targeted source column: objectDensity
threex_gfap <- threex_pathology_source[threex_pathology_source$stain == 'GFAP',]
threex_gfap_nas <- sum(is.na(threex_gfap$objectDensity))
stopifnot(nrow(threex_pathology[threex_pathology$type == GFAP_display_name,]) == nrow(threex_gfap) - threex_gfap_nas)

# S100B Correct for "N/A" values in targeted source column: objectDensity
threex_s100b <- threex_pathology_source[threex_pathology_source$stain == 'S100B',]
threex_s100b_nas <- sum(is.na(threex_s100b$objectDensity))
stopifnot(nrow(threex_pathology[threex_pathology$type == S100B_display_name,]) == nrow(threex_s100b) - threex_s100b_nas)

# LAMP1 Correct for "N/A" values in targeted source column: totalObjectVolume
threex_lamp1 <- threex_pathology_source[threex_pathology_source$stain == 'LAMP1',]
threex_lamp1_nas <- sum(is.na(threex_lamp1$totalObjectVolume))
stopifnot(nrow(threex_pathology[threex_pathology$type == LAMP1_display_name,]) == nrow(threex_lamp1) - threex_lamp1_nas)

# AT8 Correct for "N/A" values in targeted source column: totalObjectVolume
threex_at8 <- threex_pathology_source[threex_pathology_source$stain == 'AT8',]
threex_at8_nas <- sum(is.na(threex_at8$totalObjectVolume))
stopifnot(nrow(threex_pathology[threex_pathology$type == AT8_display_name,]) == nrow(threex_at8) - threex_at8_nas)

# HT7 Correct for "N/A" values in targeted source column: totalObjectVolume
threex_ht7 <- threex_pathology_source[threex_pathology_source$stain == 'HT7',]
threex_ht7_nas <- sum(is.na(threex_at8$totalObjectVolume))
stopifnot(nrow(threex_pathology[threex_pathology$type == HT7_display_name,]) == nrow(threex_ht7) -threex_ht7_nas)

# ThioS size Correct for "N/A" values in targeted source column: averageObjectVolume
threex_thioss <- threex_pathology_source[threex_pathology_source$stain == 'ThioS',]
threex_thioss_nas <- sum(is.na(threex_thioss$averageObjectVolume))
stopifnot(nrow(threex_pathology[threex_pathology$type == ThioS_size_display_name,]) == nrow(threex_thioss) - threex_thioss_nas)

# Thios density Correct for "N/A" values in targeted source column: objectDensity
threex_thiosd <- threex_pathology_source[threex_pathology_source$stain == 'ThioS',]
threex_thiosd_nas <- sum(is.na(threex_thiosd$objectDensity))
stopifnot(nrow(threex_pathology[threex_pathology$type == ThioS_density_display_name,]) == nrow(threex_thiosd) - threex_thiosd_nas)


# 5xFAD
# Iba1 Correct for "N/A" values in targeted source columns: 'X.objects..sqmm', 'mean.object.area..squm.'
fivex_ibas <- fivex_iba1_thios_source[fivex_iba1_thios_source$Stain == 'Iba1',]
fivex_iba_nas <- sum(fivex_ibas$X.objects..sqmm == "N/A") + sum(fivex_ibas$mean.object.area..squm. == "N/A")
stopifnot(nrow(fivex_pathology[fivex_pathology$type == IBA1_display_name,]) == nrow(fivex_ibas) * 2 - fivex_iba_nas)

# GFAP Correct for "N/A" values in targeted source columns: 'X.objects..sqmm', 'mean.object.area..sqmm.'
fivex_gfap <- fivex_gfap_s100b_source[fivex_gfap_s100b_source$Stain == 'GFAP',]
fivex_gfap_nas <- sum(fivex_gfap$X.objects..sqmm == "N/A") + sum(fivex_gfap$mean.object.area..sqmm. == "N/A")
stopifnot(nrow(fivex_pathology[fivex_pathology$type == GFAP_display_name,]) == nrow(fivex_gfap) * 2 - fivex_gfap_nas)

# S100B Correct for "N/A" values in targeted source column: 'X.objects..sqmm'
fivex_s100bs <- fivex_gfap_s100b_source[fivex_gfap_s100b_source$Stain == 'S100B',]
fivex_s100b_nas <- sum(fivex_s100bs$X.objects..sqmm == "N/A")
stopifnot(nrow(fivex_pathology[fivex_pathology$type == S100B_display_name,]) == nrow(fivex_s100bs) - fivex_s100b_nas)

# LAMP1 Correct for "N/A" values in targeted source column: 'Area..'
fivex_lamp1s <- fivex_lamp1_source[fivex_lamp1_source$Stain == 'LAMP1',]
fivex_lamp1_nas <- sum(fivex_lamp1s$Area.. == "N/A")
stopifnot(nrow(fivex_pathology[fivex_pathology$type == LAMP1_display_name,]) == nrow(fivex_lamp1s) - fivex_lamp1_nas)

# ThioS size Correct for N/A values in targeted source column: 'mean.object.area..squm.'
fivex_thios <- fivex_lamp1_source[fivex_lamp1_source$Stain == 'ThioS',]
fivex_thios_size_nas <- !is.na(fivex_thios$mean.object.area..squm)
stopifnot(nrow(fivex_pathology[fivex_pathology$type == ThioS_size_display_name,]) == nrow(fivex_thios) - fivex_thios_size_nas)

# ThioS density Correct for N/A values in targeted source column: 'X.objects..sqmm'
thios_density_nas <-is.na(fivex_thios$X.objects..sqmm)
stopifnot(nrow(fivex_pathology[fivex_pathology$type == ThioS_density_display_name,]) == nrow(fivex_thios) - thios_density_nas)


# Trem2-R47H_NSS
# IBA1 Correct for "N/A" values in targeted source column: objectDensity
nss_iba1 <- nss_pathology_source[nss_pathology_source$stain == 'IBA1',]
nss_iba1_nas <- sum(is.na(nss_iba1$objectDensity))
stopifnot(nrow(nss_pathology[nss_pathology$type == IBA1_display_name,]) == nrow(nss_iba1) - nss_iba1_nas)

# GFAP Correct for "N/A" values in targeted source column: objectDensity
nss_gfap <- nss_pathology_source[nss_pathology_source$stain == 'GFAP',]
nss_gfap_nas <- sum(is.na(nss_gfap$objectDensity))
stopifnot(nrow(nss_pathology[nss_pathology$type == GFAP_display_name,]) == nrow(nss_gfap) - nss_gfap_nas)

# S100B Correct for "N/A" values in targeted source column: objectDensity
nss_s100b <- nss_pathology_source[nss_pathology_source$stain == 'S100B',]
nss_s100b_nas <- sum(is.na(nss_s100b$objectDensity))
stopifnot(nrow(nss_pathology[nss_pathology$type == S100B_display_name,]) == nrow(nss_s100b) - nss_s100b_nas)

# LAMP1 Correct for "N/A" values in targeted source column: totalObjectVolume
nss_lamp1 <- nss_pathology_source[nss_pathology_source$stain == 'LAMP1',]
nss_lamp1_nas <- sum(is.na(nss_lamp1$totalObjectVolume))
stopifnot(nrow(nss_pathology[nss_pathology$type == LAMP1_display_name,]) == nrow(nss_lamp1) - nss_lamp1_nas)

# ThioS size Correct for "N/A" values in targeted source column: averageObjectVolume
nss_thios <- nss_pathology_source[nss_pathology_source$stain == 'ThioS',]
nss_thioss_nas <- sum(is.na(nss_thios$averageObjectVolume))
stopifnot(nrow(nss_pathology[nss_pathology$type == ThioS_size_display_name,]) == nrow(nss_thios) - nss_thioss_nas)

# Thios density Correct for "N/A" values in targeted source column: objectDensity
nss_thiosd_nas <- sum(is.na(nss_thios$objectDensity))
stopifnot(nrow(nss_pathology[nss_pathology$type == ThioS_density_display_name,]) == nrow(nss_thios) - nss_thiosd_nas)


# Abca7*V1599M
# IBA1 Correct for "N/A" values in targeted source column: objectDensity
abca7_iba1 <- abca7_pathology_source[abca7_pathology_source$stain == 'IBA1',]
abca7_iba1_nas <- sum(abca7_iba1$objectDensity == "N/A")
stopifnot(nrow(abca7_pathology[abca7_pathology$type == IBA1_display_name,]) == nrow(abca7_iba1) - abca7_iba1_nas)

# GFAP Correct for "N/A" values in targeted source column: objectDensity
abca7_gfap <- abca7_pathology_source[abca7_pathology_source$stain == 'GFAP',]
abca7_gfap_nas <- sum(abca7_gfap$objectDensity == "N/A")
stopifnot(nrow(abca7_pathology[abca7_pathology$type == GFAP_display_name,]) == nrow(abca7_gfap) - abca7_gfap_nas)

# S100B Correct for "N/A" values in targeted source column: objectDensity
abca7_s100b <- abca7_pathology_source[abca7_pathology_source$stain == 'S100B',]
abca7_s100b_nas <- sum(abca7_s100b$objectDensity == "N/A")
stopifnot(nrow(abca7_pathology[abca7_pathology$type == S100B_display_name,]) == nrow(abca7_s100b) - abca7_s100b_nas)

# LAMP1 Correct for "N/A" values in targeted source column: totalObjectVolume
abca7_lamp1 <- abca7_pathology_source[abca7_pathology_source$stain == 'LAMP1',]
abca7_lamp1_nas <- sum(is.na(abca7_lamp1$totalObjectVolume))
stopifnot(nrow(abca7_pathology[abca7_pathology$type == LAMP1_display_name,]) == nrow(abca7_lamp1) - abca7_lamp1_nas)

# ThioS size Correct for "N/A" values in targeted source column: averageObjectVolume
abca7_thios <- abca7_pathology_source[abca7_pathology_source$stain == 'ThioS',]
abca7_thioss_nas <- sum(abca7_thios$averageObjectVolume == "N/A")
stopifnot(nrow(abca7_pathology[abca7_pathology$type == ThioS_size_display_name,]) == nrow(abca7_thios) - abca7_thioss_nas)

# Thios density Correct for "N/A" values in targeted source column: objectDensity
abca7_thiosd_nas <- sum(abca7_thios$objectDensity == "N/A")
stopifnot(nrow(abca7_pathology[abca7_pathology$type == ThioS_density_display_name,]) == nrow(abca7_thios) - abca7_thiosd_nas)
```



## Join source files to metadata to populate age, tissue, and genotype values
```{r}
# 3xTG-AD
threex_final <- join_to_metadata(threex_individual, threex_biospecimen, threex_pathology)[keeps]

# 5xFAD
fivex_with_8mo <- join_to_metadata(fivex_individual, fivex_biospecimen, fivex_pathology)[keeps]
# drop unwanted 8 month samples from 5xFAD
fivex_final <- fivex_with_8mo[fivex_with_8mo$ageDeath != 8,]

# Trem2-R47H_NSS
nss_final <- join_to_metadata(nss_individual, nss_biospecimen, nss_pathology)[keeps]

# Abca7*V1599M
abca7_final <- join_to_metadata(abca7_individual, abca7_biospecimen, abca7_pathology)[keeps]
```


## Merge files, write csv file, and upload to synapse
```{r}
pathology_merged <- bind_rows(threex_final, fivex_final, nss_final, abca7_final)


# standardize genotype values for display
pathology_final <- mutate(pathology_merged, genotype = case_when(

        # 3xTG-AD genotypes
        genotype == "3xTg-AD_homozygous"  ~ "3xTG-AD",
        genotype == "3XTg-AD_noncarrier" ~ "B6129",

        # 5xFAD genotypes
        genotype %in% c("5XFAD_hemizygous","5XFAD_carrier") ~ "5xFAD",
        genotype == "5XFAD_noncarrier"  ~ "C57BL/6J",

        # NSS genotypes
        genotype == "Trem2-R47H_NSS_homozygous"  ~ "Trem2-R47H_NSS",
        genotype == "5XFAD_carrier, Trem2-R47H_NSS_homozygous"  ~ "5xFADTrem2-R47H_NSS",

        # Abca7 genotypes
        genotype == "Abca7-V1599M_homozygous"  ~ "Abca7*V1599M",
        genotype == "5XFAD_carrier, Abca7-V1599M_homozygous"  ~ "5xFADAbca7*V1599M",

        TRUE ~ genotype))

# print final data frame below
pathology_final

# write csv file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}
write.csv(pathology_final, file.path(output_dir, "pathology.csv"), row.names = FALSE)

# upload to synapse
syn_file <- File(file.path(output_dir, "pathology.csv"),
                 description = "Harmonized Pathology data for Model-AD Explorer 2.0.",
                 parent = synapse_folder)

res <- synStore(syn_file, forceVersion = FALSE,
                used = data_sources,
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-40_Create_harmonized_pathology_source_file.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```

