
# Create Biomarker source file (MG-88)

This notebook generates a single harmonized source file (syn61250724) containing the data required to surface Biomarker evidence in the Model-AD Explorer.

Models that are currently handled by this notebook:
* 3xTG-AD
* 5xFAD
* Trem2-R47H_NSS
* Abca7*V1599M

ADDING NEW MODELS
As data becomes available for additional models, the data for those models will need to be incorporated into the harmonized source file.

To incorporate each new model into this notebook:

1. 'Download data and metadata files from Synapse' section:
- Add a section to download the new model's metadata and data files

2. 'Generate model-specific intermediate files' section
- Add a code block that generates an intermediate results file for the new model

3. 'Join source files to metadata to populate age, tissue, and genotype values' section
- Update to join metadata to the new model's intermediate file

4. 'Sanity check nrow in generated files' section
- Add validation for the new model's intermediate file

5. 'Merge files, write csv file, and upload to synapse' section
- Add the new model's intermediate file to the bind_rows call on the first line
- Add genotype-genotype display name mappings for the new model


ADDING NEW MEASURES
As additional measurement types become available, those measurement types will need to be incorporated into the harmonized source file.

To incorporate a new measurement type into the notebook:

1. 'Download data and metadata files from Synapse' section
- Add the data file(s) containing the new measurement to the appropriate model-specific section(s)
- Adjust the metadata file version(s) as required to bring in metadata for the new measurement

2. 'Generate model-specific intermediate files' section
- For each model, add the code required to process the new source file(s)

4. 'Sanity check nrow in generated files' section
- Add a sanity check for the new measure for each model that has it


## Library setup
Install required packages:
- synapser (also requires a Synapse account with an auth token)
- tidyverse
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}

if (!require("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

library(synapser)
library(tidyverse)
```

## Utility functions & variables
```{r}
# *******
# FILTERS
# *******

# column spec for intermediate per-model result files
intermediate_keeps <- c('specimenID', 'individualID', 'model', 'type', 'measurement', 'units')

# column specs for final harmonized output file
keeps <- c('individual_id', 'model', 'type', 'measurement', 'units', 'age_death', 'tissue', 'sex', 'genotype')

# ages that we want to display results for
age_keeps <- c(4, 12, 18)

# **************
# DISPLAY LABELS
# **************

# Model and genotype display names (rnaseq_genotype_label_map.csv)
threex_display_name <- '3xTg-AD'
fivex_display_name <- '5xFAD (UCI)'
abca7_display_name <- 'Abca7*V1599M'
nss_display_name <- 'Trem2-R47H_NSS'

# *********
# FUNCTIONS
# *********

# Vector to store all downloaded synIds for Synapse provenance (populated via download_csv function)
data_sources <- character()

# Download a csv file from Synapse by synId and read it into a dataframe
# - synId: synapse ID of the file to download
# - df_name: the name to use for the resulting dataframe
download_csv <- function(synId, df_name) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("./", "input")
  id_parts <- strsplit(synId, '.', fixed = TRUE)
  id_value <- id_parts[[1]][1]
  version_value <- id_parts[[1]][2]

  if (is.na(version_value)) {
    version_value <- NULL
  }

  cat(paste0("Downloading ", synId, "..."))

  file <- synGet(id_value, version = version_value, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId, "\n- Modified on ", file$properties$modifiedOn, "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  data <- read.csv(path)
  df_name <- as.data.frame(data)
}

transform_results <- function(results, measurement_columns, model_name, fields_to_keep) {

# If no fraction_type column, add one with empty strings
if (!"fraction_type" %in% names(results)) { results$fraction_type <- "" }

results %>% pivot_longer(
                cols = all_of(measurement_columns),
                names_to = "type",
                values_to = "measurement"

       #
       ) %>% mutate(type = if_else(type %in% c("Abeta40", "Abeta42"), paste(fraction_type, type), type)

        # drop rows with no measurement
        ) %>% filter(
                !is.na(measurement)

        # Populate model column
        ) %>% add_column(
                model = model_name

        # Drop unused columns
        ) %>% select(
                all_of(fields_to_keep)
        )
}

# Join a data file to a pair of metadata files to get required age, sex, ageDeath, and genotype values
# - df: A dataframe containing result rows
# - individual_metadata: A dataframe containing individual-level metadata
# - biospecimen_metadata: A dataframe containing specimen metadata
join_to_metadata <- function(results, individual_metadata, biospecimen_metadata) {

  # Join metadata files by individualID
  individual_metadata <- select(individual_metadata, individualID, sex, ageDeath, genotype)
  biospecimen_metadata <- select(biospecimen_metadata, individualID, specimenID, tissue)
  metadata <- merge(individual_metadata, biospecimen_metadata, by = 'individualID')

  # Join results to metadata
  results %>% left_join(metadata, by = c('individualID', 'specimenID')
            ) %>% left_join(genotype_label_map, by = c('model', 'genotype')

            # Correct genotypes to align with target display values
            ) %>% mutate(genotype = display_label

            # Print out nrow of results before age filtering
            ) %>% { print(nrow(.)) ;.

            # Filter out results for unwanted ages
            } %>% filter(ageDeath %in% age_keeps

            # Print out nrow of results after age filtering
            ) %>% { print(nrow(.)) ;.

            # Convert all individualID columns to string type
            } %>% mutate(individualID = as.character(individualID)

            # Rename columns
            ) %>% rename(individual_id = individualID,
                         age_death = ageDeath

            # Drop unneeded columns
            ) %>% select(all_of(keeps))
}
```


## Download data and metadata files from Synapse
Note that each file's synId and version is added to the data_sources vector via the download_csv function.
The populated data_sources vector is used to populate Synapse provenance when the harmonized csv is uploaded.
```{r}
# Genotype -> Display label map
genotype_label_map <- download_csv('syn69014401.14') %>%
  # Model name fix, plus change "5xFAD (UCI)" -> "5xFAD"
  mutate(model = model_group,
         display_label = str_replace(display_label, " \\(UCI\\)", ""))

# 3xTG-AD
# metadata
threex_biospecimen <- download_csv('syn23532198.3')
threex_individual <- download_csv('syn23532199.4') %>%
  # genotype fix
  mutate(genotype = str_replace(genotype, "_homozygous", "_carrier"),
         genotype = str_replace(genotype, "3XTg", "3xTg"))
# Abetas
threex_abs_sol_source <- download_csv('syn25913411.7')
threex_abs_insol_source <- download_csv('syn25913418.8')
# NO NFL DATA

# 5xFAD
# metadata
fivex_biospecimen <- download_csv('syn18876530.10')
fivex_individual <- download_csv('syn18880070.14') %>%
  # genotype fix
  mutate(genotype = str_replace(genotype, "_hemizygous", "_carrier"))
# Abetas
fivex_abs_sol_source <- download_csv('syn22101767.3')
fivex_abs_insol_source <- download_csv('syn22101766.1')
# NO NFL DATA

#NSS
# metadata
nss_biospecimen <- download_csv('syn29568452.3')
nss_individual <- download_csv('syn27147690.3') %>%
  # genotype fix
  mutate(genotype = str_replace(genotype, ",", ";"))
# Abetas
nss_abs_sol_4mo_source <- download_csv('syn29568621.2')
nss_abs_sol_12mo_source <- download_csv('syn29568624.2')
nss_abs_insol_4mo_source <- download_csv('syn29568619.1')
nss_abs_insol_12mo_source <- download_csv('syn29568622.1')
# NFL
nss_nfl_cortex_source <- download_csv('syn29568489.2')
nss_nfl_plasma_source <- download_csv('syn29568490.1')

#Abca7
# metadata
abca7_biospecimen <- download_csv('syn30859390.2')
abca7_individual <- download_csv('syn30859394.2') %>%
  # genotype fix
  mutate(genotype = str_replace(genotype, ",", ";"))
# Abetas
abca7_abs_sol_4mo_source <- download_csv('syn30859386.1')
abca7_abs_sol_12mo_source <- download_csv('syn30859377.1')
abca7_abs_insol_4mo_source <- download_csv('syn30859383.1')
abca7_abs_insol_12mo_source <- download_csv('syn30859387.1')
# NFL
abca7_nfl_cortex_source <- download_csv('syn30859379.1')
abca7_nfl_plasma_source <- download_csv('syn30859380.1')
```


## Generate per-model intermediate files with one measurement per row

### 3xTG-AD
```{r}
# Abeta measures
# Capture fraction types from file names in temp 'fraction_type' column and merge resulting dfs
threex_abs_sol <- threex_abs_sol_source %>% mutate(fraction_type = 'Soluble')
threex_abs_insol <- threex_abs_insol_source %>% mutate(fraction_type = 'Insoluble')
threex_abs_merged <- bind_rows(threex_abs_sol, threex_abs_insol)

# Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
threex_abs <- transform_results(threex_abs_merged, c("Abeta40", "Abeta42"), threex_display_name, intermediate_keeps)

# No NfL measures for this model
threex_biomarkers <- threex_abs

# Join results to metadata
threex_final <- join_to_metadata(threex_biomarkers, threex_individual, threex_biospecimen)
```

### 5xFAD
```{r}
# Abeta measures
# Capture fraction types from file names in temp 'fraction_type' column and merge resulting dfs
fivex_abs_sol <- fivex_abs_sol_source %>% mutate(fraction_type = 'Soluble')
fivex_abs_insol <- fivex_abs_insol_source %>% mutate(fraction_type = 'Insoluble')
fivex_abs_merged <- bind_rows(fivex_abs_sol, fivex_abs_insol

    # MG-268: Replace N/A measurement values with 0s
    ) %>% mutate(
            Abeta40 = ifelse(is.na(Abeta40), 0, Abeta40),
            Abeta42 = ifelse(is.na(Abeta42), 0, Abeta42)
)

# Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
fivex_abs <- transform_results(fivex_abs_merged, c("Abeta40", "Abeta42"), fivex_display_name, intermediate_keeps)

# No NfL measures for this model
fivex_biomarkers <- fivex_abs

# Join results to metadata
fivex_final <- join_to_metadata(fivex_biomarkers, fivex_individual, fivex_biospecimen)
```


### Trem2-R47H_NSS
```{r}
# Abeta measures
# Convert columns to consistent types
nss_abs_sol_12mo_source$individualID <- as.character(nss_abs_sol_12mo_source$individualID)
nss_abs_insol_12mo_source$individualID <- as.character(nss_abs_insol_12mo_source$individualID)
nss_abs_insol_12mo_source$Abeta42 <- suppressWarnings(as.numeric(nss_abs_insol_12mo_source$Abeta42))

# Capture fraction types from file names in temp 'fraction_type' column and merge resulting dfs
nss_abs_sol_4mo <- nss_abs_sol_4mo_source %>% mutate(fraction_type = 'Soluble')
nss_abs_sol_12mo <- nss_abs_sol_12mo_source %>% mutate(fraction_type = 'Soluble')
nss_abs_insol_4mo <- nss_abs_insol_4mo_source %>% mutate(fraction_type = 'Insoluble')
nss_abs_insol_12mo <- nss_abs_insol_12mo_source %>% mutate(fraction_type = 'Insoluble')
nss_abs_merged <- bind_rows(nss_abs_sol_4mo, nss_abs_sol_12mo, nss_abs_insol_4mo, nss_abs_insol_12mo)

# Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
nss_abs <- transform_results(nss_abs_merged, c("Abeta40", "Abeta42"), nss_display_name, intermediate_keeps)

# NfL measures
# Merge source dfs
nss_nfl_merged <- bind_rows(nss_nfl_cortex_source, nss_nfl_plasma_source)

# Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
nss_nfl <- transform_results(nss_nfl_merged, c("NfL"), nss_display_name, intermediate_keeps)

# Merge Abetas and NfL
nss_biomarkers <- bind_rows(nss_abs, nss_nfl)

# Join results to metadata
nss_final <- join_to_metadata(nss_biomarkers, nss_individual, nss_biospecimen)
```

### Abca7*V1599M
```{r}
# Abeta measures
# Capture fraction types from file names in temp 'fraction_type' column and merge resulting dfs
abca7_abs_sol_4mo <- abca7_abs_sol_4mo_source %>% mutate(fraction_type = 'Soluble')
abca7_abs_sol_12mo <- abca7_abs_sol_12mo_source %>% mutate(fraction_type = 'Soluble')
abca7_abs_insol_4mo <- abca7_abs_insol_4mo_source %>% mutate(fraction_type = 'Insoluble')
abca7_abs_insol_12mo <- abca7_abs_insol_12mo_source %>% mutate(fraction_type = 'Insoluble')
abca7_abs_merged <- bind_rows(abca7_abs_sol_4mo, abca7_abs_sol_12mo, abca7_abs_insol_4mo, abca7_abs_insol_12mo)

# Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
abca7_abs <- transform_results(abca7_abs_merged, c("Abeta40", "Abeta42"), abca7_display_name, intermediate_keeps)

# NfL measures
# Merge source dfs
abca7_nfl_merged <- bind_rows(abca7_nfl_cortex_source, abca7_nfl_plasma_source)

# Pivot targeted measurement columns; capture colname as 'unit' and value as 'measurement'
abca7_nfl <- transform_results(abca7_nfl_merged, c("NfL"), abca7_display_name, intermediate_keeps)

# Merge Abeta and NfL measures
abca7_biomarkers <- bind_rows(abca7_abs, abca7_nfl)

# Join results to metadata
abca7_final <- join_to_metadata(abca7_biomarkers, abca7_individual, abca7_biospecimen)
```

## Sanity check generated files
Checks that ensure that the processed data includes the expected number of results per model, per result type:
1. Source data vs processed data: nrow(filtered_source_data) == nrow(processed_data)
2. Source data vs expected rows: nrow(nrow(processed_data)) == nrow(legacy_explorer_data)
   - Expected row values are calculated in the 'Immunohisto expected row counts' notebook
3. There should be no NA values: all values with NA measurements should be dropped, and there shouldn't be any NA values introduced from the `left_join()` to metadata.

```{r}
# 3xTG-AD
# Correct for NA values in source data
threex_expected <- nrow(threex_abs_merged[!is.na(threex_abs_merged$Abeta40),]) + nrow(threex_abs_merged[!is.na(threex_abs_merged$Abeta42),])
# Check corrected source data vs processed data
stopifnot(threex_expected == nrow(threex_final))
# Source data vs Expected rows
stopifnot(nrow(threex_final) == 439) # 3xTg-AD + B6129: 439
# Check all fields were populated properly -- there should be no NAs
stopifnot(!any(is.na(threex_final)))

# 5xFAD
# Correct for 104 rows removed via age filtering (from join_to_metadata print statement)
fivex_expected <- nrow(fivex_abs_merged) * 2 - 104
# Check corrected source data vs processed data
stopifnot(fivex_expected == nrow(fivex_final))
# Source data vs Expected rows
stopifnot(nrow(fivex_final) == 288) # 5xFAD + C57BL/6J (w/o 8mo): 288
# Check all fields were populated properly -- there should be no NAs
stopifnot(!any(is.na(fivex_final)))

# Trem2-R47H_NSS
# Correct for NA values in source data
nss_expected_abs <- nrow(nss_abs_merged[!is.na(nss_abs_merged$Abeta40),]) + nrow(nss_abs_merged[!is.na(nss_abs_merged$Abeta42),])
nss_expected_nfl <- nrow(nss_nfl_merged[!is.na(nss_nfl_merged$NfL),])
# Check corrected source data vs processed data
stopifnot(nss_expected_abs + nss_expected_nfl == nrow(nss_final))
# Source data vs Expected rows
stopifnot(nrow(nss_final) == 493) # 5xFADTrem2-R47H_NSS + 5xFAD + Trem2-R47H_NSS + C57BL/6J: 493
# Check all fields were populated properly -- there should be no NAs
stopifnot(!any(is.na(nss_final)))

# Abca7*V1599M
# Correct for NA values in source data
abca7_expected_abs <- nrow(abca7_abs_merged[!is.na(abca7_abs_merged$Abeta40),]) + nrow(abca7_abs_merged[!is.na(abca7_abs_merged$Abeta42),])
abca7_expected_nfl <- nrow(abca7_nfl_merged)
# Check corrected source data vs processed data
stopifnot(abca7_expected_abs + abca7_expected_nfl == nrow(abca7_final))
# Source data vs Expected rows
stopifnot(nrow(abca7_final) == 543) # 5xFADAbca7*V1599M + 5xFAD + Abca7*V1599M + C57BL/6J: 543
# Check all fields were populated properly -- there should be no NAs
stopifnot(!any(is.na(abca7_final)))
```

## Merge model-specific results files & write csv file to synapse
```{r}
# Merge results for all models
biomarkers_final <- bind_rows(
      threex_final,
      fivex_final,
      nss_final,
      abca7_final
)

# write csv file
output_dir <- file.path("./", "output")
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}
write.csv(biomarkers_final, file.path(output_dir, "biomarkers.csv"), row.names = FALSE)

# upload to synapse
syn_file <- File(file.path(output_dir, "biomarkers.csv"),
                 description = "Harmonized Biomarker data for Model-AD Explorer 2.0.",
                 parent = "syn51498054")

res <- synStore(syn_file, forceVersion = FALSE,
                used = data_sources,
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-88_Create_harmonized_biomarkers_source_file.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```
