# Create ENSMUSG Reference

This notebook downloads several key files in the `model_ad_prod.yaml` config file, gets a list of all the Ensembl gene IDs used in those files, and maps them to gene symbols and aliases. The file is then uploaded to Synapse here: [syn51615422](https://www.synapse.org/#!Synapse:syn51615422).

Required packages (will automatically be installed if necessary):

-   synapser (also requires a Synapse account with an auth token)
-   BiocManager
-   biomaRt
-   tidyverse libraries dplyr, stringr, purrr
-   jsonlite
-   yaml

## Library setup

```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "https://cloud.r-project.org"))
}

if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

if (!require("biomaRt", quietly = TRUE)) {
  BiocManager::install("biomaRt")
}

if (!require("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

if (!require("jsonlite", quietly = TRUE)) {
  install.packages("jsonlite")
}

if (!require("yaml", quietly = TRUE)) {
  install.packages("yaml")
}

library(synapser)
library(biomaRt)
library(dplyr)
library(stringr)
library(purrr)
library(jsonlite)
library(yaml)
```

## Download the data source files

We load the following files to get coverage of all genes in the Model-AD data:

-   `human_transgene_allele_map`: All human transgenes across all models

-   `biodom_genes_mm`: all genes we have biodomain info for

-   `*_differential_expression`: all genes for differential expression results.

    -   Corresponding differential expression and individual expression files have the exact same set of Ensembl IDs, so we only need to load one file from the pair to get all the genes for that study. DE files are much smaller, so we load those.

```{r}
input_dir <- file.path("..", "input")
dir.create(input_dir, showWarnings = FALSE)

cfg <- read_yaml("../../../configs/model_ad_prod.yaml", eval.expr = TRUE)

# Extract the name and ID of all files in the config, narrow down to the subset we
# want for this notebook.
#
# This requires some rearrangement of cfg$datasets: When read from the file, 
# 'cfg$datasets' is an unnamed list, one item per data set, with internal named lists
# in a hierarchy. An example of the structure:
# List of 6
#  $ :List of 1
#   ..$ model_details:List of 7
#   .. ..$ files                 :List of 7
#   .. .. ..$ :List of 3
#   .. .. .. ..$ name  : chr "model_info"
#   .. .. .. ..$ id    : chr "syn61378590.13"
#   .. .. .. ..$ format: chr "csv"
#   .. .. ..$ :List of 3
#   .. .. .. ..$ name  : chr "allele_info"
#   .. .. .. ..$ id    : chr "syn64618791.6"
#   .. .. .. ..$ format: chr "csv"
# ...
#
# Remove the wrapping list around each data set object and fetching its "files" entry
files <- unlist(cfg$datasets, recursive = FALSE) |>
  lapply("[[", "files") |>
  
  # flatten from a list of lists of files to a list of files
  unlist(recursive = FALSE) |>
  
  # remove duplicate entries
  unique() |>
  
  # make each entry a single-row data frame with columns "name", "id", "format"
  lapply(as.data.frame) |>
  
  # concat into one data frame
  list_rbind() |>
  
  # subset to the files we want
  subset(name %in% c("human_transgene_allele_map", "biodom_genes_mm") |
           grepl("differential_expression", name))

print(files)

# Check: There should be at least 8 files and two of them need to have specific names
stopifnot(nrow(files) >= 8)
stopifnot(any(files$name == "human_transgene_allele_map"))
stopifnot(any(files$name == "biodom_genes_mm"))
```

## Download the data files from synapse

Note: if this is your first time logging into Synapse from R, you will need to change `synLogin()` to `synLogin(authToken=<your_token_string>)`.

```{r}
synLogin()

files$path <- sapply(files$id, function(id) {
  syn_file <- synGet(entity = id, downloadLocation = input_dir)
  return(syn_file$path)
})
```

## Get all Ensembl IDs from all files

```{r}
# Names of the fields containing Ensembl IDs in the files
ensembl_fields <- c("ensembl_id", "ensembl_gene_id")

ens <- lapply(files$path, function(file) {
  print(str_glue("Reading {basename(file)}..."))
  
  data <- read.csv(file)
  
  # Figure out which column exists in this data frame, and check we found exactly
  # one matching column
  column <- intersect(ensembl_fields, colnames(data))
  stopifnot(length(column) == 1)
  
  return(data[, column])
}) |>
  unlist() |>
  unique() |>
  na.omit()

print(str_glue("Ensembl IDs found: {length(ens)}"))
```

## Get gene symbols and aliases from Ensembl

### Mouse

```{r}
ens_mouse <- ens[grepl("ENSMUSG", ens)]
mart_mouse <- useMart(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")

genes_mouse <- getBM(attributes = c("ensembl_gene_id", "external_gene_name",
                                    "external_synonym"),
                     filters = "ensembl_gene_id",
                     values = ens_mouse,
                     mart = mart_mouse)

print(dim(genes_mouse))
print(str_glue("{sum(ens %in% genes_mouse$ensembl_gene_id)} of {length(ens_mouse)} ",
               "mouse Ensembl IDs returned by BioMart."))
print(head(genes_mouse))
```

### Human

```{r}
ens_human <- ens[grepl("ENSG", ens)]
mart_human <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

genes_human <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", 
                                    "external_synonym"),
                     filters = "ensembl_gene_id",
                     values = ens_human,
                     mart = mart_human)

print(dim(genes_human))
print(str_glue("{sum(ens %in% genes_human$ensembl_gene_id)} of {length(ens_human)} ",
               "human Ensembl IDs returned by BioMart."))
print(head(genes_human))
```

### Add Ensembl IDs missing from BioMart back to the data frame

```{r}
genes <- rbind(genes_mouse, genes_human)
missing_ids <- setdiff(ens, genes$ensembl_gene_id)

genes <- rbind(genes,
               data.frame(ensembl_gene_id = missing_ids, 
                          external_gene_name = "",
                          external_synonym = ""))
```

## Re-format results for Model AD Explorer

```{r}
# Helper function to get rid of duplicates and empty strings in the synonyms list. Returns
# the remaining values wrapped in a list object. The list object will be empty if there are
# no aliases.
clean_aliases <- function(synonyms) {
  aliases <- unique(synonyms) |> na.omit() |> setdiff("")
  return(list(aliases))
}

# There may be multiple rows for a single Ensembl ID, corresponding to different values for
# `external_synonym`. This collapses down to a single row per ID, with all synonyms in a list
genes_df <- genes |> 
  # Collapse to one row per ID/gene name
  group_by(ensembl_gene_id, external_gene_name) |>
  summarize(alias = clean_aliases(external_synonym), .groups = "drop") |>
  
  # Rename the gene symbol column
  dplyr::rename(gene_symbol = external_gene_name) |>
  
  # Sort by Ensembl ID
  arrange(ensembl_gene_id)

output_dir <- file.path("..", "output")
dir.create(output_dir, showWarnings = FALSE)
write_json(genes_df, file.path(output_dir, "mouse_gene_metadata.json"), pretty = TRUE)

print(head(genes_df))
```

## Upload JSON file to Synapse

```{r}
syn_file <- File(file.path(output_dir, "mouse_gene_metadata.json"),
                 description = "Ensembl IDs, gene symbols, and alias symbols for each gene used in the Model AD Explorer.",
                 parent = "syn51498054")
res <- synStore(syn_file, forceVersion = FALSE,
                used = files$id,
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-8_Create_ENSMUSG_Reference.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))
```
