
# Create Biomarker source file (MG-88)

This notebook generates a single harmonized source file containing the data required to surface Biomarker evidence in the Model-AD Explorer.

Required packages (will automatically be installed if necessary):

- synapser (also requires a Synapse account with an auth token)
- tidyverse


## Library setup
```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "http://cran.fhcrc.org"))
}

if (!require("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

library(synapser)
library(tidyverse)
```

## Utility functions & variables
```{r}
# Harmonized column spec for intermediate per-model files
keeps <- c('individualID', 'specimenID', 'model', 'type', 'measurement', 'units')

# Vector to store all downloaded synIds for Synapse provenance
data_sources <- character()


# Download a file from Synapse by synId and read it into a dataframe
download_file <- function(synId, df_name, type) {

  synLogin()

  # Capture all downloaded synIds for provenance
  data_sources <<- c(data_sources, synId)

  input_dir <- file.path("..", "input")
  id_parts <- strsplit(synId, '.', fixed = TRUE)
  id_value <- id_parts[[1]][1]
  version_value <- id_parts[[1]][2]

  if (is.na(version_value)) {
    version_value <- NULL
  }

  cat(paste0("Downloading ", synId, "..."))


  file <- synGet(id_value, version = version_value, downloadLocation = input_dir, ifcollision='overwrite.local')
  path <- file$path

  cat("\nDownload on:  ", date())
  cat("\n", synId, "\n- Modified on ", file$properties$modifiedOn, "\n- Version ", file$properties$versionNumber)
  cat("\npath: ", file$path, "\n")

  if (type == 'f') {
    data <- read_feather(path)
  } else {
    if (type == 'json') { data <- fromJSON(path) }
    if (type == 'csv' || type == 'table') { data <- read.csv(path) }
    if (type == 'tsv') { data <- read.csv(path, sep = "\t") }
    if (type == 'rda') { data <- load(path)}
    if (type == 'rds') {data <- readRDS(path)}
    df_name <- as.data.frame(data)
  }
}

# Join a data file to a pair of metadata files to get required age, sex, ageDeath, and genotype values
join_to_metadata <- function(df, individual_metadata, biospecimen_metadata) {

  individual_metadata <- subset(individual_metadata, select=c(individualID, sex, ageDeath, genotype))
  biospecimen_metadata <- subset(biospecimen_metadata, select=c(individualID, specimenID, tissue))
  metadata <- merge(individual_metadata, biospecimen_metadata, by = 'individualID')

  #df_out <- merge(df, metadata) # MG-12 - drop rows that are missing from biospecimen metadata
  df_out <- merge(df, metadata, fixed=TRUE)

  df_out
}
```


## Download data and metadata files from Synapse
Note that each file's synId and version is added to the data_sources vector via the download_file function.
The populated data_sources vector is used to populate Synapse provenance when the harmonized csv is uploaded.
```{r}
# 3xTG-AD
# metadata
threex_biospecimen <- download_file('syn23532198.3', type = "csv")
threex_individual <- download_file('syn23532199.4', type = "csv")
# Abetas
threex_abs_sol_source <- download_file('syn25913411.7', type = "csv")
threex_abs_insol_source <- download_file('syn25913418.8', type = "csv")
# NO NFL DATA

# 5xFAD
# metadata
fivex_biospecimen <- download_file('syn18876530.10', type = "csv")
fivex_individual <- download_file('syn18880070.14', type = "csv")
# Abetas
fivex_abs_sol_source <- download_file('syn22101767.3', type = "csv")
fivex_abs_insol_source <- download_file('syn22101766.1', type = "csv")
# NO NFL DATA

#NSS
# metadata
nss_biospecimen <- download_file('syn29568452.3', type = "csv")
nss_individual <- download_file('syn27147690.3', type = "csv")
# Abetas
nss_abs_sol_4mo_source <- download_file('syn29568621.2', type = "csv")
nss_abs_sol_12mo_source <- download_file('syn29568624.2', type = "csv")
nss_abs_insol_4mo_source <- download_file('syn29568619.1', type = "csv")
nss_abs_insol_12mo_source <- download_file('syn29568622.1', type = "csv")
# NFL
nss_nfl_cortex_source <- download_file('syn29568489.2', type = "csv")
nss_nfl_plasma_source <- download_file('syn29568490.1', type = "csv")

#Abca7
# metadata
abca7_biospecimen <- download_file('syn30859390.2', type = "csv")
abca7_individual <- download_file('syn30859394.2', type = "csv")
# Abetas
abca7_abs_sol_4mo_source <- download_file('syn30859386.1', type = "csv")
abca7_abs_sol_12mo_source <- download_file('syn30859377.1', type = "csv")
abca7_abs_insol_4mo_source <- download_file('syn30859383.1', type = "csv")
abca7_abs_insol_12mo_source <- download_file('syn30859387.1', type = "csv")
# NFL
abca7_nfl_cortex_source <- download_file('syn30859379.1', type = "csv")
abca7_nfl_plasma_source <- download_file('syn30859380.1', type = "csv")
```


## Generate per-model intermediate files with one measurement per row

### 3xTG-AD
```{r}
# Abeta measures
# Capture fraction types from file names in temp 'fraction_type' column and merge resulting dfs
threex_abs_sol <- threex_abs_sol_source %>% mutate(fraction_type = 'Soluble')
threex_abs_insol <- threex_abs_insol_source %>% mutate(fraction_type = 'Insoluble')
threex_abs_merged <- bind_rows(threex_abs_sol, threex_abs_insol)

# populate 'model' column
threex_abs_merged$model <- '3xTG-AD'

# Pivot Abeta columns; capture temp 'abeta_type' values and populate the target 'measurement' column
threex_abs_pivot <- threex_abs_merged %>%
        pivot_longer(
                cols = c(Abeta40, Abeta42),
                names_to = "abeta_type",
                values_to = "measurement"
        )

# Concatenate 'fraction_type' and 'abeta_type' columns to populate target 'type' column values
threex_abs_pivot$type <- paste(threex_abs_pivot$fraction_type, threex_abs_pivot$abeta_type)

# Drop unwanted columns and rows with no measurement value
threex_abs_with_nas <- threex_abs_pivot[keeps]
threex_abs <- threex_abs_with_nas[!is.na(threex_abs_with_nas$measurement),]

# No NfL measures for this model

threex_biomarkers <- threex_abs
```

### 5xFAD
```{r}
# Abeta measures
# Capture fraction types from file names in temp 'fraction_type' column and merge resulting dfs
fivex_abs_sol <- fivex_abs_sol_source %>% mutate(fraction_type = 'Soluble')
fivex_abs_insol <- fivex_abs_insol_source %>% mutate(fraction_type = 'Insoluble')
fivex_abs_merged <- bind_rows(fivex_abs_sol, fivex_abs_insol)

# populate 'model' column
fivex_abs_merged$model <- '5xFAD'

# Pivot Abeta columns; capture temp 'abeta_type' values and populate the target 'measurement' column
fivex_abs_pivot <- fivex_abs_merged %>%
        pivot_longer(
                cols = c(Abeta40, Abeta42),
                names_to = "abeta_type",
                values_to = "measurement"
        )

# Concatenate 'fraction_type' and 'abeta_type' columns to populate target 'type' column values
fivex_abs_pivot$type <- paste(fivex_abs_pivot$fraction_type, fivex_abs_pivot$abeta_type)

# Drop unwanted columns and rows with no measurement value
fivex_abs_with_nas <- fivex_abs_pivot[keeps]
fivex_abs <- fivex_abs_with_nas[!is.na(fivex_abs_with_nas$measurement),]

# No NfL measures for this model
fivex_biomarkers <- fivex_abs
```

### Trem2-R47H_NSS
```{r}

# Abeta measures
# Convert columns to consistent types
nss_abs_sol_12mo_source$individualID <- as.character(nss_abs_sol_12mo_source$individualID)
nss_abs_insol_12mo_source$individualID <- as.character(nss_abs_insol_12mo_source$individualID)
nss_abs_insol_12mo_source$Abeta42 <- suppressWarnings(as.numeric(nss_abs_insol_12mo_source$Abeta42))

# Capture fraction types from file names in temp 'fraction_type' column and merge resulting dfs
nss_abs_sol_4mo <- nss_abs_sol_4mo_source %>% mutate(fraction_type = 'Soluble')
nss_abs_sol_12mo <- nss_abs_sol_12mo_source %>% mutate(fraction_type = 'Soluble')
nss_abs_insol_4mo <- nss_abs_insol_4mo_source %>% mutate(fraction_type = 'Insoluble')
nss_abs_insol_12mo <- nss_abs_insol_12mo_source %>% mutate(fraction_type = 'Insoluble')
nss_abs_merged <- bind_rows(nss_abs_sol_4mo, nss_abs_sol_12mo, nss_abs_insol_4mo, nss_abs_insol_12mo)

# populate 'model' column
nss_abs_merged$model <- 'Trem2-R47H_NSS'

# Pivot Abeta columns; capture temp 'abeta_type' values and populate the target 'measurement' column
nss_abs_pivot <- nss_abs_merged %>%
        pivot_longer(
                cols = c(Abeta40, Abeta42),
                names_to = "abeta_type",
                values_to = "measurement"
        )

# Concatenate 'fraction_type' and 'abeta_type' columns to populate target 'type' column values
nss_abs_pivot$type <- paste(nss_abs_pivot$fraction_type, nss_abs_pivot$abeta_type)

# Drop unwanted columns and rows with no measurement value
nss_abs_with_nas <- nss_abs_pivot[keeps]
nss_abs <- nss_abs_with_nas[!is.na(nss_abs_with_nas$measurement),]

# NfL measures
# Merge source dfs
nss_nfl_merged <- bind_rows(nss_nfl_cortex_source, nss_nfl_plasma_source)

# Populate 'model' column
nss_nfl_merged$model <- 'Trem2-R47H_NSS'

# Pivot nfl column and populate the target 'measurement' column
nss_nfl_pivot <- nss_nfl_merged %>%
        pivot_longer(
                cols = c(NfL),
                names_to = "type",
                values_to = "measurement"
        )

# Drop unwanted columns and rows with no measurement value
nss_nfl_with_nas <- nss_nfl_pivot[keeps]
nss_nfl <- nss_nfl_with_nas[!is.na(nss_nfl_with_nas$measurement),]

# Merge Abetas and NfL
nss_biomarkers <- bind_rows(nss_abs, nss_nfl)

```

### Abca7*V1599M
```{r}
# Abeta measures
# Capture fraction types from file names in temp 'fraction_type' column and merge resulting dfs
abca7_abs_sol_4mo <- abca7_abs_sol_4mo_source %>% mutate(fraction_type = 'Soluble')
abca7_abs_sol_12mo <- abca7_abs_sol_12mo_source %>% mutate(fraction_type = 'Soluble')
abca7_abs_insol_4mo <- abca7_abs_insol_4mo_source %>% mutate(fraction_type = 'Insoluble')
abca7_abs_insol_12mo <- abca7_abs_insol_12mo_source %>% mutate(fraction_type = 'Insoluble')
abca7_abs_merged <- bind_rows(abca7_abs_sol_4mo, abca7_abs_sol_12mo, abca7_abs_insol_4mo, abca7_abs_insol_12mo)

# populate 'model' column
abca7_abs_merged$model <- 'Abca7*v1599M'

# Pivot Abeta columns; capture temp 'abeta_type' values and populate the target 'measurement' column
abca7_abs_pivot <- abca7_abs_merged %>%
        pivot_longer(
                cols = c(Abeta40, Abeta42),
                names_to = "abeta_type",
                values_to = "measurement"
        )

# Concatenate 'fraction_type' and 'abeta_type' columns to populate target 'type' column values
abca7_abs_pivot$type <- paste(abca7_abs_pivot$fraction_type, abca7_abs_pivot$abeta_type)

# Drop unwanted columns and rows with no measurement value
abca7_abs_with_nas <- abca7_abs_pivot[keeps]
abca7_abs <- abca7_abs_with_nas[!is.na(abca7_abs_with_nas$measurement),]

# NfL measures
# Merge source dfs
abca7_nfl_merged <- bind_rows(abca7_nfl_cortex_source, abca7_nfl_plasma_source)

# Populate 'model' column
abca7_nfl_merged$model <- 'Abca7*v1599M'

# Pivot nfl column and populate the target 'measurement' column
abca7_nfl_pivot <- abca7_nfl_merged %>%
        pivot_longer(
                cols = c(NfL),
                names_to = "type",
                values_to = "measurement"
        )

# Drop unwanted columns and rows with no measurement value
abca7_nfl_with_nas <- abca7_nfl_pivot[keeps]
abca7_nfl <- abca7_nfl_with_nas[!is.na(abca7_nfl_with_nas$measurement),]

# Merge abs and nfl
abca7_biomarkers <- bind_rows(abca7_abs, abca7_nfl)
```


## Join source files to metadata to populate age, tissue, and genotype values
```{r}
keeps <- c('model', 'type', 'measurement', 'units', 'ageDeath', 'tissue', 'sex', 'genotype')

# 3xTG-AD
threex_final <- join_to_metadata(threex_biomarkers, threex_individual, threex_biospecimen)[keeps]

# 5xFAD
fivex_with_8mo <- join_to_metadata(fivex_biomarkers, fivex_individual, fivex_biospecimen)[keeps]
# drop unwanted 8 month samples from 5xFAD
fivex_final <- fivex_with_8mo[fivex_with_8mo$ageDeath != 8,]

# Trem2-R47H_NSS
nss_final <- join_to_metadata(nss_biomarkers, nss_individual, nss_biospecimen)[keeps]

# Abca7*V1599M
abca7_final <- join_to_metadata(abca7_biomarkers, abca7_individual, abca7_biospecimen)[keeps]
```


## Sanity check nrow in generated files
```{r}
# 3xTG-AD
# before dropping NA values
stopifnot(nrow(threex_abs_sol) + nrow(threex_abs_insol) == nrow(threex_abs_merged))
stopifnot(nrow(threex_abs_merged) * 2 == nrow(threex_abs_pivot))
# after dropping NA values
stopifnot(nrow(threex_abs) == nrow(threex_biomarkers))
stopifnot(nrow(threex_biomarkers) == nrow(threex_final))

# 5xFAD
# before dropping NA values
stopifnot(nrow(fivex_abs_sol) + nrow(fivex_abs_insol) == nrow(fivex_abs_merged))
stopifnot(nrow(fivex_abs_merged) * 2 == nrow(fivex_abs_with_nas))
# after dropping NA values
stopifnot(nrow(fivex_abs) == nrow(fivex_biomarkers))
# correct for rows lost due to MG-12 (missing biospecimen metadata) and 8mo age filtering
stopifnot(nrow(fivex_biomarkers) == nrow(fivex_final) + 4 + nrow(fivex_with_8mo[fivex_with_8mo$ageDeath == 8,]))

# Trem2-R47H_NSS
# before dropping NA values
stopifnot(nrow(nss_abs_sol_4mo) + nrow(nss_abs_sol_12mo) + nrow(nss_abs_insol_4mo) + nrow(nss_abs_insol_12mo) == nrow(nss_abs_merged))
stopifnot(nrow(nss_abs_merged) * 2 == nrow(nss_abs_with_nas))
stopifnot(nrow(nss_nfl_cortex_source) + nrow(nss_nfl_plasma_source) == nrow(nss_nfl_with_nas))
stopifnot(nrow(nss_nfl_cortex_source) + nrow(nss_nfl_plasma_source) == nrow(nss_nfl_with_nas))
# after dropping NA values
stopifnot(nrow(nss_nfl) + nrow(nss_abs) == nrow(nss_biomarkers))
stopifnot(nrow(nss_biomarkers) == nrow(nss_final))

# Abca7*V1599M
# before dropping NA values
stopifnot(nrow(abca7_abs_sol_4mo) + nrow(abca7_abs_sol_12mo) + nrow(abca7_abs_insol_4mo) + nrow(abca7_abs_insol_12mo) == nrow(abca7_abs_merged))
stopifnot(nrow(abca7_abs_merged) * 2 == nrow(abca7_abs_with_nas))
stopifnot(nrow(abca7_nfl_cortex_source) + nrow(abca7_nfl_plasma_source) == nrow(abca7_nfl_with_nas))
# after dropping NA values
stopifnot(nrow(abca7_nfl) + nrow(abca7_abs) == nrow(abca7_biomarkers))
stopifnot(nrow(abca7_biomarkers) == nrow(abca7_final))
```


## merge files, write csv file, and upload to synapse
```{r}
biomarkers_merged <- bind_rows(threex_final, fivex_final, nss_final, abca7_final)


# standardize genotype values for display
biomarkers_final <- mutate(biomarkers_merged,
                           genotype = case_when(genotype %in% c("5XFAD_hemizygous","5XFAD_carrier") ~ "5xFAD",
                                                genotype == "3xTg-AD_homozygous"  ~ "3xTG-AD",
                                                genotype == "Trem2-R47H_NSS_homozygous"  ~ "Trem2-R47H_NSS",
                                                genotype == "5XFAD_carrier, Trem2-R47H_NSS_homozygous"  ~ "5xFADTrem2-R47H_NSS",
                                                genotype == "5XFAD_noncarrier"  ~ "C57BL/6J",
                                                genotype == "Abca7-V1599M_homozygous"  ~ "Abca7*V1599M",
                                                genotype == "5XFAD_carrier, Abca7-V1599M_homozygous"  ~ "5xFADAbca7*V1599M",
                                                TRUE ~ genotype))

# print final dataframe below
biomarkers_final

# write csv file
output_dir <- file.path("./output")
write.csv(biomarkers_final, file.path(output_dir, "biomarkers.csv"), row.names = FALSE)

# upload to synapse
syn_file <- File(file.path(output_dir, "biomarkers.csv"),
                 description = "Harmonized Biomarker data for Model-AD Explorer 2.0.",
                 parent = "syn51498054")

res <- synStore(syn_file, forceVersion = FALSE,
                used = data_sources,
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-88_Create_harmonized_biomarkers_source_file.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))

```

