# Create ENSMUSG Reference

This notebook downloads all data files listed in [data_sources.csv](https://github.com/Sage-Bionetworks/magora/blob/main/data-raw/gene_expressions/data_sources.csv), gets a list of all the Ensembl gene IDs used across all the files, and maps them to gene symbols and aliases. The file is then uploaded to Synapse here: [syn51615422](https://www.synapse.org/#!Synapse:syn51615422).

Required packages (will automatically be installed if necessary):

-   synapser (also requires a Synapse account with an auth token)
-   BiocManager
-   biomaRt
-   dplyr
-   jsonlite

## Library setup

```{r}
if (!require("synapser", quietly = TRUE)) {
  install.packages("synapser", repos = c("http://ran.synapse.org",
                                         "http://cran.fhcrc.org"))
}

if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

if (!require("biomaRt", quietly = TRUE)) {
  BiocManager::install("biomaRt")
}

if (!require("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}

if (!require("jsonlite", quietly = TRUE)) {
  install.packages("jsonlite")
}

library(synapser)
library(biomaRt)
library(dplyr)
library(arrow)
library(jsonlite)
```

## Download the data sources file

```{r}
input_dir <- file.path("..", "input")
dir.create(input_dir, showWarnings = FALSE)

data_sources_url <- "https://raw.githubusercontent.com/Sage-Bionetworks/magora/d4f21caf94694fee7659fc7d30748b7e64e08488/data-raw/gene_expressions/data_sources.csv"

data_sources_file <- file.path(input_dir, "data_sources.csv")
download.file(data_sources_url, data_sources_file, method = "wget")

```

## Download the data files from synapse

Note: if this is your first time logging into Synapse from R, you will need to change `synLogin()` to `synLogin(authToken=<your_token_string>)`.

```{r}
synLogin()
data_sources <- read.csv(data_sources_file)

files <- lapply(1:nrow(data_sources), function(row) {
  syn_file <- synGet(entity = data_sources$id[row],
                     version = data_sources$version[row],
                     downloadLocation = input_dir)
  return(syn_file$path)
})
```

## Get all Ensembl IDs from all files

```{r}
ens <- lapply(files, function(F) {
  data <- read.csv(F)
  return(data$GeneID)
})

ens <- unique(unlist(ens))

print(length(ens))
```

## Get gene symbols and aliases from Ensembl

```{r}
mart <- useMart(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")

genes <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "external_synonym"),
               filters = "ensembl_gene_id",
               values = ens,
               mart = mart)

print(dim(genes))
print(head(genes))
```

## Re-format results for Model AD Explorer

```{r}
genes_df <- genes %>% group_by(ensembl_gene_id, external_gene_name) %>%
              summarise(alias = list(unique(external_synonym))) %>%
              rename(gene_symbol = external_gene_name) %>%
              arrange(ensembl_gene_id)

output_dir <- file.path("..", "output")
dir.create(output_dir, showWarnings = FALSE)
write_json(genes_df, file.path(output_dir, "mouse_gene_metadata.json"), pretty = TRUE)

print(head(genes_df))
```

## Upload JSON file to Synapse

```{r}
syn_file <- File(file.path(output_dir, "mouse_gene_metadata.json"),
                 description = "Ensembl IDs, gene symbols, and alias symbols for each gene used in the Model AD Explorer.",
                 parent = "syn51498054")
res <- synStore(syn_file, forceVersion = FALSE,
                used = paste(data_sources$id, data_sources$version, sep = "."),
                executed = "https://github.com/Sage-Bionetworks/agora-data-tools/blob/dev/data_analysis/model_ad/notebooks/MG-8_Create_ENSMUSG_Reference.rmd")

print(paste0("ID: ", res$id, " v", res$versionNumber, ", Name: ", res$name))
```
